### 2024-03-12 letterB.qmd
All files that made the draft

---
title: "Draft open letter to editor"
format: gfm
editor: visual
---

Things David knows he still needs to do:

-   more proofreading

-   clear easy to follow citations

-   clear links to source data

To the editor and editorial board,

I am formally writing to point out the sgnificant errors in the paper from Gibson (2024)[^1], published in your journal. In particular, the paper asserts two claims:

[^1]: Reference to a footnote.

-   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of check names all[^2].
-   That Gibson's method, producing positive excess mortality, is more effective than Karlinsky & Kobak in determining mortality as it adjusts for population change due to immigration changes.

[^2]: Ariel Karlinsky, Dmitry Kobak (2021) Tracking excess mortality across countries during the COVID-19 pandemic with the World Mortality Dataset eLife 10:e69336

Neither of these claims are accurate.

# The underpinng of New Zealand's mortality measurement

For centuries, it has been established practice among actuaries, demographers, and health researchers to measure mortality using death rates by age were such data is available[^3]. This method superseded Crude Mortality Rates, where CMR takes into account total population change in comparisons by dividing the total number of deaths by the total number of people available to die, improving CMR by acknowledging that different ages have different risks of death so comparing age based risk of death on a uniform basis. In New Zealand, tutorials on making mortality comparisons by using death rates by age and standardising to a reference population have been published since the 1890s[^4].

[^3]: 1767 whatever

[^4]: NZ statistical handbook 1890

Age Standardised Mortality direct adjusts for both population growth and aging, as it is based on the number of deaths of the current resident population divided by the living current resident population, stratified by age group and time period. This directly measures the population change effects Gibson is trying to model.

While calculation of New Zealand's cummulative excess mortality for 2020-2022 can vary by a couple of percentage points depending on the exact method, using a death rates by age (standardised) method the Institute of Actuaries find New Zealand's excess mortality from week 10 2020 to the end of 2022 to be -4% (that there are 4% fewer deaths than would be expected based on precovid trends)[^5].

[^5]: Actuaries Institute: CMI Working Paper 180 https://www.actuaries.org.uk/learn-and-develop/continuous-mortality-investigation/cmi-working-papers/mortality-projections/cmi-working-paper-180

The Our World in Data graph, and the K&K method that underlies it is an approximation, and while if it is a better or worse approximation to Gibson's method will be discussed later, while recognising the Our World In Data graph is an approximation there are still several practically useful implicit features about it.

-   Being based on trends of deaths, it enables a common basis comparison with countries that lack age stratified current population information
-   It makes weekly comparisons available.
-   Able to be checked by any member of the public for the latest current worldwide information. This last is a function of Our World in Data presentation rather than the method, but it is still a practical benefit.

While the existance of centuries of mortality measurement (not mentioned in Gibson's article) does not directly address if New Zealand medical professionals are aware that best practice mortality calculations recognise that people of different ages die at different rates when using the publicly available Our World in Data graphs in media pieces, it does establish that there is an established methodology, not mentiond in Gibson's article, underpinnning claims of New Zealand negative excess, and that method has been used for centuries, world-wide, across multiple disciplines. And we can check if New Zealand medical professionals are aware of this by asking them- *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ’s are!)"*(Ashley Bloomfield) [^6]

[^6]: Ashley Bloomfield, 2024, personal communication

## Age Standardised Mortality Review

As Age Standardised Mortality is a standard across multiple disciplines, but is not raised in the Gibson paper, a brief review is in order.

Death rates for an age stratified group, for a time period in a given location, measure that rate at which that age group dies. These can be directly compared to death rates in other time periods and locations to compare if the group experience is more or less deadly. As it is a rate, it automatically adjusts for change in the population size of the age group.

While this enables direct comparisons of people of the same age, it does not enable total population comparisons (as total populations may have different age structures). Standardisation with a reference population addresses this by assessing the total population deaths on the basis of death rates by age if the places and times being compared had the same age structure.

*Figure 1*

![ASMR steps](figure1.png){fig-alt="ASMR process" fig-align="center"}

While some variation in calculation results can stem from choice of standard population, due to the variation of young and old within a population, pragmatically any population that can be argued to be analytically useful should produce similar results. Using single year ages groups by sex to 94 then 95+, and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -5% for 2020 to 2022. Using a standard reference population of 2021 when the borders were restricted gives a cumulative excess of -6%. Using a pre-covid standard population of 2019 gives a cumulative excess of -5%. Using the standard population of 1961, used by StatsNZ to maintain long-term continuity, cumulative excess is -5%.

### Baselines and excess mortality

Excess mortality is the difference to what is expected for mortality, so excess calculations using any method are sensitive what is considered normal- the baseline. For Age Standardised Mortality the normal process is to use a linear trend of preceding years, and in the case of a major multi-year event like a pandemic, the linear trend of pre-pandemic years.

As New Zealand, no different to most other advanced economies, has seen the rate of decrease in mortality slow over time, the length of the linear trend baseline period should be short enough not to be introducing error as a result of applying a straight line of best fit to curved data.

![baseline](figure2.png){fig-alt="Overlong baseline error" fig-align="center"}

Based on the influenza season patterns in recent pre-pandemic years, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline[^7]. New Zealand has seen a very similar history in this period to Australia, with the added limitation that incorporating 2011, the year of the Christchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is a poor influence point to use in a baseline as it is not a re-occurring annual event.

[^7]: Australian Bureau of Statistics. https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology

While many international comparisons use a 5 year baseline range to increase the number of countries for which weekly level data is available, in practice the results for New Zealand do not differ substantively between a 5 (2015-2019) and 7 (2013-2019) year baseline period. A 5 year linear baseline with a 2022 standard population gives a cumulative excess for the period of -5%, while a 7 year baseline gives a cumulative baseline of -5%.

## Comparison of Gibson and Karlinsky & Kobak

Starting from a commonly recognised best practice measure of mortality, age standised mortality, giving a cumulative excess mortality for 2020-2022 in the range of -4% to -6% depending on analytic assumptions, we can compare other methods to that.

The Our World in Data Cummalative Excess (Projection based on Karlinsky & Kobak) gives cumulative excess mortality for 2020-2022 as 0%

Gibson's Population adjusted for changes in growth rates model gives an excess mortality for 2020-2022 range of 1.7% to 5.3% with the central estimate rounding to 4%.

Though not mentioned in the paper, adding the Population as a model term to a death based model should be functionally equivalent to using the commonly known Crude Mortality Rate (deaths among the total population / total living population). And a 2015-2019 linear regression of the annual CMR gives an excess of 1%.

Gibson's method produces results further from actuarial standard methods than the K&K method. The reason for Gibson's model preforming so poorly compared to K&K lies in implicitly misapplying death rates by age. Young people migrate a lot and die little. Old people migrate little and die a lot. Apply a correction based on the amount of migration to the mortality of the total population (dominated by people not migrating) introduces unnecessary structural error.

![Comparison of ages of migration and death](immigration.png){fig-alt="Young people migrate, old people die" fig-align="center"}

Conversely, because the K&K method only uses death data, it implicitly limits itself to the population that dies. The population mainly dying are not migrating, so it does not diverge as far from ASM as the Gibson method when population changes take place that do not effect the population at risk of death.

Gibson asserted that it was not important to incorporate aging

*If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process.*

It does show up before covid. If you do exactly the same type of analysis, but before covid and the arguments around closing the border, of comparing the crude death rate (which ignores aging but adjusts for total population size) of the 3 years (2017-2019) after the 5 year baseline (2012-2016) to the expected values of the baseline, then the most recent three years are a cumulative 10% above baseline.

![Even if all years are before covid, the post-baseline years are above trend](figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}

I suggest anyone interested in this topic go explore population ages and death rates at StatsNZ themeselves, and while there have a look at New Zealand death rates by age, or indeed the public, downloadable standardised death rates.

\[ Note only for the draft, for those with whom I had very early discussions that the predicted results from Gibsons model are impossible if you look at the change in population from 2019 to 2020 and the change in deaths from 2019 to 2020, due to the young migrate/old die thing. There were 2 reasons: (1) The model Gibson used in his article is sufficiently undocumented I cannot exactly replicate is claims over 3 years, so making a claim about 1 year is filed with assumptions that there would be distracting quibble fights about. (2) While it demonstrates this particular model is impossible, that distracts from the general point that Age Standardised Mortality Rates exist, are industry standard, and shouldn't be ignored.\]

### 2024-03-12 NZfourpane.R

@@ -1,0 +1,266 @@
+library(dplyr)
+library(tidyr)
+
+Dth <- read.csv("VSD_Death_by_Age_Dec.csv", skip=2)
+Dth00_59 <- Dth |> select(Year = X., Less.than.1.year:X59.years) |> 
+  filter(!is.na(X50.years)) |> 
+  mutate(Year = as.numeric(Year)) |> 
+  gather(key="Age",value="Dths", -Year) |> 
+  summarise(.by=Year, Deaths = sum(Dths)) |> 
+  filter(Year >2012) |> 
+  mutate(dataset = "Resident Deaths",
+         Age = "age 0 to 59\n33% of deaths") |> 
+  rename(Number = Deaths)
+Dth60_79 <- Dth |> select(Year = X., X60.years:X79.years) |> 
+  filter(!is.na(X70.years)) |> 
+  mutate(Year = as.numeric(Year)) |> 
+  gather(key="Age",value="Dths", -Year) |> 
+  summarise(.by=Year, Deaths = sum(Dths)) |> 
+  filter(Year >2012) |> 
+  mutate(dataset = "Resident Deaths",
+         Age = "age 60 to 79\n27% of deaths") |> 
+  rename(Number = Deaths)
+Dth80_plus <- Dth |> select(Year = X., X80.years:X100.years.and.over.1) |> 
+  filter(!is.na(X85.years)) |> 
+  mutate(Year = as.numeric(Year)) |> 
+  gather(key="Age",value="Dths", -Year) |> 
+  summarise(.by=Year, Deaths = sum(Dths)) |> 
+  filter(Year >2012) |> 
+  mutate(dataset = "Resident Deaths",
+         Age = "age 80 plus\n40% of deaths") |> 
+  rename(Number = Deaths)
+Tdth00 <- sum(Dth00_59$Deaths)
+Tdth60 <- sum(Dth60_79$Deaths)
+Tdth80 <- sum(Dth80_plus$Deaths)
+Tdth80/(Tdth00 + Tdth60 + Tdth80)
+
+Pop <- read.csv("DPE_Mean_Pop_by_Age_Dec.csv", skip=3)
+Pop00_59 <- Pop |> select(Year = X., X0.Years:X59.Years) |> 
+  filter(!is.na(X50.Years)) |> 
+  mutate(Year = as.numeric(Year)) |> 
+  gather(key="Age",value="pop", -Year) |> 
[…]

### 2024-03-12 figure4.R

@@ -1,0 +1,87 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+library(patchwork)
+source("theme.R")
+six_cols <- colorblind_pal()(6)
+
+CMR <- read.csv("DMM_CDRdec.csv", skip=1, 
+                 col.names = c("Year","CDR1K")) |> 
+  filter(!is.na(CDR1K)) |> 
+  mutate(Year = as.numeric(Year)) 
+
+mod7y <- lm(CDR1K ~ Year, data=CMR |> filter(Year %in% 2013:2019))
+CMR$mod7y <- mod7y$coefficients[1] + 
+  CMR$Year * mod7y$coefficients[2]
+
+mod5y <- lm(CDR1K ~ Year, data=CMR |> filter(Year %in% 2015:2019))
+CMR$mod5y <- mod5y$coefficients[1] + 
+  CMR$Year * mod5y$coefficients[2]
+CMR$percent5year <- 100*CMR$CDR1K/CMR$mod5y-100
+CMR$percent7year <- 100*CMR$CDR1K/CMR$mod7y-100
+
+excessdf <- data.frame(Baseline = c("2015-2019", "2013-2019", "2015-2019", "2013-2019",
+                          "2015-2019", "2013-2019", "2015-2019", "2013-2019"),
+           Year = c("2020","2020","2021","2021","2022","2022","nett","nett"),
+           value = c(round(CMR$percent5year[CMR$Year == 2020],2),
+                     round(CMR$percent7year[CMR$Year == 2020],2),
+                     round(CMR$percent5year[CMR$Year == 2021],2),
+                     round(CMR$percent7year[CMR$Year == 2021],2),
+                     round(CMR$percent5year[CMR$Year == 2022],2),
+                     round(CMR$percent7year[CMR$Year == 2022],2),
+                     round(sum(CMR$percent5year[CMR$Year %in% 2020:2022]),2),
+                     round(sum(CMR$percent7year[CMR$Year %in% 2020:2022]),2)))
+innergraph <- ggplot(excessdf, aes(x=Year,y=Baseline, label=value))+
+  geom_tile(fill="white", colour="black") +geom_text() + theme_tufte()  +
+  scale_x_discrete(position = "top")+
+  labs(title="Excess as percent of expected", y="Baseline\n ",x="")
+  
+ggplot(CMR |> filter(Year %in% 2012:2023), 
+              aes(x=Year, y=CDR1K))+
[…]

### 2024-03-12 intextCalcs.R

@@ -1,0 +1,105 @@
+library(dplyr)
+library(tidyr)
+
+# figure 01a Male death rates by age 
+autoAge <- paste(2:99,"years")
+autoAge <- c("Less than 1 year","1 year",autoAge,"100 years and over")
+VSD <- read.csv("VSD_deaths_age_sex.csv", skip=1,
+                col.names = c("Year","Age","Male","Female")) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
+  fill(Year) |> 
+  filter(Age %in% autoAge) |> 
+  group_by(Year) |> 
+  slice(2:n()) |> 
+  ungroup() |> 
+  mutate(Age = ifelse(Age == "Less than 1 year","0 years",Age),
+         AgeN = as.numeric(gsub(" .*","",Age)),
+         AgeGroup = case_when(AgeN > 94 ~ "95 plus",
+                              TRUE ~ as.character(AgeN)),
+         Year = as.numeric(Year)) |> 
+  summarise(.by = c(Year,AgeGroup),
+            Males=sum(Male),
+            Females=sum(Female)) |> 
+  filter(Year > 2005)
+
+autoAge <- paste(0:94,"Years")
+autoAge <- c(autoAge,"95 Years and Over")
+DPE <- read.csv("DPEpopulation_by_age_sex.csv", skip=1,
+                col.names = c("Year","Age","Sex","MeanPop")) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year),
+         Age = ifelse(Age == " ", NA_character_,Age)) |> 
+  fill(Year,Age) |> 
+  filter(Sex %in% c("Male","Female"),
+         Age %in% autoAge) |> 
+  mutate(AgeN = as.numeric(gsub(" .*","",Age)),
+         AgeGroup = case_when(AgeN > 94 ~ "95 plus",
+                              TRUE ~ as.character(AgeN)),
+         AnnualPop = as.numeric(MeanPop),
+         Year = as.numeric(Year)) |> 
+  summarise(.by = c(Year,AgeGroup,Sex),
+            Population=sum(AnnualPop, na.rm=TRUE)) |> 
+  filter(Year > 2005)
+
+Males <- DPE |> 
+  filter(Sex == "Male") |> 
+  inner_join(VSD,by = join_by(Year, AgeGroup)) |> 
[…]

### 2024-03-12 rawfiles/letterhome.qmd
reorganising to tidy the repo look

--- 2024-03-12 letterB.qmd
@@
 title: "Draft open letter to editor"
-format: gfm
+format: html
 editor: visual
@@
 
-Things David knows he still needs to do:
-
--   more proofreading
-
--   clear easy to follow citations
-
--   clear links to source data
-
 To the editor and editorial board,
 
-I am formally writing to point out the sgnificant errors in the paper from Gibson (2024)[^1], published in your journal. In particular, the paper asserts two claims:
+I am formally writing to point out the significant errors in the paper from Gibson (2024)[^1], published in your journal. In particular, the paper asserts two claims:
 
@@
 
--   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of check names all[^2].
--   That Gibson's method, producing positive excess mortality, is more effective than Karlinsky & Kobak in determining mortality as it adjusts for population change due to immigration changes.
+-   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of KRansky et all[^2].
+-   That Gibson's method, producing postive excess mortality, is more effective than Kransky & in determining mortality as it adjusts for population change due to immigration changes.
 
-[^2]: Ariel Karlinsky, Dmitry Kobak (2021) Tracking excess mortality across countries during the COVID-19 pandemic with the World Mortality Dataset eLife 10:e69336
+[^2]: Kransky's article reference
 
@@
 For centuries, it has been established practice among actuaries, demographers, and health researchers to measure mortality using death rates by age were such data is available[^3].
-This method superseded Crude Mortality Rates, where CMR takes into account total population change in comparisons by dividing the total number of deaths by the total number of people available to die, improving CMR by acknowledging that different ages have different risks of death so comparing age based risk of death on a uniform basis.
-In New Zealand, tutorials on making mortality comparisons by using death rates by age and standardising to a reference population have been published since the 1890s[^4].
+In New Zealand, tutorials on making mortality comparisons by using death rates by age and standardising to a reference population have been published since the 1890s\[\^NZ statistical handbook 1890\].
 
@@
 
-[^4]: NZ statistical handbook 1890
+Age Standardised Mortality direct adjusts for population growth and aging, as it is based on the number of deaths of the current resident population divided by the living current resident population, stratified by age group and time period.
+This directly measures the population change effects Gibson is trying to model, as any population change that does not change risk of dying leaves the rate of living people who die unchanged.
 
-Age Standardised Mortality direct adjusts for both population growth and aging, as it is based on the number of deaths of the current resident population divided by the living current resident population, stratified by age group and time period.
-This directly measures the population change effects Gibson is trying to model.
-
-While calculation of New Zealand's cummulative excess mortality for 2020-2022 can vary by a couple of percentage points depending on the exact method, using a death rates by age (standardised) method the Institute of Actuaries find New Zealand's excess mortality from week 10 2020 to the end of 2022 to be -4% (that there are 4% fewer deaths than would be expected based on precovid trends)[^5].
-
-[^5]: Actuaries Institute: CMI Working Paper 180 https://www.actuaries.org.uk/learn-and-develop/continuous-mortality-investigation/cmi-working-papers/mortality-projections/cmi-working-paper-180
+While calculation of New Zealand's cummulative excess mortality for 2020-2022 can vary by a couple of percentage points depending on the exact method, using a death rates by age (standardised) method the Institute of Actuaries find New Zealand's excess mortality from week 10 2020 to the end of 2022 to be -4% (that there are 4% fewer deaths than would be expected based on precovid trends)\[\^actuaries\].
 
@@
 -   Being based on trends of deaths, it enables a common basis comparison with countries that lack age stratified current population information
--   It makes weekly comparisons available.
+-   Weekly
 -   Able to be checked by any member of the public for the latest current worldwide information.
@@
 While the existance of centuries of mortality measurement (not mentioned in Gibson's article) does not directly address if New Zealand medical professionals are aware that best practice mortality calculations recognise that people of different ages die at different rates when using the publicly available Our World in Data graphs in media pieces, it does establish that there is an established methodology, not mentiond in Gibson's article, underpinnning claims of New Zealand negative excess, and that method has been used for centuries, world-wide, across multiple disciplines.
-And we can check if New Zealand medical professionals are aware of this by asking them- *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ’s are!)"*(Ashley Bloomfield) [^6]
-
-[^6]: Ashley Bloomfield, 2024, personal communication
+And we can check if New Zealand medical professionals are aware of this by asking them- " the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ's are!)" (Bloomfield, 2024) \[\^Pers Comm.\]
 
@@
 
-While this enables direct comparisons of people of the same age, it does not enable total population comparisons (as total populations may have different age structures).
+While this enables direct comparisons of people of the same age, it does not enable total population comparisons (as total populations be may have different age structures).
 Standardisation with a reference population addresses this by assessing the total population deaths on the basis of death rates by age if the places and times being compared had the same age structure.
@@
 Excess mortality is the difference to what is expected for mortality, so excess calculations using any method are sensitive what is considered normal- the baseline.
-For Age Standardised Mortality the normal process is to use a linear trend of preceding years, and in the case of a major multi-year event like a pandemic, the linear trend of pre-pandemic years.
+For Age Standardised Mortality the normal process is to use a linear trend of preceeding years, and in the case of a major multiyear event like a pandemic which does not yet have a post-acute pandemic trend established, the linear trend of pre-pandemic years.
 
@@
 
-![baseline](figure2.png){fig-alt="Overlong baseline error" fig-align="center"}
+*Figure 2 goes here*
 
-Based on the influenza season patterns in recent pre-pandemic years, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline[^7].
-New Zealand has seen a very similar history in this period to Australia, with the added limitation that incorporating 2011, the year of the Christchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is a poor influence point to use in a baseline as it is not a re-occurring annual event.
-
-[^7]: Australian Bureau of Statistics. https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology
+Based on the influenza season patterns in recent pre-pandemic years, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline\[\^ABS link\].
+New Zealand has seen a very similar history in this period to Australia, with the added limitation that incorporating 2011, the year of the Christhchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is a poor addition to a baseline as it is not a re-occuring annual event.
 
@@
 
-## Comparison of Gibson and Karlinsky & Kobak
+## Comparison with Gibson
 
-Starting from a commonly recognised best practice measure of mortality, age standised mortality, giving a cumulative excess mortality for 2020-2022 in the range of -4% to -6% depending on analytic assumptions, we can compare other methods to that.
+With the existance of a common, internationally recognised best practice measure of mortality, age standised mortality, giving a cumulative excess mortality for 2020-2022 in the range of -4% to -6% depending on analytic assumptions.
 
-The Our World in Data Cummalative Excess (Projection based on Karlinsky & Kobak) gives cumulative excess mortality for 2020-2022 as 0%
-
-Gibson's Population adjusted for changes in growth rates model gives an excess mortality for 2020-2022 range of 1.7% to 5.3% with the central estimate rounding to 4%.
-
-Though not mentioned in the paper, adding the Population as a model term to a death based model should be functionally equivalent to using the commonly known Crude Mortality Rate (deaths among the total population / total living population).
-And a 2015-2019 linear regression of the annual CMR gives an excess of 1%.
-
-Gibson's method produces results further from actuarial standard methods than the K&K method.
-The reason for Gibson's model preforming so poorly compared to K&K lies in implicitly misapplying death rates by age.
-Young people migrate a lot and die little.
-Old people migrate little and die a lot.
-Apply a correction based on the amount of migration to the mortality of the total population (dominated by people not migrating) introduces unnecessary structural error.
-
-![Comparison of ages of migration and death](immigration.png){fig-alt="Young people migrate, old people die" fig-align="center"}
-
-Conversely, because the K&K method only uses death data, it implicitly limits itself to the population that dies.
-The population mainly dying are not migrating, so it does not diverge as far from ASM as the Gibson method when population changes take place that do not effect the population at risk of death.
-
-Gibson asserted that it was not important to incorporate aging
-
-*If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process.*
-
-It does show up before covid.
-If you do exactly the same type of analysis, but before covid and the arguments around closing the border, of comparing the crude death rate (which ignores aging but adjusts for total population size) of the 3 years (2017-2019) after the 5 year baseline (2012-2016) to the expected values of the baseline, then the most recent three years are a cumulative 10% above baseline.
-
-![Even if all years are before covid, the post-baseline years are above trend](figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}
-
-I suggest anyone interested in this topic go explore population ages and death rates at StatsNZ themeselves, and while there have a look at New Zealand death rates by age, or indeed the public, downloadable standardised death rates.
-
-\[ Note only for the draft, for those with whom I had very early discussions that the predicted results from Gibsons model are impossible if you look at the change in population from 2019 to 2020 and the change in deaths from 2019 to 2020, due to the young migrate/old die thing.
-There were 2 reasons: (1) The model Gibson used in his article is sufficiently undocumented I cannot exactly replicate is claims over 3 years, so making a claim about 1 year is filed with assumptions that there would be distracting quibble fights about.
-(2) While it demonstrates this particular model is impossible, that distracts from the general point that Age Standardised Mortality Rates exist, are industry standard, and shouldn't be ignored.\]
+The reason for Gibson's model preforming so poorly compared to

### 2024-03-12 rawfiles/KnK.R

@@ -1,0 +1,55 @@
+library(dplyr)
+library(tidyr)
+library(lubridate)
+NZstmf <- read.csv("NZL_NPstmfout.csv") |> 
+  filter(Sex == "b")
+NZQTR <- read.csv("TotalQuarterlyPopulationAsAt.csv", skip=3,
+                  col.names = c("YRQR","Pop")) |> 
+  filter(!is.na(Pop)) |> 
+  separate(YRQR, into=c("YR", "QR"), sep="Q", convert = TRUE) |> 
+  mutate(asDate = ceiling_date(ISOdate(YR,QR*3,25), 
+                               unit="quarter") - days(1),
+         Week=isoweek(asDate),
+         Year=isoyear(asDate)) |> 
+  select(Year,Week,Pop)
+## bluntly apply the quarter as a block rather than
+## imputing each week, but it should be in the right ballpark
+
+joint <- NZstmf |> left_join(NZQTR, by = join_by(Year, Week)) |> 
+  fill(Pop) |> mutate(CDR1K = 1000*Total/Pop,
+                      logPop = log(Pop))
+
+KnKexpect <- function(x, weekly_data=joint){
+  targetRow <- weekly_data[x,]
+  sexis <- targetRow$Sex
+  weekno <- targetRow$Week
+  if (weekno == 53){weekno <- 52}
+  regresset <- weekly_data[weekly_data$Year %in% 2015:2019 & 
+                             weekly_data$Week==weekno &
+                             weekly_data$Sex==sexis,]
+  Knkmodel <- lm(Total ~ Year, data=regresset)
+  targetRow$KnKexpected <- Knkmodel$coefficients[1] +
+    Knkmodel$coefficients[2] * targetRow$Year
+  DRmodel <- lm(CDR1K ~ Year, data=regresset)
+  targetRow$CDRexpected <- DRmodel$coefficients[1] +
+    DRmodel$coefficients[2] * targetRow$Year
+  Popmodel <- lm(Total ~ Year + Pop, data=regresset)
+  targetRow$Popexpected <- predict(Popmodel,newdata=targetRow)
+  Logmodel <- lm(Total ~ Year + logPop, data=regresset)
+  targetRow$Logexpected <- predict(Logmodel,newdata=targetRow)
+  return(targetRow)
+}
+
+w.Expected <- lapply(1:nrow(NZstmf), KnKexpect)
+df.Expected <- bind_rows(w.Expected) 
+
[…]

### 2024-03-12 rawfiles/annualFigure.R

@@ -1,0 +1,91 @@
+
+
+
+
+source("Serbia.R")
+source("NewZealand.R")
+source("Sweden.R")
+source("Bulgaria.R")
+rm(list=setdiff(ls(), c("BUL", "NZ",
+                        "SWE", "SRB")))
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+source("theme.R")
+Bmod <- lm(Standth ~ Year, data=BUL |> 
+             filter(Year %in% 2013:2019))
+BUL$exp <- Bmod$coefficients[1] + 
+  Bmod$coefficients[2] * BUL$Year
+Cexcess = round(sum(100 * BUL$Standth[21:23] / 
+                   BUL$exp[21:23]-100),2)
+Rexcess = round(sum(BUL$Standth[21:23]) - sum(BUL$exp[21:23]),0)
+BUL$Country = paste0(BUL$Country,"<P><P>Cummulative excess <span style='color:", six_cols[3],"'>",
+                     "2020 to 2022</span> ",
+                     Cexcess, "% (",Rexcess," people)")
+NZmod <- lm(Standth ~ Year, data=NZ |> 
+          filter(Year %in% 2013:2019))
+NZ$exp <- NZmod$coefficients[1] + 
+  NZmod$coefficients[2] * NZ$Year
+Cexcess = round(sum(100 * NZ$Standth[21:23] / 
+                      NZ$exp[21:23]-100),2)
+Rexcess = round(sum(NZ$Standth[21:23]) - sum(NZ$exp[21:23]),0)
+NZ$Country = paste0(NZ$Country,"<P><P>Cummulative excess <span style='color:", six_cols[3],"'>",
+                    "2020 to 2022</span> ",
+                    Cexcess, "% (",Rexcess," people)")
+SEmod <- lm(Standth ~ Year, data=SWE |> 
+             filter(Year %in% 2013:2019))
+SWE$exp <- SEmod$coefficients[1] + 
+  SEmod$coefficients[2] * SWE$Year
+Cexcess = round(sum(100 * SWE$Standth[21:23] / 
+                      SWE$exp[21:23]-100),2)
+Rexcess = round(sum(SWE$Standth[21:23]) - sum(SWE$exp[21:23]),0)
+SWE$Country = paste0(SWE$Country,"<P><P>Cummulative excess <span style='color:", six_cols[3],"'>",
+                     "2020 to 2022</span> ",
[…]

### 2024-03-12 rawfiles/figure2.R

@@ -1,0 +1,93 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+library(patchwork)
+
+ASMR <- read.csv("DMM_Standardised_Death_Rates.csv", skip=1, 
+                 col.names = c("Year","ASMR1K")) |> 
+  filter(!is.na(ASMR1K)) |> 
+  mutate(Year = as.numeric(Year)) |> 
+  filter(Year > 1990)
+
+longmod <- lm(ASMR1K ~ Year, data=ASMR |> filter(Year < 2017))
+ASMR$longExp <- longmod$coefficients[1] + 
+  ASMR$Year * longmod$coefficients[2]
+
+shortmod <- lm(ASMR1K ~ Year, data=ASMR |> filter(Year %in% 2010:2016))
+ASMR$shortExp <- shortmod$coefficients[1] + 
+  ASMR$Year * shortmod$coefficients[2]
+ASMR$shortExp[ASMR$Year < 2010] <- NA_real_
+
+xis<- c(2017,
+        2018,
+        2019,
+        2019,
+        2018,
+        2017)
+yis <- c(
+  3.180031,
+  3.073993,
+  2.967956,
+  3.351429,
+  3.406429,
+  3.461429)
+
+structural_over17_19 <- sum(ASMR$shortExp[ASMR$Year %in% 2017:2019] /
+  ASMR$longExp[ASMR$Year %in% 2017:2019])
+
+ggplot(ASMR |> filter(Year < 2020), 
+              aes(x=Year, y=ASMR1K))+
+  annotate("polygon",x=xis,y=yis,fill="skyblue")+
+  geom_point(size=0.3) +
+         geom_line(aes(y=longExp), linetype=2)+
+         geom_line(aes(y=shortExp), linetype=2) +
+  geom_line(data=ASMR |> filter(Year < 2017), aes(y=longExp)) +
+  geom_line(data=ASMR |> filter(Year < 2017),aes(y=shortExp)) +
+  theme_minimal(base_family="Open Sans",
+                base_size = 6) +
+  theme(
+    plot.background = element_rect(fill = "#FAFAFA", colour="#FAFAFA"),
+    panel.background = element_rect(fill = "#FFFFFF", colour = "#FFFFFF"),
+    panel.spacing.x = unit(9, "points"),
+    panel.spacing.y = unit(3, "points"),
+    panel.grid = element_blank(),
[…]

### 2024-03-12 rawfiles/figure5.R

@@ -1,0 +1,48 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+library(patchwork)
+source("theme.R")
+six_cols <- colorblind_pal()(6)
+
+VSD <- read.csv("DMM_CDRdec.csv", skip=1) |> 
+  rename(Year=1, Deaths=2) |> 
+  filter(!is.na(Deaths)) |> 
+  mutate(Year = as.numeric(Year)) 
+
+model <- lm(Deaths ~ Year, data=VSD |> filter(Year %in% 2012:2016))
+VSD$expected <- model$coefficients[1] + 
+  VSD$Year * model$coefficients[2]
+
+ggplot(VSD |> filter(Year %in% 2010:2019), aes(x=Year, y=Deaths))+
+  geom_point(size=0.3) +
+         geom_line(aes(y=expected), linetype=2, colour=six_cols[2])+
+  geom_line(data=VSD |> filter(Year %in% 2012:2016), aes(y=expected), colour=six_cols[2]) +
+  theme_minimal(base_family="Open Sans",
+                base_size = 6) +
+  theme(
+    plot.background = element_rect(fill = "#FAFAFA", colour="#FAFAFA"),
+    panel.background = element_rect(fill = "#FFFFFF", colour = "#FFFFFF"),
+    panel.spacing.x = unit(9, "points"),
+    panel.spacing.y = unit(3, "points"),
+    panel.grid = element_blank(),
+    strip.background = element_rect(fill= "#FFFFFF", colour="#EFEFEF"),
+    strip.text = element_text(margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")),
+    strip.placement = "inside",
+    axis.title.y.left = element_text(margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"))
+  ) +
+  labs(title="The crude death in 2017-2019 are above the 5 year linear trend for 2012-2016",
+       subtitle="This is because, before covid, the accelerating aging population is accelerating deaths",
+       y="Crude Death Rate", x="")
+
+ggsave(filename="figure5.png",
+       height=4, width = 6, dpi=300, units = "in", bg = "white")
+
+VSD$perc = 100*VSD$Deaths/VSD$expected-100
+
+
+## https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology
+
+

### 2024-03-12 rawfiles/figures.R

@@ -1,0 +1,258 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+library(patchwork)
+library(forcats)
+
+# figure 01a Male death rates by age since 2013
+autoAge <- paste(2:99,"years")
+autoAge <- c("Less than 1 year","1 year",autoAge,"100 years and over")
+VSD <- read.csv("VSD_deaths_age_sex.csv", skip=1,
+                col.names = c("Year","Age","Male","Female")) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
+  fill(Year) |> 
+  filter(Age %in% autoAge) |> 
+  group_by(Year) |> 
+  slice(2:n()) |> 
+  ungroup() |> 
+  mutate(Age = ifelse(Age == "Less than 1 year","0 years",Age),
+         AgeN = as.numeric(gsub(" .*","",Age)),
+         AgeGroup = case_when(AgeN > 79 ~ "80 plus",
+                              AgeN > 59 ~ "60 to 79",
+                              AgeN > 39 ~ "40 to 59",
+                              AgeN > 19 ~ "20 to 39",
+                              TRUE ~ "0 to 19"),
+         Year = as.numeric(Year)) |> 
+  summarise(.by = c(Year,AgeGroup),
+            Males=sum(Male),
+            Females=sum(Female)) |> 
+  filter(Year > 2012)
+
+autoAge <- paste(0:94,"Years")
+autoAge <- c(autoAge,"95 Years and Over")
+DPE <- read.csv("DPEpopulation_by_age_sex.csv", skip=1,
+                col.names = c("Year","Age","Sex","MeanPop")) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year),
+         Age = ifelse(Age == " ", NA_character_,Age)) |> 
+  fill(Year,Age) |> 
+  filter(Sex %in% c("Male","Female"),
+         Age %in% autoAge) |> 
+  mutate(AgeN = as.numeric(gsub(" .*","",Age)),
+         AgeGroup = case_when(AgeN > 79 ~ "80 plus",
+                              AgeN > 59 ~ "60 to 79",
+                              AgeN > 39 ~ "40 to 59",
+                              AgeN > 19 ~ "20 to 39",
+                              TRUE ~ "0 to 19"),
[…]

### 2024-03-12 KnK.R
reorganising to tidy the repo look

@@ -1,55 +1,0 @@
-library(dplyr)
-library(tidyr)
-library(lubridate)
-NZstmf <- read.csv("NZL_NPstmfout.csv") |> 
-  filter(Sex == "b")
-NZQTR <- read.csv("TotalQuarterlyPopulationAsAt.csv", skip=3,
-                  col.names = c("YRQR","Pop")) |> 
-  filter(!is.na(Pop)) |> 
-  separate(YRQR, into=c("YR", "QR"), sep="Q", convert = TRUE) |> 
-  mutate(asDate = ceiling_date(ISOdate(YR,QR*3,25), 
-                               unit="quarter") - days(1),
-         Week=isoweek(asDate),
-         Year=isoyear(asDate)) |> 
-  select(Year,Week,Pop)
-## bluntly apply the quarter as a block rather than
-## imputing each week, but it should be in the right ballpark
-
-joint <- NZstmf |> left_join(NZQTR, by = join_by(Year, Week)) |> 
-  fill(Pop) |> mutate(CDR1K = 1000*Total/Pop,
-                      logPop = log(Pop))
-
-KnKexpect <- function(x, weekly_data=joint){
-  targetRow <- weekly_data[x,]
-  sexis <- targetRow$Sex
-  weekno <- targetRow$Week
-  if (weekno == 53){weekno <- 52}
-  regresset <- weekly_data[weekly_data$Year %in% 2015:2019 & 
-                             weekly_data$Week==weekno &
-                             weekly_data$Sex==sexis,]
-  Knkmodel <- lm(Total ~ Year, data=regresset)
-  targetRow$KnKexpected <- Knkmodel$coefficients[1] +
-    Knkmodel$coefficients[2] * targetRow$Year
-  DRmodel <- lm(CDR1K ~ Year, data=regresset)
-  targetRow$CDRexpected <- DRmodel$coefficients[1] +
-    DRmodel$coefficients[2] * targetRow$Year
-  Popmodel <- lm(Total ~ Year + Pop, data=regresset)
-  targetRow$Popexpected <- predict(Popmodel,newdata=targetRow)
-  Logmodel <- lm(Total ~ Year + logPop, data=regresset)
-  targetRow$Logexpected <- predict(Logmodel,newdata=targetRow)
-  return(targetRow)
-}
-
-w.Expected <- lapply(1:nrow(NZstmf), KnKexpect)
-df.Expected <- bind_rows(w.Expected) 
-
[…]

### 2024-03-12 annualFigure.R

@@ -1,91 +1,0 @@
-
-
-
-
-source("Serbia.R")
-source("NewZealand.R")
-source("Sweden.R")
-source("Bulgaria.R")
-rm(list=setdiff(ls(), c("BUL", "NZ",
-                        "SWE", "SRB")))
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-source("theme.R")
-Bmod <- lm(Standth ~ Year, data=BUL |> 
-             filter(Year %in% 2013:2019))
-BUL$exp <- Bmod$coefficients[1] + 
-  Bmod$coefficients[2] * BUL$Year
-Cexcess = round(sum(100 * BUL$Standth[21:23] / 
-                   BUL$exp[21:23]-100),2)
-Rexcess = round(sum(BUL$Standth[21:23]) - sum(BUL$exp[21:23]),0)
-BUL$Country = paste0(BUL$Country,"<P><P>Cummulative excess <span style='color:", six_cols[3],"'>",
-                     "2020 to 2022</span> ",
-                     Cexcess, "% (",Rexcess," people)")
-NZmod <- lm(Standth ~ Year, data=NZ |> 
-          filter(Year %in% 2013:2019))
-NZ$exp <- NZmod$coefficients[1] + 
-  NZmod$coefficients[2] * NZ$Year
-Cexcess = round(sum(100 * NZ$Standth[21:23] / 
-                      NZ$exp[21:23]-100),2)
-Rexcess = round(sum(NZ$Standth[21:23]) - sum(NZ$exp[21:23]),0)
-NZ$Country = paste0(NZ$Country,"<P><P>Cummulative excess <span style='color:", six_cols[3],"'>",
-                    "2020 to 2022</span> ",
-                    Cexcess, "% (",Rexcess," people)")
-SEmod <- lm(Standth ~ Year, data=SWE |> 
-             filter(Year %in% 2013:2019))
-SWE$exp <- SEmod$coefficients[1] + 
-  SEmod$coefficients[2] * SWE$Year
-Cexcess = round(sum(100 * SWE$Standth[21:23] / 
-                      SWE$exp[21:23]-100),2)
-Rexcess = round(sum(SWE$Standth[21:23]) - sum(SWE$exp[21:23]),0)
-SWE$Country = paste0(SWE$Country,"<P><P>Cummulative excess <span style='color:", six_cols[3],"'>",
-                     "2020 to 2022</span> ",
[…]

### 2024-03-12 figure2.R

@@ -1,93 +1,0 @@
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-library(patchwork)
-
-ASMR <- read.csv("DMM_Standardised_Death_Rates.csv", skip=1, 
-                 col.names = c("Year","ASMR1K")) |> 
-  filter(!is.na(ASMR1K)) |> 
-  mutate(Year = as.numeric(Year)) |> 
-  filter(Year > 1990)
-
-longmod <- lm(ASMR1K ~ Year, data=ASMR |> filter(Year < 2017))
-ASMR$longExp <- longmod$coefficients[1] + 
-  ASMR$Year * longmod$coefficients[2]
-
-shortmod <- lm(ASMR1K ~ Year, data=ASMR |> filter(Year %in% 2010:2016))
-ASMR$shortExp <- shortmod$coefficients[1] + 
-  ASMR$Year * shortmod$coefficients[2]
-ASMR$shortExp[ASMR$Year < 2010] <- NA_real_
-
-xis<- c(2017,
-        2018,
-        2019,
-        2019,
-        2018,
-        2017)
-yis <- c(
-  3.180031,
-  3.073993,
-  2.967956,
-  3.351429,
-  3.406429,
-  3.461429)
-
-structural_over17_19 <- sum(ASMR$shortExp[ASMR$Year %in% 2017:2019] /
-  ASMR$longExp[ASMR$Year %in% 2017:2019])
-
-ggplot(ASMR |> filter(Year < 2020), 
-              aes(x=Year, y=ASMR1K))+
-  annotate("polygon",x=xis,y=yis,fill="skyblue")+
-  geom_point(size=0.3) +
-         geom_line(aes(y=longExp), linetype=2)+
-         geom_line(aes(y=shortExp), linetype=2) +
-  geom_line(data=ASMR |> filter(Year < 2017), aes(y=longExp)) +
-  geom_line(data=ASMR |> filter(Year < 2017),aes(y=shortExp)) +
-  theme_minimal(base_family="Open Sans",
-                base_size = 6) +
-  theme(
-    plot.background = element_rect(fill = "#FAFAFA", colour="#FAFAFA"),
-    panel.background = element_rect(fill = "#FFFFFF", colour = "#FFFFFF"),
-    panel.spacing.x = unit(9, "points"),
-    panel.spacing.y = unit(3, "points"),
-    panel.grid = element_blank(),
[…]

### 2024-03-12 figure4.R

@@ -1,87 +1,0 @@
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-library(patchwork)
-source("theme.R")
-six_cols <- colorblind_pal()(6)
-
-CMR <- read.csv("DMM_CDRdec.csv", skip=1, 
-                 col.names = c("Year","CDR1K")) |> 
-  filter(!is.na(CDR1K)) |> 
-  mutate(Year = as.numeric(Year)) 
-
-mod7y <- lm(CDR1K ~ Year, data=CMR |> filter(Year %in% 2013:2019))
-CMR$mod7y <- mod7y$coefficients[1] + 
-  CMR$Year * mod7y$coefficients[2]
-
-mod5y <- lm(CDR1K ~ Year, data=CMR |> filter(Year %in% 2015:2019))
-CMR$mod5y <- mod5y$coefficients[1] + 
-  CMR$Year * mod5y$coefficients[2]
-CMR$percent5year <- 100*CMR$CDR1K/CMR$mod5y-100
-CMR$percent7year <- 100*CMR$CDR1K/CMR$mod7y-100
-
-excessdf <- data.frame(Baseline = c("2015-2019", "2013-2019", "2015-2019", "2013-2019",
-                          "2015-2019", "2013-2019", "2015-2019", "2013-2019"),
-           Year = c("2020","2020","2021","2021","2022","2022","nett","nett"),
-           value = c(round(CMR$percent5year[CMR$Year == 2020],2),
-                     round(CMR$percent7year[CMR$Year == 2020],2),
-                     round(CMR$percent5year[CMR$Year == 2021],2),
-                     round(CMR$percent7year[CMR$Year == 2021],2),
-                     round(CMR$percent5year[CMR$Year == 2022],2),
-                     round(CMR$percent7year[CMR$Year == 2022],2),
-                     round(sum(CMR$percent5year[CMR$Year %in% 2020:2022]),2),
-                     round(sum(CMR$percent7year[CMR$Year %in% 2020:2022]),2)))
-innergraph <- ggplot(excessdf, aes(x=Year,y=Baseline, label=value))+
-  geom_tile(fill="white", colour="black") +geom_text() + theme_tufte()  +
-  scale_x_discrete(position = "top")+
-  labs(title="Excess as percent of expected", y="Baseline\n ",x="")
-  
-ggplot(CMR |> filter(Year %in% 2012:2023), 
-              aes(x=Year, y=CDR1K))+
[…]

### 2024-03-13 rawfiles/Draft_letter.qmd
first big update after asking for feedback. Thanks everyone across many plaforms

---
title: Draft Letter
author: David Hood
date: last-modified
format: gfm
---

Things David knows he still needs to do:

-   more proofreading.

-   standardised citation format.

-   clear links to source data.

-   standard figure look

*The timetable is: I am welcoming changes up to Thursday 21st and regularly updating the draft here until then. On Thursday 21st I will make what I expect to be any final changes. On Friday 22nd I will be distribution the final draft to people who have expressed an interest in co-signing, with an aim to send it to the journal early in the week beginning Monday 25th.*

To the editor and editorial board,

I am formally writing to point out the significant errors in the paper from Gibson (2024)[^1], published in your journal, as I feel both you and your readership should be aware of the problems before basing any decisions on the claims made. In particular, the paper asserts two claims:

[^1]: John Gibson (22 Feb 2024): Cumulative excess deaths in New Zealand in the COVID-19 era: biases from ignoring changes in population growth rates, New Zealand Economic Papers, DOI: 10.1080/00779954.2024.2314770

-   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of check names all[^2].
-   That Gibson's method, producing positive excess mortality, is more effective than Karlinsky & Kobak in determining mortality as it adjusts for population change due to immigration changes.

[^2]: Ariel Karlinsky, Dmitry Kobak (2021) Tracking excess mortality across countries during the COVID-19 pandemic with the World Mortality Dataset eLife 10:e69336. Code available from https://github.com/dkobak/excess-mortality

Neither of these claims are accurate.

# The underpinning of New Zealand's mortality measurement

For centuries, it has been established practice among actuaries, demographers, and health researchers to measure mortality using death rates by age where such data is available[^3]. This method superseded Crude Mortality Rates, where CMR takes into account total population change in comparisons by dividing the total number of deaths by the total number of people available to die, improving CMR by acknowledging that different ages have different risks of death, so comparing age-based risk of death on a uniform basis. In New Zealand, tutorials on making mortality comparisons by using death rates by age and standardising to a reference population have been published since the 1890s[^4].

[^3]: Price, Richard (1773). Observations on Reversionary Payments. United Kingdom: (n.p.).

[^4]: Report on the Statistics of New Zealand, 1890. Published by George Didsbury, Government Printer, Wellington., 1891. https://www3.stats.govt.nz/historic_publications/1890-official-handbook/1890-official-handbook.html Age standardised mortality is in the section "Adjusted death-rate a better means of comparison".

Age Standardised Mortality directly adjusts for both population growth and aging, as it is based on the number of deaths of the current resident population divided by the living current resident population, stratified by age group and time period. This directly measures the population change effects Gibson is trying to model.

While calculation of New Zealand's cumulative excess mortality for 2020-2022 can vary by a couple of percentage points depending on the exact method, using a death rates by age (standardised) method the Institute of Actuaries finds New Zealand's excess mortality from week 10 2020 to the end of 2022 to be -4% (that there are 4% fewer deaths than would be expected based on pre-COVID-19 trends)[^5].

[^5]: Actuaries Institute: CMI Working Paper 180 https://www.actuaries.org.uk/learn-and-develop/continuous-mortality-investigation/cmi-working-papers/mortality-projections/cmi-working-paper-180

The Our World in Data graph, and the Karlinsky and Kobak method that underlies it is an approximation (if it is a better or worse approximation to Gibson's method will be discussed later). While recognising the Our World In Data graph is an approximation, there are still several practically useful implicit features about it.

-   Being based on trends of deaths, it enables a common basis comparison with countries that lack age stratified current population information
-   It makes weekly comparisons available.
-   Able to be checked by any member of the public for the latest current worldwide information. This last is a function of Our World in Data presentation rather than the method, but it is still a practical benefit.

While the existence of centuries of mortality measurement (not mentioned in Gibson's article) does not directly address if New Zealand medical professionals are aware that best practice mortality calculations recognise that people of different ages die at different rates when using the publicly available Our World in Data graphs in media pieces, it does establish that there is an established methodology, not mentioned in Gibson's article, underpinning claims of New Zealand's negative excess, and that method has been used for centuries, world-wide, across multiple disciplines. We can check if New Zealand medical professionals are aware of this by asking them: *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ’s are!)"*(Ashley Bloomfield) [^6]

[^6]: Ashley Bloomfield, 2024, personal communication

## Age Standardised Mortality Review

As Age Standardised Mortality is a standard across multiple disciplines, but is not raised in the Gibson paper, a brief review is in order.

Death rates for an age stratified group, for a time period in a given location, measure the rate at which that age group dies. These can be directly compared to death rates in other time periods and locations to measure if the group experience is more or less deadly. As it is a rate, it automatically adjusts for change in the population size of the age group.

While this enables direct comparisons of people of the same age, it does not enable total population comparisons (as total populations may have different age structures). Standardisation with a reference population addresses this by assessing the total population deaths on the basis of death rates by age if the places and times being compared had the same age structure.

*Figure 1*

![ASMR steps](md_figures/figure1.png){fig-alt="ASMR process" fig-align="center"}

Some variation in calculation results can be caused by the choice of standard population, due to the variation of young and old within a population. However, pragmatically, any population that can be argued to be analytically useful should produce similar results. Using single year ages groups by sex to 94 then 95+, and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -5% for 2020 to 2022. Using a standard reference population of 2021 when the borders were restricted gives a cumulative excess of -6%. Using a pre-COVID-19 standard population of 2019 gives a cumulative excess of -5%. Using the standard population of 1961, used by Stats NZ to maintain long-term continuity, cumulative excess is -5%.

### Baselines and excess mortality

Excess mortality is the difference to what is expected for mortality, so excess calculations using any method are sensitive what is considered normal: the baseline. For Age Standardised Mortality the regular process is to use a linear trend of preceding years, and in the case of a major multi-year event like a pandemic, the linear trend of pre-pandemic years.

As New Zealand, no different to most other advanced economies, has seen the rate of decrease in mortality slow over time, the length of the linear trend baseline period should be short enough not to be introducing error as a result of applying a straight line of best fit to curved data.

![baseline](md_figures/figure2.png){fig-alt="Overlong baseline error" fig-align="center"}

Based on the influenza season patterns in recent pre-pandemic years, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline[^7]. New Zealand has seen a very similar history in this period to Australia, with the added limitation that incorporating 2011, the year of the Christchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is a poor influence point to use in a baseline as it is not a re-occurring annual event.

[^7]: Australian Bureau of Statistics. https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology

While many international comparisons use a 5 year baseline range to increase the number of countries for which weekly level data is available, in practice the results for New Zealand do not differ substantively between a 5 (2015-2019) and 7 (2013-2019) year baseline period. A 5 year linear baseline and a 7 year baseline both give a cumulative excess for 2020-2022 of -5% when using a 2022 standard population.

## Comparison of Gibson and Karlinsky & Kobak

Starting from a commonly recognised best practice measure of mortality, age standised mortality, giving a cumulative excess mortality for 2020-2022 in the range of -4% to -6% depending on analytic assumptions, we can compare other methods to that.

The Our World in Data cumulative excess (Projection based on Karlinsky & Kobak) gives cumulative excess mortality for 2020-2022 as 0%.

Gibson's Population adjusted for changes in growth rates model gives an excess mortality for 2020-2022 range of 1.7% to 5.3% with the central estimate rounding to 4%.

Though not mentioned in the paper, adding the Population as a model term to a death based model would normally be functionally equivalent to using the commonly known Crude Mortality Rate (deaths among the total population / total living population). A 2015-2019 linear regression of the annual CMR gives an excess of 1%.

Gibson's method produces results further from actuarial standard methods than the Karlinsky & Kobak method. The reason for Gibson's model performing so poorly compared to Karlinsky & Kobak lies in implicitly misapplying death rates by age. Young people frequently migrate and infrequently die. Old people infrequently migrate and frequently die. Applying a correction based on the amount of migration to the mortality of the total population (dominated by people not migrating) introduces unnecessary structural error.

![Comparison of ages of migration and death](md_figures/immigration.png){fig-alt="Young people migrate, old people die" fig-align="center"}

Conversely, because the Karlinsky & Kobak method only uses death data, it implicitly limits itself to the population that dies. The population mainly dying are not migrating, so the model does not diverge as far from ASM as the Gibson method when border changes cause major fluctuations in migration.

Gibson asserted that it was not important to incorporate aging:

*If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process.*

But, it does show up before COVID-19. If you do exactly the same type of analysis, but before covid and the arguments around closing the border, of comparing the crude death rate (which ignores aging but adjusts for total population size) of the 3 years (2017-2019) after the 5 year baseline (2012-2016) to the expected values of the baseline, then the most recent three years are a cumulative 10% above baseline.

![Even if all years are before COVID-19, the post-baseline years are above trend](md_figures/figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}

I suggest anyone interested in this topic go explore population ages and death rates at Stats NZ themselves, and while there have a look at New Zealand death rates by age, or indeed the public, downloadable, standardised death rates. And then ask yourself, given the ready availability of high quality data, is the best default course of action to use the source that automatically compensates for the most confounders?

### 2024-03-13 rawfiles/NZfourpane.R

@@ -1,266 +1,0 @@
-library(dplyr)
-library(tidyr)
-
-Dth <- read.csv("VSD_Death_by_Age_Dec.csv", skip=2)
-Dth00_59 <- Dth |> select(Year = X., Less.than.1.year:X59.years) |> 
-  filter(!is.na(X50.years)) |> 
-  mutate(Year = as.numeric(Year)) |> 
-  gather(key="Age",value="Dths", -Year) |> 
-  summarise(.by=Year, Deaths = sum(Dths)) |> 
-  filter(Year >2012) |> 
-  mutate(dataset = "Resident Deaths",
-         Age = "age 0 to 59\n33% of deaths") |> 
-  rename(Number = Deaths)
-Dth60_79 <- Dth |> select(Year = X., X60.years:X79.years) |> 
-  filter(!is.na(X70.years)) |> 
-  mutate(Year = as.numeric(Year)) |> 
-  gather(key="Age",value="Dths", -Year) |> 
-  summarise(.by=Year, Deaths = sum(Dths)) |> 
-  filter(Year >2012) |> 
-  mutate(dataset = "Resident Deaths",
-         Age = "age 60 to 79\n27% of deaths") |> 
-  rename(Number = Deaths)
-Dth80_plus <- Dth |> select(Year = X., X80.years:X100.years.and.over.1) |> 
-  filter(!is.na(X85.years)) |> 
-  mutate(Year = as.numeric(Year)) |> 
-  gather(key="Age",value="Dths", -Year) |> 
-  summarise(.by=Year, Deaths = sum(Dths)) |> 
-  filter(Year >2012) |> 
-  mutate(dataset = "Resident Deaths",
-         Age = "age 80 plus\n40% of deaths") |> 
-  rename(Number = Deaths)
-Tdth00 <- sum(Dth00_59$Deaths)
-Tdth60 <- sum(Dth60_79$Deaths)
-Tdth80 <- sum(Dth80_plus$Deaths)
-Tdth80/(Tdth00 + Tdth60 + Tdth80)
-
-Pop <- read.csv("DPE_Mean_Pop_by_Age_Dec.csv", skip=3)
-Pop00_59 <- Pop |> select(Year = X., X0.Years:X59.Years) |> 
-  filter(!is.na(X50.Years)) |> 
-  mutate(Year = as.numeric(Year)) |> 
-  gather(key="Age",value="pop", -Year) |> 
[…]

### 2024-03-13 rawfiles/fig1files/theme.R

@@ -1,0 +1,28 @@
+six_cols <- colorblind_pal()(6)
+footer <- paste0("\nGraph made ", trimws(format(Sys.time(), format="%e %b %Y")),". Graphs: thoughtfulnz.quarto.pub/nzcovidreport/ Contact: thoughtfulnz on mastodon.nz or bsky.app\n")
+
+theme_davidhood <- function(){
+  theme_minimal(base_family="Open Sans",
+                base_size = 9) %+replace%  
+    theme(axis.line.x = element_line(linewidth=0.1),
+          axis.line.y = element_line(linewidth=0.1),
+          axis.ticks = element_line(linewidth=0.2),
+          panel.grid = element_blank(),
+          strip.background = element_rect(fill= "#FFFFFF", colour="#EFEFEF"),
+          strip.text = element_text(margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")),
+          strip.placement = "inside",
+          panel.background = element_rect(fill = "#FFFFFF", colour = "#FFFFFF"),
+          panel.spacing = unit(1.5, "lines"),
+          plot.title = element_text(lineheight = 1.18, size=11,
+                                    margin=margin(t = 5, r = 15, b = 5, l = 15, unit = "pt"),
+                                    hjust=1, vjust=0),
+          plot.subtitle = element_text(lineheight = 1.18, size=7,
+                                    margin=margin(t = 0, r = 15, b = 5, l = 15, unit = "pt"),
+                                    hjust=0),
+          plot.background = element_rect(fill = "#FCFCFC", colour="#FCFCFC"),
+          plot.caption = element_text(margin=margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
+                                      lineheight = 1.15,
+                                      size=7, hjust=1),
+          plot.caption.position = "plot")
+  
+}

### 2024-03-13 rawfiles/Draft_letter.qmd
first big update after asking for feedback. Thanks everyone across many plaforms
lunchtime Thursday- an added paragraph and a little caption rewriting

@@
 
-Things David knows he still needs to do:
+Things David knows he still needs to do (that don't alter overall content):
 
@@
 
--   standardised citation format.
+-   standard citation format.
 
@@
 -   standard figure look
+
+-   Captions under graphs.
 
@@
 
--   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of check names all[^2].
+-   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of Karlinsky & Kobak[^2].
 -   That Gibson's method, producing positive excess mortality, is more effective than Karlinsky & Kobak in determining mortality as it adjusts for population change due to immigration changes.
@@
 
-The Our World in Data graph, and the Karlinsky and Kobak method that underlies it is an approximation (if it is a better or worse approximation to Gibson's method will be discussed later).
-While recognising the Our World In Data graph is an approximation, there are still several practically useful implicit features about it.
+The Our World in Data graph, and the Karlinsky and Kobak method that underlies it, is an approximation (if it is a better or worse approximation to Gibson's method will be discussed later).
+While recognising the Our World In Data graph is an approximation, there are still several practically useful implicit features about it:
 
@@
 While the existence of centuries of mortality measurement (not mentioned in Gibson's article) does not directly address if New Zealand medical professionals are aware that best practice mortality calculations recognise that people of different ages die at different rates when using the publicly available Our World in Data graphs in media pieces, it does establish that there is an established methodology, not mentioned in Gibson's article, underpinning claims of New Zealand's negative excess, and that method has been used for centuries, world-wide, across multiple disciplines.
-We can check if New Zealand medical professionals are aware of this by asking them: *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ’s are!)"*(Ashley Bloomfield) [^6]
+We can check if New Zealand medical professionals are aware of this by asking them: *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ’s are!)"*(Ashley Bloomfield) [^6].
 
@@
 However, pragmatically, any population that can be argued to be analytically useful should produce similar results.
-Using single year ages groups by sex to 94 then 95+, and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -5% for 2020 to 2022.
+Using single year ages groups by sex to 94 then 95+[^7], and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -5% for 2020 to 2022.
 Using a standard reference population of 2021 when the borders were restricted gives a cumulative excess of -6%.
@@
 Using the standard population of 1961, used by Stats NZ to maintain long-term continuity, cumulative excess is -5%.
+
+[^7]: NZ annual deaths are available from <https://infoshare.stats.govt.nz> in Population : Deaths : Deaths by age and sex (Annual-Dec).
+NZ mean annual population is available from Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec) : Mean year ended.
 
@@
 
-![baseline](md_figures/figure2.png){fig-alt="Overlong baseline error" fig-align="center"}
+![Figure 2: Structural error introduced by overly long baselines.
+Longterm (age) standardised death rates from https://infoshare.stats.govt.nz in Population : Death Rates - DMM : Standardised death rates (Maori and total population) (Annual-Dec) : Total Population](md_figures/figure2.png){fig-alt="Overlong baseline error" fig-align="center"}
 
-Based on the influenza season patterns in recent pre-pandemic years, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline[^7].
-New Zealand has seen a very similar history in this period to Australia, with the added limitation that incorporating 2011, the year of the Christchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is a poor influence point to use in a baseline as it is not a re-occurring annual event.
+Based on the influenza season patterns in recent pre-pandemic years, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline[^8].
+New Zealand has seen a very similar history in this period to Australia, with the added limitation that using 2011, the year of the Christchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is an extremely poor choice to include in baselines as an earthquake hitting the second largest city and people living in broken housing through the winter is not a reoccuring annual event.
 
-[^7]: Australian Bureau of Statistics. https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology
+[^8]: Australian Bureau of Statistics. https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology
 
@@
 
-![Even if all years are before COVID-19, the post-baseline years are above trend](md_figures/figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}
+![Even if all years are before COVID-19, the post-baseline years are above trend](md_figures/figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"} Of more surprise is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the accepted method of comparing death rates while ignoring the effects of age.
+While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same answers, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of the regression.
+But, compared to the differences with methods acknowledging age, such problems are minor.
 

### 2024-03-14 rawfiles/Draft_letter.qmd
replaced figure 3 with individual ages rather than 5 year steps

@@
 While the existence of centuries of mortality measurement (not mentioned in Gibson's article) does not directly address if New Zealand medical professionals are aware that best practice mortality calculations recognise that people of different ages die at different rates when using the publicly available Our World in Data graphs in media pieces, it does establish that there is an established methodology, not mentioned in Gibson's article, underpinning claims of New Zealand's negative excess, and that method has been used for centuries, world-wide, across multiple disciplines.
-We can check if New Zealand medical professionals are aware of this by asking them: *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ’s are!)"*(Ashley Bloomfield) [^6].
+We can check if New Zealand medical professionals are aware of this by asking them: *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ's are!)"*(Ashley Bloomfield) [^6].
 
@@
 
-![Comparison of ages of migration and death](md_figures/immigration.png){fig-alt="Young people migrate, old people die" fig-align="center"}
+![Figure 3: Ages of migration and ages of death have little overlap](md_figures/figure3.png){fig-alt="Young people migrate, old people die" fig-align="center"}
 
@@
 
-![Even if all years are before COVID-19, the post-baseline years are above trend](md_figures/figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"} Of more surprise is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the accepted method of comparing death rates while ignoring the effects of age.
+![Figure 4: Applying the same test to pre-covid years shows the same aging population effects](md_figures/figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}
+
+Of more surprise is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the accepted method of comparing death rates while ignoring the effects of age.
 While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same answers, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of the regression.

### 2024-03-14 rawfiles/Fig2_baseline.R

@@ -1,0 +1,59 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+source("theme.R")
+
+ASMR <- read.csv("StatsNZData/DMM_SDR_TotPop.csv", skip=1, 
+                 col.names = c("Year","ASMR1K")) |> 
+  filter(!is.na(ASMR1K)) |> 
+  mutate(Year = as.numeric(Year)) |> 
+  filter(Year > 1990)
+
+longmod <- lm(ASMR1K ~ Year, data=ASMR |> filter(Year < 2017))
+ASMR$longExp <- longmod$coefficients[1] + 
+  ASMR$Year * longmod$coefficients[2]
+
+shortmod <- lm(ASMR1K ~ Year, data=ASMR |> filter(Year %in% 2010:2016))
+ASMR$shortExp <- shortmod$coefficients[1] + 
+  ASMR$Year * shortmod$coefficients[2]
+ASMR$shortExp[ASMR$Year < 2010] <- NA_real_
+
+xis<- c(2017,
+        2018,
+        2019,
+        2019,
+        2018,
+        2017)
+yis <- c(
+  3.180031,
+  3.073993,
+  2.967956,
+  3.351429,
+  3.406429,
+  3.461429)
+
+structural_over17_19 <- sum(ASMR$shortExp[ASMR$Year %in% 2017:2019] /
+  ASMR$longExp[ASMR$Year %in% 2017:2019])
+
+ggplot(ASMR |> filter(Year < 2020), 
+              aes(x=Year, y=ASMR1K))+
+  annotate("polygon",x=xis,y=yis,fill=six_cols[3])+
+  geom_point(size=0.3) +
+         geom_line(aes(y=longExp), linetype=2)+
+         geom_line(aes(y=shortExp), linetype=2) +
+  geom_line(data=ASMR |> filter(Year < 2017), aes(y=longExp)) +
+  geom_line(data=ASMR |> filter(Year < 2017),aes(y=shortExp)) +
+  theme_davidhood() +
+  theme(plot.title = element_markdown(lineheight = 1.18, size=11,
+                                      margin=margin(t = 5, r = 15, b = 5, l = 0, unit = "pt"),
+                                      hjust=0, vjust=0)) +
+  labs(title=paste0("<span style='color:", six_cols[3],"'>Structural Error</span> from an overlong baseline period in predicting 2017-2019"),
[…]

### 2024-03-14 rawfiles/Fig3_immigration.R

@@ -1,0 +1,71 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+source("theme.R")
+six_cols <- colorblind_pal()(6)
+
+VSD <- read.csv("StatsNZData/VSD_DeathsDec_age_sex.csv",
+                skip=1, col.names = c("Year","Age",
+                                      "Male","Female")) |> 
+  filter(!is.na(Male)) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
+  fill(Year) |> 
+  mutate(Year = as.numeric(Year)) |> 
+  filter(Year == 2019)|> 
+  slice(26:126) |> 
+  mutate(Deaths = Male + Female,
+         Age = ifelse(Age == "Less than 1 year",
+                      "0 years",Age),
+         AgeN = as.numeric(gsub(" yea.*","",Age)),
+         AgeN = ifelse(AgeN > 89, 90, AgeN)) |> 
+  summarise(.by=AgeN, Deaths=sum(Deaths))
+
+DPE <- read.csv("StatsNZData/DPE_PopDecMean_age_sex.csv",
+                skip=2, col.names = c("Year","Age",
+                                     "Male","Female"))  |> 
+  filter(!is.na(Male)) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
+  fill(Year) |> 
+  mutate(Year = as.numeric(Year)) |> 
+  filter(Year == 2019) |> 
+  slice(c(1:95,134)) |> 
+  mutate(Lives = Male + Female,
+         AgeN = as.numeric(gsub(" Yea.*","",Age)),
+         AgeN = ifelse(AgeN > 89, 90, AgeN)) |> 
+  summarise(.by=AgeN, Lives=sum(Lives))
+
+ITM <- read.csv("StatsNZData/ITM_2019AgeSex.csv", skip=4, 
+                col.names = c("Age","Female","Male")) |> 
+  slice(c(1:90,127)) |> 
+  mutate(Moves = as.numeric(Male) + as.numeric(Female),
+         AgeN = as.numeric(gsub(" Yea.*","",Age)),
+         AgeN = ifelse(AgeN > 89, 90, AgeN)) |> 
+  summarise(.by=AgeN, Moves=sum(Moves))
+
+Deaths <- VSD |> 
+  inner_join(DPE, by = join_by(AgeN)) |> 
[…]

### 2024-03-14 rawfiles/theme.R

@@ -7,16 +7,18 @@ theme_davidhood <- function(){
     theme(axis.line.x = element_line(linewidth=0.1),
           axis.line.y = element_line(linewidth=0.1),
           axis.ticks = element_line(linewidth=0.2),
-          panel.grid = element_blank(),
+          axis.title.y.left = element_text(margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")),
           strip.background = element_rect(fill= "#FFFFFF", colour="#EFEFEF"),
           strip.text = element_text(margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")),
           strip.placement = "inside",
           panel.background = element_rect(fill = "#FFFFFF", colour = "#FFFFFF"),
-          panel.spacing = unit(1.5, "lines"),
+          panel.spacing.x = unit(9, "points"),
+          panel.spacing.y = unit(3, "points"),
+          panel.grid = element_blank(),
           plot.title = element_text(lineheight = 1.18, size=11,
-                                    margin=margin(t = 5, r = 15, b = 5, l = 15, unit = "pt"),
-                                    hjust=1, vjust=0),
-          plot.subtitle = element_text(lineheight = 1.18, size=7,
+                                    margin=margin(t = 5, r = 0, b = 5, l = 15, unit = "pt"),
+                                    hjust=0, vjust=0),
+          plot.subtitle = element_text(lineheight = 1.18, size=8,
                                     margin=margin(t = 0, r = 15, b = 5, l = 15, unit = "pt"),
                                     hjust=0),
           plot.background = element_rect(fill = "#FCFCFC", colour="#FCFCFC"),

### 2024-03-14 rawfiles/Draft_letter.qmd
replaced figure 4 with straight up graph of old people

@@
 
-But, it does show up before COVID-19.
-If you do exactly the same type of analysis, but before covid and the arguments around closing the border, of comparing the crude death rate (which ignores aging but adjusts for total population size) of the 3 years (2017-2019) after the 5 year baseline (2012-2016) to the expected values of the baseline, then the most recent three years are a cumulative 10% above baseline.
+The assumption that aging is a long term process so should show up as a long slow process only applies if there are no sudden demographic shifts among age groups that alter the risk of death.
+The number of 75 and older residents, when compared to the 2012-2016 trend, and that trend can be seen increasing in 2019.
+This matches a "sudden" demographic shift of a spike in the age distribution that began decades earlier and naturally aged into ranges that play a more significant role in total deaths.
 
-![Figure 4: Applying the same test to pre-covid years shows the same aging population effects](md_figures/figure5.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}
+![Figure 4: The rate of increase in elderly accelerated in 2019](md_figures/figure4.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}
 

### 2024-03-14 rawfiles/Fig3_immigration.R

@@ -62,7 +62,7 @@ ggplot(aes(x=AgeN, y=percent, fill=series))+
   scale_fill_manual(values=six_cols, name="Series:")+
   scale_x_continuous(breaks=c(0,30,60,90),
                      labels=c("under 1","30","60","90+")) +
-  labs(title="2019 Deaths vs. Migrants as percentage of resident population",
+  labs(title="2019 Deaths vs. Migrants\nas percentage of resident population",
        y="Percentage", x="Age",
        caption="Source: infoshare.stats.govt.nz")
 ggsave(filename="md_figures/figure3.png",

### 2024-03-14 rawfiles/Fig4_75plus.R

@@ -1,0 +1,42 @@
+# table comparing CDR and ASMR predictions
+
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+source("theme.R")
+six_cols <- colorblind_pal()(6)
+
+autoAge <- paste(0:94,"Years")
+autoAge <- c(autoAge,"95 Years and Over")
+DPE <- read.csv("StatsNZData/DPE_PopDecMean_age_sex.csv", skip=2,
+                col.names = c("Year","Age","Male","Female")) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
+  fill(Year) |> 
+  filter(Age %in% autoAge) |> 
+  mutate(AgeN = as.numeric(gsub(" .*","",Age)),
+         Year = as.numeric(Year)) |> 
+  gather(key="Sex",value="MeanPopulation", Male:Female) |> 
+  filter(AgeN >=75) |> 
+  summarise(.by=Year,
+            Living75plus = sum(MeanPopulation))
+
+LifeModel <- lm(Living75plus ~ Year, data=DPE |> filter(Year %in% 2012:2016))
+DPE$Expected <- LifeModel$coefficients[1] + LifeModel$coefficients[2]*DPE$Year
+
+ggplot(DPE, aes(x=Year, y=Living75plus))+
+  geom_point() +
+  geom_line(aes(y=Expected), colour=six_cols[2], linetype=2) +
+  geom_line(data=DPE |> filter(Year < 2017), aes(y=Expected), colour=six_cols[2]) +
+  theme_davidhood() +
+  scale_x_continuous(breaks=c(2012,2016,2020))+
+  labs(title="Actual 75 years and older vs 2012-2016 trend",
+       y="Residents 75 years and Older", x="Year",
+       caption="Source: infoshare.stats.govt.nz")
+ggsave(filename="md_figures/figure4.png",
+       height=4, width = 6, dpi=300, units = "in", bg = "white")
+ggsave(filename="../md_figures/figure4.png",
+       height=4, width = 6, dpi=300, units = "in", bg = "white")
+

### 2024-03-14 rawfiles/table1_CDRvsASMR.R

@@ -1,0 +1,57 @@
+# table comparing CDR and ASMR predictions
+
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+library(patchwork)
+library(forcats)
+
+autoAge <- paste(2:99,"years")
+autoAge <- c("Less than 1 year","1 year",autoAge,"100 years and over")
+VSD <- read.csv("StatsNZData/VSD_DeathsDec_age_sex.csv", skip=1,
+                col.names = c("Year","Age","Male","Female")) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
+  fill(Year) |> 
+  filter(Age %in% autoAge) |> 
+  group_by(Year, Age) |> 
+  slice(1) |> 
+  ungroup() |> 
+  mutate(Age = ifelse(Age == "Less than 1 year","0 years",Age),
+         AgeN = as.numeric(gsub(" .*","",Age)),
+         Year = as.numeric(Year)) |> 
+  gather(key="Sex",value="Deaths", Male:Female) |> 
+  mutate(AgeN = ifelse(AgeN > 94,95, AgeN)) |> 
+  summarise(.by=c(Year,Sex,AgeN),
+            Deaths = sum(Deaths))
+
+
+autoAge <- paste(0:94,"Years")
+autoAge <- c(autoAge,"95 Years and Over")
+DPE <- read.csv("StatsNZData/DPE_PopDecMean_age_sex.csv", skip=2,
+                col.names = c("Year","Age","Male","Female")) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
+  fill(Year) |> 
+  filter(Age %in% autoAge) |> 
+  mutate(AgeN = as.numeric(gsub(" .*","",Age)),
+         Year = as.numeric(Year)) |> 
+  gather(key="Sex",value="MeanPopulation", Male:Female) |> 
+  select(Year, AgeN, Sex, MeanPopulation)
+
+Stnd <- DPE |> filter(Year == 2018) |> 
+  select(AgeN, Sex, StndPopulation = MeanPopulation)
+
+totalStndPop = sum(Stnd$StndPopulation)
+
+ASMR <- VSD |> 
+  inner_join(DPE, by = join_by(Year, Sex, AgeN)) |> 
+  inner_join(Stnd, by = join_by(Sex, AgeN)) |> 
[…]

### 2024-03-16 rawfiles/Draft_letter.qmd
tweaking Richard's wording suggestion

@@
 
-I am formally writing to point out the significant errors in the paper from Gibson (2024)[^1], published in your journal, as I feel both you and your readership should be aware of the problems before basing any decisions on the claims made.
+I am formally writing to point out the significant errors in the paper from Gibson (2024)[^1], published in your journal.
+It is important both you and your readership are aware of its shortcomings before basing any decisions on the claims made.
 In particular, the paper asserts two claims:

### 2024-03-19 rawfiles/Draft_letter.qmd
suggestions from Monday and Tuesday

@@
 
-[^1]: John Gibson (22 Feb 2024): Cumulative excess deaths in New Zealand in the COVID-19 era: biases from ignoring changes in population growth rates, New Zealand Economic Papers, DOI: 10.1080/00779954.2024.2314770
+[^1]: John Gibson (22 Feb 2024): Cumulative excess deaths in New Zealand in the COVID-19 era: biases from ignoring changes in population growth rates, New Zealand Economic Papers, DOI: 10.1080/00779954.2024.2314770 available from https://doi.org/10.1080/00779954.2024.2314770
 
@@
 
--   Being based on trends of deaths, it enables a common basis comparison with countries that lack age stratified current population information
+-   Being based on trends of deaths, it enables a common, consistent basis comparison with countries that lack age stratified current population information.
+As Gibson notes, Karlinsky and Kobak were initially able to provide comparison data for 103 countries (later 127).
+By contrast the mortality.org has weekly death rates by age (suitable for age standard mortality comparisons) for 38 countries[^6].
 -   It makes weekly comparisons available.
@@
 
+[^6]: <https://mortality.org/File/GetDocument/Public/STMF/DOC/STMFmetadata.pdf>
+
 While the existence of centuries of mortality measurement (not mentioned in Gibson's article) does not directly address if New Zealand medical professionals are aware that best practice mortality calculations recognise that people of different ages die at different rates when using the publicly available Our World in Data graphs in media pieces, it does establish that there is an established methodology, not mentioned in Gibson's article, underpinning claims of New Zealand's negative excess, and that method has been used for centuries, world-wide, across multiple disciplines.
-We can check if New Zealand medical professionals are aware of this by asking them: *"the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ's are!)"*(Ashley Bloomfield) [^6].
+We can check if New Zealand medical professionals are aware of this by asking them.
 
-[^6]: Ashley Bloomfield, 2024, personal communication
+> "*the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ's are!)*"(Ashley Bloomfield) [^7].
+
+[^7]: Ashley Bloomfield, 2024, personal communication
 
@@
 
-*Figure 1*
-
-![ASMR steps](md_figures/figure1.png){fig-alt="ASMR process" fig-align="center"}
+The total of standard deaths can optionally be divided by the standard population to make age standardised mortality rates.
 
@@
 However, pragmatically, any population that can be argued to be analytically useful should produce similar results.
-Using single year ages groups by sex to 94 then 95+[^7], and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -5% for 2020 to 2022.
+Using single year age death rates by sex to 94 then 95+[^8], and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -5% for 2020 to 2022.
 Using a standard reference population of 2021 when the borders were restricted gives a cumulative excess of -6%.
@@
 
-[^7]: NZ annual deaths are available from <https://infoshare.stats.govt.nz> in Population : Deaths : Deaths by age and sex (Annual-Dec).
+[^8]: NZ annual deaths are available from <https://infoshare.stats.govt.nz> in Population : Deaths : Deaths by age and sex (Annual-Dec).
 NZ mean annual population is available from Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec) : Mean year ended.
-
-### Baselines and excess mortality
 
@@
 
-![Figure 2: Structural error introduced by overly long baselines.
-Longterm (age) standardised death rates from https://infoshare.stats.govt.nz in Population : Death Rates - DMM : Standardised death rates (Maori and total population) (Annual-Dec) : Total Population](md_figures/figure2.png){fig-alt="Overlong baseline error" fig-align="center"}
+Based on the influenza season patterns in recent pre-pandemic years, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline[^9].
+New Zealand has seen a very similar history in this period to Australia, with the added limitation that using 2011, the year of the Christchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is an extremely poor choice to include in baselines as such an event is not a reoccurring annual one.
 
-Based on the influenza season patterns in recent pre-pandemic years, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline[^8].
-New Zealand has seen a very similar history in this period to Australia, with the added limitation that using 2011, the year of the Christchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is an extremely poor choice to include in baselines as an earthquake hitting the second largest city and people living in broken housing through the winter is not a reoccuring annual event.
+[^9]: Australian Bureau of Statistics. https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology
 
-[^8]: Australian Bureau of Statistics. https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology
-
-While many international comparisons use a 5 year baseline range to increase the number of countries for which weekly level data is available, in practice the results for New Zealand do not differ substantively between a 5 (2015-2019) and 7 (2013-2019) year baseline period.
+While many international covid comparisons use a 5 year baseline range to increase the number of countries for which weekly level data is available, in practice the results for New Zealand do not differ substantively between a 5 (2015-2019) and 7 (2013-2019) year baseline period.
 A 5 year linear baseline and a 7 year baseline both give a cumulative excess for 2020-2022 of -5% when using a 2022 standard population.
@@
 
-Starting from a commonly recognised best practice measure of mortality, age standised mortality, giving a cumulative excess mortality for 2020-2022 in the range of -4% to -6% depending on analytic assumptions, we can compare other methods to that.
+Starting from a commonly recognised best practice measure of mortality, age standised mortality, which gives a cumulative excess mortality for 2020-2022 in the range of -4% to -6%, we can compare other methods.
 
@@
 
-Though not mentioned in the paper, adding the Population as a model term to a death based model would normally be functionally equivalent to using the commonly known Crude Mortality Rate (deaths among the total population / total living population).
+Though not mentioned in the paper, using Population as a model term to a death based model would normally be functionally equivalent to using the commonly known Crude Mortality Rate (deaths among the total population / total living population).
 A 2015-2019 linear regression of the annual CMR gives an excess of 1%.
@@
 Old people infrequently migrate and frequently die.
-Applying a correction based on the amount of migration to the mortality of the total population (dominated by people not migrating) introduces unnecessary structural error.
+Applying a correction based on the amount of migration to the mortality of the total population (dominated by people not migrating) introduces structural error.
 
-![Figure 3: Ages of migration and ages of death have little overlap](md_figures/figure3.png){fig-alt="Young people migrate, old people die" fig-align="center"}
+![Figure 1: Ages of migration and ages of death have little overlap](md_figures/figure3.png){fig-alt="Young people migrate, old people die" fig-align="center"}
 
 Conversely, because the Karlinsky & Kobak method only uses death data, it implicitly limits itself to the population that dies.
-The population mainly dying are not migrating, so the model does not diverge as far from ASM as the Gibson method when border changes cause major fluctuations in migration.
+The population mainly dying are not migrating, so the model does not diverge as far from age standardised mortality as the Gibson method when border changes cause major fluctuations in migration.
+
+If 2020 had the same single year of age death rates as 2019, and new migrant 2020 death rates were the same as the equivalently aged non-migratory population in 2019, then we can calculate the expected deaths from the 35,744 reduction in nett migrationfrom 2019 to 2020.
+Using single age death rates the youthful migrating population would be expected to result in 85 fewer deaths for the 35,744 reduction in total population.
+
+Conversely, the expected increase in deaths due to aging can be calculated by aging the 2019 resident population by one year, and subtracting annual deaths of that age, and applying the single year death rates of 2019 (the same assumption as deaths expected from migration).
+For the resident population, there would be an expected rise of 837 deaths while the total population decreased by about 34,000.
+
+Gibson's decision not to apply total population but not age to the model excludes an increase in expected deaths of an approximate order of magnitude greater than the effects of migration.
+But all these complexities are avoided by incorporating both migration and aging by using age standardised mortality.
 
@@
 
-*If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process.*
+> *If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process.*
 
 The assumption that aging is a long term process so should show up as a long slow process only applies if there are no sudden demographic shifts among age groups that alter the risk of death.
-The number of 75 and older residents, when compared to the 2012-2016 trend, and that trend can be seen increasing in 2019.
+The number of 75 and older residents, when compared to the 2012-2016 trend, was increasing in 2019.
 This matches a "sudden" demographic shift of a spike in the age distribution that began decades earlier and naturally aged into ranges that play a more significant role in total deaths.
 
-![Figure 4: The rate of increase in elderly accelerated in 2019](md_figures/figure4.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}
+![Figure 2: The rate of increase in elderly accelerated in 2019](md_figures/figure4.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}
 
-Of more surprise is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the accepted method of comparing death rates while ignoring the effects of age.
-While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same answers, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of the regression.
+Of more surprise is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the normal method of comparing death rates while ignoring the effects of age.
+While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same answers, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of Karlinsky & Kobak.
 But, compared to the differences with methods acknowledging age, such problems are minor.
 
-I suggest anyone interested in this topic go explore population ages and death rates at Stats NZ themselves, and while there have a look at New Zealand death rates by age, or indeed the public, downloadable, standardised death rates.
+Gibson's results are also at variance with other countries statistical authorities, the Australian Bureau of Statistic's age based mortality analysis has excess mortality of 3.4% across 2020-2022[^10].
+Karlinsky & Kobak's approximation of 4% is much closer than Gibson's result of -3%.
+
+[^10]: https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-august-2023#key-statistics
+
+I suggest anyone interested in the topic of New Zealand mortality go explore population ages and death rates at Stats NZ themselves, and while there have a look at New Zealand death rates by age, or indeed the public, downloadable, standardised death rates.
 And then ask yourself, given the ready availability of high quality data, is the best default course of action to use the source that automatically compensates for the most confounders?

### 2024-03-19 rawfiles/Draft_letter_preMar19.qmd
suggestions from Monday and Tuesday

--- 2024-03-16 rawfiles/Draft_letter.qmd
@@
 
+If 2020 had the same single year of age death rates as 2019, and new migrant 2020 death rates were the same as the equivalently aged non-migratory population in 2019, the 35,744 reduction of nett migration, and total population growth, would be expected to result in a reduction in 85 deaths due to the youthfully low death rates of migrants.
+
+Conversely, if the 2019 resident population naturally aged by one year combined with subtracting deaths of that age, and death rates of 2020 were the same as 2019, there would be an expected increase of 837 deaths while decreasing total population by about 34,000.
+Gibson's decision not to apply age to the model excludes an increase in expected deaths of an approximate order of magnitude greater than that included.
+But all these complexities are avoided by incorporating aging, such as with current best practice.
+
 Gibson asserted that it was not important to incorporate aging:

### 2024-03-19 rawfiles/imm1920.R

@@ -1,0 +1,120 @@
+# working out that if 2020 had the same death rates by age as 2019
+# and the loss in migration, how much would the death rate go up by.
+
+library(dplyr)
+library(tidyr)
+
+# ITM Migration
+# ITM has lowest open category of 90+
+# so sets the age limits for others.
+autoAge <- paste(0:89,"Years")
+autoAge <- c(autoAge,"90 Years and over")
+ITM <- read.csv("StatsNZData/ITM_agesex20192020.csv", skip=3,
+                col.names = c("Sex","Age","y19","y20")) |> 
+  mutate(Sex = ifelse(Sex == " ", NA_character_,Sex)) |> 
+  fill(Sex) |> 
+  filter(Age %in% autoAge) |> 
+  mutate(change = y20-y19,
+         AgeN = as.numeric(gsub(" Yea.*","",Age)))
+
+# VSD deaths
+autoAge <- paste(0:99,"years")
+autoAge <- c("Less than 1 year", "1 year", autoAge,"100 years and over")
+VSD <- read.csv("StatsNZData/VSD_DeathsDec_age_sex.csv", skip=1,
+                col.names = c("Year","Age","Male","Female")) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
+  fill(Year) |> 
+  filter(Age %in% autoAge, Year == 2019) |> 
+  group_by(Age) |> 
+  slice(1) |> 
+  ungroup() |> 
+  gather(key="Sex", value="Deaths", Male:Female) |> 
+  mutate(Age = ifelse(Age == "Less than 1 year",
+                      "0 years", Age),
+    AgeN = as.numeric(gsub(" yea.*","",Age)),
+         AgeN = ifelse(AgeN > 90, 90, AgeN)) |> 
+  summarise(.by=c(Sex,AgeN),
+            RawDeaths = sum(Deaths)) 
+
+  
+# DPE Population
+autoAge <- paste(0:94,"Years")
+autoAge <- c(autoAge,"95 Years and Over")
+DPE <- read.csv("DPEpopulation_by_age_sex.csv", skip=1,
+                col.names = c("Year","Age","Sex","MeanPop")) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year),
+         Age = ifelse(Age == " ", NA_character_,Age)) |> 
[…]

### 2024-03-21 rawfiles/Draft_letter.qmd
many revisions

@@
 format: gfm
+bibliography: references.bib
 ---
 
-Things David knows he still needs to do (that don't alter overall content):
-
--   more proofreading.
-
--   standard citation format.
-
--   clear links to source data.
-
--   standard figure look
-
--   Captions under graphs.
-
-*The timetable is: I am welcoming changes up to Thursday 21st and regularly updating the draft here until then.
-On Thursday 21st I will make what I expect to be any final changes.
-On Friday 22nd I will be distribution the final draft to people who have expressed an interest in co-signing, with an aim to send it to the journal early in the week beginning Monday 25th.*
+*The timetable is: On Friday 22nd I will be distribution the final draft to people who have expressed an interest in co-signing, with an aim to send it to the journal early in the week beginning Monday 25th.*
 
@@
 
-I am formally writing to point out the significant errors in the paper from Gibson (2024)[^1], published in your journal.
+I am formally writing to point out the significant errors in the [@gibson2024] paper, published in your journal.
 It is important both you and your readership are aware of its shortcomings before basing any decisions on the claims made.
@@
 
-[^1]: John Gibson (22 Feb 2024): Cumulative excess deaths in New Zealand in the COVID-19 era: biases from ignoring changes in population growth rates, New Zealand Economic Papers, DOI: 10.1080/00779954.2024.2314770 available from https://doi.org/10.1080/00779954.2024.2314770
+-   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of [@karlinsky2021][^1].
+-   That Gibson's method, producing positive excess mortality, is more effective than Karlinsky & Kobak in determining excess mortality as it adjusts for immigration based changes in the rate of population growth.
 
--   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of Karlinsky & Kobak[^2].
--   That Gibson's method, producing positive excess mortality, is more effective than Karlinsky & Kobak in determining mortality as it adjusts for population change due to immigration changes.
-
-[^2]: Ariel Karlinsky, Dmitry Kobak (2021) Tracking excess mortality across countries during the COVID-19 pandemic with the World Mortality Dataset eLife 10:e69336.
-Code available from https://github.com/dkobak/excess-mortality
+[^1]: The analysis code for [@karlinsky2021] is available from https://github.com/dkobak/excess-mortality
 
@@
 
-# The underpinning of New Zealand's mortality measurement
+## The underpinning of New Zealand's mortality measurement
 
-For centuries, it has been established practice among actuaries, demographers, and health researchers to measure mortality using death rates by age where such data is available[^3].
-This method superseded Crude Mortality Rates, where CMR takes into account total population change in comparisons by dividing the total number of deaths by the total number of people available to die, improving CMR by acknowledging that different ages have different risks of death, so comparing age-based risk of death on a uniform basis.
-In New Zealand, tutorials on making mortality comparisons by using death rates by age and standardising to a reference population have been published since the 1890s[^4].
+For centuries, it has been established practice among actuaries, demographers, and health researchers to measure mortality using age-adjusted death rates where such data is available [@price1773].
+This method supersedes the use of Crude Mortality Rates[^2] by acknowledging that different ages have different risks of death, so comparing age-based risk of death on a uniform basis.
+In New Zealand, tutorials on making mortality comparisons by using death rates by age and standardising to a reference population have been published since the 1890s [@didsbury1891].
 
-[^3]: Price, Richard (1773). Observations on Reversionary Payments. United Kingdom: (n.p.).
+[^2]: The Crude Mortality Rate (CMR) is a simple calculation of the total number of deaths in a population divided by that population.  The CMR allows comparability of populations of different sizes, but does not make any allowance for differences in composition between populations
 
-[^4]: Report on the Statistics of New Zealand, 1890.
-Published by George Didsbury, Government Printer, Wellington., 1891.
-https://www3.stats.govt.nz/historic_publications/1890-official-handbook/1890-official-handbook.html Age standardised mortality is in the section "Adjusted death-rate a better means of comparison".
-
-Age Standardised Mortality directly adjusts for both population growth and aging, as it is based on the number of deaths of the current resident population divided by the living current resident population, stratified by age group and time period.
+Age Standardised Mortality directly adjusts for both population growth and aging, as it is based on the number of deaths of the current resident population divided by the current resident population, stratified by age group and time period.
 This directly measures the population change effects Gibson is trying to model.
 
-While calculation of New Zealand's cumulative excess mortality for 2020-2022 can vary by a couple of percentage points depending on the exact method, using a death rates by age (standardised) method the Institute of Actuaries finds New Zealand's excess mortality from week 10 2020 to the end of 2022 to be -4% (that there are 4% fewer deaths than would be expected based on pre-COVID-19 trends)[^5].
-
-[^5]: Actuaries Institute: CMI Working Paper 180 https://www.actuaries.org.uk/learn-and-develop/continuous-mortality-investigation/cmi-working-papers/mortality-projections/cmi-working-paper-180
+While calculation of New Zealand's cumulative excess mortality for 2020-2022 can vary by a couple of percentage points depending on the exact method adopted, using an age-standardised death rates method the UK's Institute and Faculty of Actuaries finds New Zealand's excess mortality from week 10 2020 to the end of 2022 to be -4% (that there are 4% fewer deaths than would be expected based on pre-COVID-19 trends) [@instituteandfacultyofactuaries2024].
 
@@
 As Gibson notes, Karlinsky and Kobak were initially able to provide comparison data for 103 countries (later 127).
-By contrast the mortality.org has weekly death rates by age (suitable for age standard mortality comparisons) for 38 countries[^6].
+By contrast the mortality.org has weekly death rates by age (suitable for age standard mortality comparisons) for 38 countries[^3].
 -   It makes weekly comparisons available.
--   Able to be checked by any member of the public for the latest current worldwide information.
+-   It is able to be checked by any member of the public for the latest current worldwide information.
 This last is a function of Our World in Data presentation rather than the method, but it is still a practical benefit.
 
-[^6]: <https://mortality.org/File/GetDocument/Public/STMF/DOC/STMFmetadata.pdf>
+[^3]: <https://mortality.org/File/GetDocument/Public/STMF/DOC/STMFmetadata.pdf>
 
@@
 
-> "*the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ's are!)*"(Ashley Bloomfield) [^7].
-
-[^7]: Ashley Bloomfield, 2024, personal communication
+> "*the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ's are!)*" [@bloomfield2024]
 
@@
 These can be directly compared to death rates in other time periods and locations to measure if the group experience is more or less deadly.
-As it is a rate, it automatically adjusts for change in the population size of the age group.
+As it is a rate, it automatically adjusts for differences in the size of the age group.
 
@@
 
-The total of standard deaths can optionally be divided by the standard population to make age standardised mortality rates.
+The total of standardised deaths can optionally be divided by the standard population to make age standardised mortality rates.
 
-Some variation in calculation results can be caused by the choice of standard population, due to the variation of young and old within a population.
-However, pragmatically, any population that can be argued to be analytically useful should produce similar results.
-Using single year age death rates by sex to 94 then 95+[^8], and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -5% for 2020 to 2022.
+Some variation in calculated results can be caused by the choice of standard population, due to the variation of young and old within a population.
+However, pragmatically, variations arising from the choice of standard population are likely to be immaterial to the results.
+Using single year age death rates by sex to 94 then 95+, and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -5% for 2020 to 2022.
 Using a standard reference population of 2021 when the borders were restricted gives a cumulative excess of -6%.
 Using a pre-COVID-19 standard population of 2019 gives a cumulative excess of -5%.
-Using the standard population of 1961, used by Stats NZ to maintain long-term continuity, cumulative excess is -5%.
+Using the standard population of 1961, used by Stats NZ to maintain long-term continuity, cumulative excess is -6%[^4].
 
-[^8]: NZ annual deaths are available from <https://infoshare.stats.govt.nz> in Population : Deaths : Deaths by age and sex (Annual-Dec).
+[^4]: NZ annual deaths are available from <https://infoshare.stats.govt.nz> in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
 NZ mean annual population is available from Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec) : Mean year ended.
+The 1961 age standardised mortality rate figures are in Populstion: Death Rates - DMM : Standardised death rates (Maori and total population) (Annual-Dec).
+Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
 
-Excess mortality is the difference to what is expected for mortality, so excess calculations using any method are sensitive what is considered normal: the baseline.
-For Age Standardised Mortality the regular process is to use a linear trend of preceding years, and in the case of a major multi-year event like a pandemic, the linear trend of pre-pandemic years.
+Excess mortality is the difference between actual mortality and what is expected.
+The expected value is an estimate based on certain underlying premises, so excess mortality calculations using any method are sensitive what is considered "normal": the baseline.
+For Age Standardised Mortality the regular process is to use a linear trend of preceding years, and where trying to measure the impact of a major multi-year event like a pandemic, the linear trend of pre-pandemic years.
 
@@
 
-Based on the influenza season patterns in recent pre-pandemic years, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline[^9].
-New Zealand has seen a very similar history in this period to Australia, with the added limitation that using 2011, the year of the Christchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is an extremely poor choice to include in baselines as such an event is not a reoccurring annual one.
+After thorough investigation of the data, the Australian Bureau of Statistics (ABS) considers 2013-2019 to give the best baseline for the pandemic period [@australianbureauofstatistics2023].
+New Zealand has seen a similar trend in mortality in this period to Australia, with the added limitation that using 2011, the year of the Christchurch Earthquake and the residents of the second largest city living in broken housing through the following winter, is an extremely poor choice to include in baselines as such an event is not a reoccurring annual one.
 
-[^9]: Australian Bureau of Statistics. https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology
+While many international studies use a 5 year baseline range to increase the number of countries for which weekly level data is available, in practice the results for New Zealand do not differ substantively between a 5 (2015-2019) and 7 (2013-2019) year baseline period.
+Using a 2022 standard population and a 2015-2019 baseline gives -5% excess mortality for 2020-2022, while using a 2013-2019 baseline gives -6%[^5].
 
-While many international covid comparisons use a 5 year baseline range to increase the number of countries for which weekly level data is available, in practice the results for New Zealand do not differ substantively between a 5 (2015-2019) and 7 (2013-2019) year baseline period.
-A 5 year linear baseline and a 7 year baseline both give a cumulative excess for 2020-2022 of -5% when using a 2022 standard population.
+[^5]: Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
 
@@
 
-Starting from a commonly recognised best practice measure of mortality, age standised mortality, which gives a cumulative excess mortality for 2020-2022 in the range of -4% to -6%, we can compare other methods.
+Starting from a commonly recognised best practice measure of mortality, age stanardised mortality, which gives a cumulative excess mortality for 2020-2022 in the range of -4% to -6%, we can compare other methods.
 
@@
 
-Gibson's Population adjusted for changes in growth rates model gives an excess mortality for 2020-2022 range of 1.7% to 5.3% with the central estimate rounding to 4%.
+Gibson's model gives an excess mortality for 2020-2022 range of 1.7% to 5.3% with the central estimate rounding to 4%.
 
-Though not mentioned in the paper, using Population as a model term to a death based model would normally be functionally equivalent to using the commonly known Crude Mortality Rate (deaths among the total population / total living population).
-A 2015-2019 linear regression of the annual CMR gives an excess of 1%.
+Though not mentioned in Gibson's paper, using total population (without aging) as a model term would normally be functionally equivalent to the commonly known Crude Mortality Rate (deaths among the total population / total living population).
+A 2015-2019 linear regression of the annual CMR gives an 2020-2022 cumulative excess of 1%[^6].
+
+[^6]: NZ crude death rates are available from <https://infoshare.stats.govt.nz> in Death Rates - DMM Table: Crude death rate (Maori and total population) (Annual-Dec).
+Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
 
 Gibson's method produces results further from actuarial standard methods than the Karlinsky & Kobak method.
-The reason for Gibson's model performing so poorly compared to Karlinsky & Kobak lies in implicitly misapplying death rates by age.
+The reason for Gibson's model performing so poorly compared to Karlinsky & Kobak lies in the age imbalance of death rates.
 Young people frequently migrate and infrequently die.
@@
 
-![Figure 1: Ages of migration and ages of death have little overlap](md_figures/figure3.png){fig-alt="Young people migrate, old people die" fig-align="center"}
+![Figure 1: Ages of migration and ages of death have little overlap](md_figures/figure1.png){fig-alt="Young people migrate, old people die" fig-align="center"}
 
-Conversely, because the Karlinsky & Kobak method only uses death data, it implicitly limits itself to the population that dies.
+Conversely, because the Karlinsky & Kobak method uses only death data, it implicitly limits itself to the population that dies.
 The population mainly dying are not migrating, so the model does not diverge as far from age standardised mortality as the Gibson method when border changes cause major fluctuations in migration.
 
-If 2020 had the same single year of age death rates as 2019, and new migrant 2020 death rates were the same as the equivalently aged non-migratory population in 2019, then we can calculate the expected deaths from the 35,744 reduction in nett migrationfrom 2019 to 2020.
-Using single age death rates the youthful migrating population would be expected to result in 85 fewer deaths for the 35,744 reduction in total population.
+If 2020 had the same single year of age death rates as 2019, and new migrant 2020 death rates were the same as the equivalently aged non-migratory population in 2019, then we can calculate the expected deaths from the 35,744 reduction in nett migration from 2019 to 2020.
+Using single age death rates the youthful migrating population would be expected to result in 85 fewer deaths for the 35,744 reduction in total population[^7].
+
+[^7]: NZ migration data available from <https://infoshare.stats.govt.nz> in Tourism : International Travel and Migration - ITM : Table: Estimated migration by direction, age group and sex, 12/16-month rule (Annual-Dec).
+NZ death data available in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
+NZ population data available in Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec).
+Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
 
 Conversely, the expected increase in deaths due to aging can be calculated by aging the 2019 resident population by one year, and subtracting annual deaths of that age, and applying the single year death rates of 2019 (the same assumption as deaths expected from migration).
-For the resident population, there would be an expected rise of 837 deaths while the total population decreased by about 34,000.
+For the resident population, there would be an expected rise of 837 deaths while the total population decreased by about 34,000[^8].
 
-Gibson's decision not to apply total population but not age to the model excludes an increase in expected deaths of an approximate order of magnitude greater than the effects of migration.
+[^8]: NZ death data available from <https://infoshare.stats.govt.nz> in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
+NZ population data available in Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec).
+Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
+
+Gibson's decision not to apply total population but not age to the model minimises an expected increase in deaths of an approximate order of magnitude greater than the effects of migration.
 But all these complexities are avoided by incorporating both migration and aging by using age standardised mortality.
@@
 The assumption that aging is a long term process so should show up as a long slow process only applies if there are no sudden demographic shifts among age groups that alter the risk of death.
-The number of 75 and older residents, when compared to the 2012-2016 trend, was increasing in 2019.
+The number of 75 and older residents, when compared to the 2012-2016 trend, was also accelerating in 2019.
 This matches a "sudden" demographic shift of a spike in the age distribution that began decades earlier and naturally aged into ranges that play a more significant role in total deaths.
 
-![Figure 2: The rate of increase in elderly accelerated in 2019](md_figures/figure4.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}
+![Figure 2: The rate of increase in elderly accelerated in 2019](md_figures/figure2.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}
 
@@
 
-Gibson's results are also at variance with other countries statistical authorities, the Australian Bureau of Statistic's age based mortality analysis has excess mortality of 3.4% across 2020-2022[^10].
-Karlinsky & Kobak's approximation of 4% is much closer than Gibson's result of -3%.
-
-[^10]: https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-august-2023#key-statistics
+Karlinsky & Kobak also appears superior for comparisons between countries, as Gibson's results are a worse approximation with ASMR based methods calculated by other countries statistical authorities.
+The Australian Bureau of Statistic's age based mortality analysis has excess mortality of 3.4% across 2020-2022 [@australianbureauofstatistics2023].
+Karlinsky & Kobak's approximation of 4% is much closer to the ABS result than Gibson's result of -3%.
 

### 2024-03-21 rawfiles/Fig2_baseline.R

@@ -1,59 +1,0 @@
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-source("theme.R")
-
-ASMR <- read.csv("StatsNZData/DMM_SDR_TotPop.csv", skip=1, 
-                 col.names = c("Year","ASMR1K")) |> 
-  filter(!is.na(ASMR1K)) |> 
-  mutate(Year = as.numeric(Year)) |> 
-  filter(Year > 1990)
-
-longmod <- lm(ASMR1K ~ Year, data=ASMR |> filter(Year < 2017))
-ASMR$longExp <- longmod$coefficients[1] + 
-  ASMR$Year * longmod$coefficients[2]
-
-shortmod <- lm(ASMR1K ~ Year, data=ASMR |> filter(Year %in% 2010:2016))
-ASMR$shortExp <- shortmod$coefficients[1] + 
-  ASMR$Year * shortmod$coefficients[2]
-ASMR$shortExp[ASMR$Year < 2010] <- NA_real_
-
-xis<- c(2017,
-        2018,
-        2019,
-        2019,
-        2018,
-        2017)
-yis <- c(
-  3.180031,
-  3.073993,
-  2.967956,
-  3.351429,
-  3.406429,
-  3.461429)
-
-structural_over17_19 <- sum(ASMR$shortExp[ASMR$Year %in% 2017:2019] /
-  ASMR$longExp[ASMR$Year %in% 2017:2019])
-
-ggplot(ASMR |> filter(Year < 2020), 
-              aes(x=Year, y=ASMR1K))+
-  annotate("polygon",x=xis,y=yis,fill=six_cols[3])+
-  geom_point(size=0.3) +
-         geom_line(aes(y=longExp), linetype=2)+
-         geom_line(aes(y=shortExp), linetype=2) +
-  geom_line(data=ASMR |> filter(Year < 2017), aes(y=longExp)) +
-  geom_line(data=ASMR |> filter(Year < 2017),aes(y=shortExp)) +
-  theme_davidhood() +
-  theme(plot.title = element_markdown(lineheight = 1.18, size=11,
-                                      margin=margin(t = 5, r = 15, b = 5, l = 0, unit = "pt"),
-                                      hjust=0, vjust=0)) +
-  labs(title=paste0("<span style='color:", six_cols[3],"'>Structural Error</span> from an overlong baseline period in predicting 2017-2019"),
[…]

### 2024-03-21 rawfiles/fig1files/figures.R

@@ -1,258 +1,0 @@
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-library(patchwork)
-library(forcats)
-
-# figure 01a Male death rates by age since 2013
-autoAge <- paste(2:99,"years")
-autoAge <- c("Less than 1 year","1 year",autoAge,"100 years and over")
-VSD <- read.csv("VSD_deaths_age_sex.csv", skip=1,
-                col.names = c("Year","Age","Male","Female")) |> 
-  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
-  fill(Year) |> 
-  filter(Age %in% autoAge) |> 
-  group_by(Year) |> 
-  slice(2:n()) |> 
-  ungroup() |> 
-  mutate(Age = ifelse(Age == "Less than 1 year","0 years",Age),
-         AgeN = as.numeric(gsub(" .*","",Age)),
-         AgeGroup = case_when(AgeN > 79 ~ "80 plus",
-                              AgeN > 59 ~ "60 to 79",
-                              AgeN > 39 ~ "40 to 59",
-                              AgeN > 19 ~ "20 to 39",
-                              TRUE ~ "0 to 19"),
-         Year = as.numeric(Year)) |> 
-  summarise(.by = c(Year,AgeGroup),
-            Males=sum(Male),
-            Females=sum(Female)) |> 
-  filter(Year > 2012)
-
-autoAge <- paste(0:94,"Years")
-autoAge <- c(autoAge,"95 Years and Over")
-DPE <- read.csv("DPEpopulation_by_age_sex.csv", skip=1,
-                col.names = c("Year","Age","Sex","MeanPop")) |> 
-  mutate(Year = ifelse(Year == " ", NA_character_,Year),
-         Age = ifelse(Age == " ", NA_character_,Age)) |> 
-  fill(Year,Age) |> 
-  filter(Sex %in% c("Male","Female"),
-         Age %in% autoAge) |> 
-  mutate(AgeN = as.numeric(gsub(" .*","",Age)),
-         AgeGroup = case_when(AgeN > 79 ~ "80 plus",
-                              AgeN > 59 ~ "60 to 79",
-                              AgeN > 39 ~ "40 to 59",
-                              AgeN > 19 ~ "20 to 39",
-                              TRUE ~ "0 to 19"),
[…]

### 2024-03-21 rawfiles/fig1files/theme.R

@@ -1,28 +1,0 @@
-six_cols <- colorblind_pal()(6)
-footer <- paste0("\nGraph made ", trimws(format(Sys.time(), format="%e %b %Y")),". Graphs: thoughtfulnz.quarto.pub/nzcovidreport/ Contact: thoughtfulnz on mastodon.nz or bsky.app\n")
-
-theme_davidhood <- function(){
-  theme_minimal(base_family="Open Sans",
-                base_size = 9) %+replace%  
-    theme(axis.line.x = element_line(linewidth=0.1),
-          axis.line.y = element_line(linewidth=0.1),
-          axis.ticks = element_line(linewidth=0.2),
-          panel.grid = element_blank(),
-          strip.background = element_rect(fill= "#FFFFFF", colour="#EFEFEF"),
-          strip.text = element_text(margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")),
-          strip.placement = "inside",
-          panel.background = element_rect(fill = "#FFFFFF", colour = "#FFFFFF"),
-          panel.spacing = unit(1.5, "lines"),
-          plot.title = element_text(lineheight = 1.18, size=11,
-                                    margin=margin(t = 5, r = 15, b = 5, l = 15, unit = "pt"),
-                                    hjust=1, vjust=0),
-          plot.subtitle = element_text(lineheight = 1.18, size=7,
-                                    margin=margin(t = 0, r = 15, b = 5, l = 15, unit = "pt"),
-                                    hjust=0),
-          plot.background = element_rect(fill = "#FCFCFC", colour="#FCFCFC"),
-          plot.caption = element_text(margin=margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
-                                      lineheight = 1.15,
-                                      size=7, hjust=1),
-          plot.caption.position = "plot")
-  
-}

### 2024-03-21 rawfiles/figure5.R

@@ -1,48 +1,0 @@
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-library(patchwork)
-source("theme.R")
-six_cols <- colorblind_pal()(6)
-
-VSD <- read.csv("DMM_CDRdec.csv", skip=1) |> 
-  rename(Year=1, Deaths=2) |> 
-  filter(!is.na(Deaths)) |> 
-  mutate(Year = as.numeric(Year)) 
-
-model <- lm(Deaths ~ Year, data=VSD |> filter(Year %in% 2012:2016))
-VSD$expected <- model$coefficients[1] + 
-  VSD$Year * model$coefficients[2]
-
-ggplot(VSD |> filter(Year %in% 2010:2019), aes(x=Year, y=Deaths))+
-  geom_point(size=0.3) +
-         geom_line(aes(y=expected), linetype=2, colour=six_cols[2])+
-  geom_line(data=VSD |> filter(Year %in% 2012:2016), aes(y=expected), colour=six_cols[2]) +
-  theme_minimal(base_family="Open Sans",
-                base_size = 6) +
-  theme(
-    plot.background = element_rect(fill = "#FAFAFA", colour="#FAFAFA"),
-    panel.background = element_rect(fill = "#FFFFFF", colour = "#FFFFFF"),
-    panel.spacing.x = unit(9, "points"),
-    panel.spacing.y = unit(3, "points"),
-    panel.grid = element_blank(),
-    strip.background = element_rect(fill= "#FFFFFF", colour="#EFEFEF"),
-    strip.text = element_text(margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")),
-    strip.placement = "inside",
-    axis.title.y.left = element_text(margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"))
-  ) +
-  labs(title="The crude death in 2017-2019 are above the 5 year linear trend for 2012-2016",
-       subtitle="This is because, before covid, the accelerating aging population is accelerating deaths",
-       y="Crude Death Rate", x="")
-
-ggsave(filename="figure5.png",
-       height=4, width = 6, dpi=300, units = "in", bg = "white")
-
-VSD$perc = 100*VSD$Deaths/VSD$expected-100
-
-
-## https://www.abs.gov.au/articles/measuring-australias-excess-mortality-during-covid-19-pandemic-until-first-quarter-2023#methodology
-
-

### 2024-03-21 rawfiles/intextCalcs.R

@@ -1,105 +1,0 @@
-library(dplyr)
-library(tidyr)
-
-# figure 01a Male death rates by age 
-autoAge <- paste(2:99,"years")
-autoAge <- c("Less than 1 year","1 year",autoAge,"100 years and over")
-VSD <- read.csv("VSD_deaths_age_sex.csv", skip=1,
-                col.names = c("Year","Age","Male","Female")) |> 
-  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
-  fill(Year) |> 
-  filter(Age %in% autoAge) |> 
-  group_by(Year) |> 
-  slice(2:n()) |> 
-  ungroup() |> 
-  mutate(Age = ifelse(Age == "Less than 1 year","0 years",Age),
-         AgeN = as.numeric(gsub(" .*","",Age)),
-         AgeGroup = case_when(AgeN > 94 ~ "95 plus",
-                              TRUE ~ as.character(AgeN)),
-         Year = as.numeric(Year)) |> 
-  summarise(.by = c(Year,AgeGroup),
-            Males=sum(Male),
-            Females=sum(Female)) |> 
-  filter(Year > 2005)
-
-autoAge <- paste(0:94,"Years")
-autoAge <- c(autoAge,"95 Years and Over")
-DPE <- read.csv("DPEpopulation_by_age_sex.csv", skip=1,
-                col.names = c("Year","Age","Sex","MeanPop")) |> 
-  mutate(Year = ifelse(Year == " ", NA_character_,Year),
-         Age = ifelse(Age == " ", NA_character_,Age)) |> 
-  fill(Year,Age) |> 
-  filter(Sex %in% c("Male","Female"),
-         Age %in% autoAge) |> 
-  mutate(AgeN = as.numeric(gsub(" .*","",Age)),
-         AgeGroup = case_when(AgeN > 94 ~ "95 plus",
-                              TRUE ~ as.character(AgeN)),
-         AnnualPop = as.numeric(MeanPop),
-         Year = as.numeric(Year)) |> 
-  summarise(.by = c(Year,AgeGroup,Sex),
-            Population=sum(AnnualPop, na.rm=TRUE)) |> 
-  filter(Year > 2005)
-
-Males <- DPE |> 
-  filter(Sex == "Male") |> 
-  inner_join(VSD,by = join_by(Year, AgeGroup)) |> 
[…]

### 2024-03-21 rawfiles/rejected_figure4.R

@@ -1,57 +1,0 @@
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-source("theme.R")
-six_cols <- colorblind_pal()(6)
-
-CMR <- read.csv("DMM_CDRdec.csv", skip=1, 
-                 col.names = c("Year","CDR1K")) |> 
-  filter(!is.na(CDR1K)) |> 
-  mutate(Year = as.numeric(Year)) 
-
-mod7y <- lm(CDR1K ~ Year, data=CMR |> filter(Year %in% 2013:2019))
-CMR$mod7y <- mod7y$coefficients[1] + 
-  CMR$Year * mod7y$coefficients[2]
-
-mod5y <- lm(CDR1K ~ Year, data=CMR |> filter(Year %in% 2015:2019))
-CMR$mod5y <- mod5y$coefficients[1] + 
-  CMR$Year * mod5y$coefficients[2]
-CMR$percent5year <- 100*CMR$CDR1K/CMR$mod5y-100
-CMR$percent7year <- 100*CMR$CDR1K/CMR$mod7y-100
-
-excessdf <- data.frame(Baseline = c("2015-2019", "2013-2019", "2015-2019", "2013-2019",
-                          "2015-2019", "2013-2019", "2015-2019", "2013-2019"),
-           Year = c("2020","2020","2021","2021","2022","2022","nett","nett"),
-           value = c(round(CMR$percent5year[CMR$Year == 2020],2),
-                     round(CMR$percent7year[CMR$Year == 2020],2),
-                     round(CMR$percent5year[CMR$Year == 2021],2),
-                     round(CMR$percent7year[CMR$Year == 2021],2),
-                     round(CMR$percent5year[CMR$Year == 2022],2),
-                     round(CMR$percent7year[CMR$Year == 2022],2),
-                     round(sum(CMR$percent5year[CMR$Year %in% 2020:2022]),2),
-                     round(sum(CMR$percent7year[CMR$Year %in% 2020:2022]),2)))
-innergraph <- ggplot(excessdf, aes(x=Year,y=Baseline, label=value))+
-  geom_tile(fill="white", colour="black") +geom_text() + theme_tufte()  +
-  scale_x_discrete(position = "top")+
-  labs(title="Excess as percent of expected", y="Baseline\n ",x="")
-  
-ggplot(CMR |> filter(Year %in% 2012:2023), 
[…]

### 2024-03-21 rawfiles/table1_CDRvsASMR.R

@@ -1,57 +1,0 @@
-# table comparing CDR and ASMR predictions
-
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-library(patchwork)
-library(forcats)
-
-autoAge <- paste(2:99,"years")
-autoAge <- c("Less than 1 year","1 year",autoAge,"100 years and over")
-VSD <- read.csv("StatsNZData/VSD_DeathsDec_age_sex.csv", skip=1,
-                col.names = c("Year","Age","Male","Female")) |> 
-  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
-  fill(Year) |> 
-  filter(Age %in% autoAge) |> 
-  group_by(Year, Age) |> 
-  slice(1) |> 
-  ungroup() |> 
-  mutate(Age = ifelse(Age == "Less than 1 year","0 years",Age),
-         AgeN = as.numeric(gsub(" .*","",Age)),
-         Year = as.numeric(Year)) |> 
-  gather(key="Sex",value="Deaths", Male:Female) |> 
-  mutate(AgeN = ifelse(AgeN > 94,95, AgeN)) |> 
-  summarise(.by=c(Year,Sex,AgeN),
-            Deaths = sum(Deaths))
-
-
-autoAge <- paste(0:94,"Years")
-autoAge <- c(autoAge,"95 Years and Over")
-DPE <- read.csv("StatsNZData/DPE_PopDecMean_age_sex.csv", skip=2,
-                col.names = c("Year","Age","Male","Female")) |> 
-  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
-  fill(Year) |> 
-  filter(Age %in% autoAge) |> 
-  mutate(AgeN = as.numeric(gsub(" .*","",Age)),
-         Year = as.numeric(Year)) |> 
-  gather(key="Sex",value="MeanPopulation", Male:Female) |> 
-  select(Year, AgeN, Sex, MeanPopulation)
-
-Stnd <- DPE |> filter(Year == 2018) |> 
-  select(AgeN, Sex, StndPopulation = MeanPopulation)
-
-totalStndPop = sum(Stnd$StndPopulation)
-
-ASMR <- VSD |> 
-  inner_join(DPE, by = join_by(Year, Sex, AgeN)) |> 
-  inner_join(Stnd, by = join_by(Sex, AgeN)) |> 
[…]

### 2024-03-21 supportFiles/Fig1_immigration.R

@@ -64,8 +64,10 @@ ggplot(aes(x=AgeN, y=percent, fill=series))+
                      labels=c("under 1","30","60","90+")) +
   labs(title="2019 Deaths vs. Migrants\nas percentage of resident population",
        y="Percentage", x="Age",
-       caption="Source: infoshare.stats.govt.nz")
-ggsave(filename="md_figures/figure3.png",
+       caption="Source: infoshare.stats.govt.nz: Deaths by age and sex (Annual-Dec)
+Estimated Resident Population by Age and Sex (1991+) (Annual-Dec)
+Estimated migration by direction, age group and sex, 12/16-month rule (Annual-Dec)")
+ggsave(filename="../rawfiles/md_figures/figure1.png",
        height=4, width = 6, dpi=300, units = "in", bg = "white")
-ggsave(filename="../md_figures/figure3.png",
+ggsave(filename="../md_figures/figure1.png",
        height=4, width = 6, dpi=300, units = "in", bg = "white")

### 2024-03-21 supportFiles/Fig2_75plus.R

@@ -1,4 +1,4 @@
-# table comparing CDR and ASMR predictions
+# 75+ trend
 
 library(dplyr)
 library(tidyr)
@@ -34,9 +34,9 @@ ggplot(DPE, aes(x=Year, y=Living75plus))+
   scale_x_continuous(breaks=c(2012,2016,2020))+
   labs(title="Actual 75 years and older vs 2012-2016 trend",
        y="Residents 75 years and Older", x="Year",
-       caption="Source: infoshare.stats.govt.nz")
-ggsave(filename="md_figures/figure4.png",
+       caption="Source: infoshare.stats.govt.nz
+Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec)")
+ggsave(filename="../rawfiles/md_figures/figure2.png",
        height=4, width = 6, dpi=300, units = "in", bg = "white")
-ggsave(filename="../md_figures/figure4.png",
+ggsave(filename="../md_figures/figure2.png",
        height=4, width = 6, dpi=300, units = "in", bg = "white")
-

### 2024-03-21 supportFiles/footnotes4and5.R

@@ -1,0 +1,92 @@
+library(dplyr)
+library(tidyr)
+
+# Deaths, male/female single years then aggregate 95+
+DeathAges <- paste(2:99,"years")
+DeathAges <- c("Less than 1 year","1 year",DeathAges,"100 years and over")
+VSD <- read.csv("StatsNZData/VSD_DeathsDec_age_sex.csv", skip=1,
+                col.names = c("Year","Age","Male","Female")) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
+  fill(Year) |> 
+  filter(Age %in% DeathAges) |> 
+  group_by(Year,Age) |> #100 years + repeated in file so slicing 2nd
+  slice(1) |> 
+  ungroup() |> 
+  mutate(Age = ifelse(Age == "Less than 1 year","0 years",Age),
+         AgeN = as.numeric(gsub(" .*","",Age)),
+         AgeGroup = case_when(AgeN > 94 ~ 95,
+                              TRUE ~ AgeN),
+         Year = as.numeric(Year)) |> 
+  summarise(.by = c(Year,AgeGroup),
+            Male=sum(Male),
+            Female=sum(Female)) |> 
+  filter(Year > 2005) |> 
+  gather(key="Sex", value="Deaths", Male, Female) 
+
+# Population Male/Female single years then 95+
+PopAges <- paste(0:94,"Years")
+PopAges <- c(PopAges,"95 Years and Over")
+DPE <- read.csv("StatsNZData/DPE_PopDecMean_age_sex.csv", skip=2,
+                col.names = c("Year","Age","Male","Female")) |> 
+  mutate(Year = ifelse(Year == " ", NA_character_,Year),
+         Age = ifelse(Age == " ", NA_character_,Age)) |> 
+  fill(Year,Age) |> 
+  filter(Age %in% PopAges) |> 
+  mutate(AgeN = as.numeric(gsub(" .*","",Age)),
+         AgeGroup = case_when(AgeN > 94 ~ 95,
+                              TRUE ~ AgeN),
+         Year = as.numeric(Year)) |> 
+  summarise(.by = c(Year,AgeGroup),
+            Male=sum(Male),
+            Female=sum(Female)) |> 
+  filter(Year > 2005) |> 
+  gather(key="Sex", value="Population", Male, Female) 
+
+# function for a liner regression off a chosen baseline with
[…]

### 2024-03-21 supportFiles/footnotes6.R

@@ -1,0 +1,18 @@
+library(dplyr)
+
+DMM <- read.csv("StatsNZData/DMM_CDRDec_TotPop.csv",
+                skip=1) |> 
+  filter(!is.na(Total.Population)) |> 
+  rename(Year=1,CDR=2) |> 
+  mutate(Year = as.numeric(Year))
+DMModel <- lm(CDR ~ Year, 
+              DMM |> filter(Year %in% 2015:2019))
+DMM$Expected <- DMModel$coefficients[1] +
+  DMModel$coefficients[2] * DMM$Year
+DMM$excess <- 100 * DMM$CDR / DMM$Expected - 100
+cumexc <- sum(DMM$excess[DMM$Year %in% 2020:2022])
+print(round(cumexc,0))
+
+
+
+

### 2024-03-21 rawfiles/letter_to_ed.qmd
tidying up supporting files

--- 2024-03-21 rawfiles/Draft_letter.qmd
@@
 ---
-title: Draft Letter
+title: Letter to Editor
 author: David Hood
@@
 -   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of [@karlinsky2021][^1].
--   That Gibson's method, producing positive excess mortality, is more effective than Karlinsky & Kobak in determining excess mortality as it adjusts for immigration based changes in the rate of population growth.
+-   That Gibson's method, producing positive excess mortality, is more effective than Karlinsky & Kobak in determining excess mortality as it adjusts for immigration-based changes in the rate of population growth.
 
@@
 
-[^2]: The Crude Mortality Rate (CMR) is a simple calculation of the total number of deaths in a population divided by that population.  The CMR allows comparability of populations of different sizes, but does not make any allowance for differences in composition between populations
+[^2]: The Crude Mortality Rate (CMR) is a simple calculation of the total number of deaths in a population divided by that population.
+The CMR allows comparability of populations of different sizes, but does not make any allowance for differences in composition between populations
 
@@
 
-[^3]: <https://mortality.org/File/GetDocument/Public/STMF/DOC/STMFmetadata.pdf>
+[^3]: The count of 38 countries was obtained from counting the countries in the STMF metadata file https://mortality.org/File/GetDocument/Public/STMF/DOC/STMFmetadata.pdf
 
@@
 
-> "*the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ's are!)*" [@bloomfield2024]
+> "...*the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ's are!)*" [@bloomfield2024]
 
@@
 
-[^4]: NZ annual deaths are available from <https://infoshare.stats.govt.nz> in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
+[^4]: NZ annual deaths are available from https://infoshare.stats.govt.nz in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
 NZ mean annual population is available from Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec) : Mean year ended.
-The 1961 age standardised mortality rate figures are in Populstion: Death Rates - DMM : Standardised death rates (Maori and total population) (Annual-Dec).
+The 1961 age standardised mortality rate figures are in Population: Death Rates - DMM : Standardised death rates (Maori and total population) (Annual-Dec).
 Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
@@
 
-Starting from a commonly recognised best practice measure of mortality, age stanardised mortality, which gives a cumulative excess mortality for 2020-2022 in the range of -4% to -6%, we can compare other methods.
+To measure the accuracy of approximations of mortality, the difference of an approximation to age-standardised mortality methods can be measured.
+Earlier age-stratified methods give a cumulative excess mortality for 2020-2022 in the -4% to -6% range.
 
@@
 
-[^6]: NZ crude death rates are available from <https://infoshare.stats.govt.nz> in Death Rates - DMM Table: Crude death rate (Maori and total population) (Annual-Dec).
+[^6]: NZ crude death rates are available from https://infoshare.stats.govt.nz in Population : Death Rates - DMM Crude death rate (Maori and total population) (Annual-Dec).
 Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
@@
 
-![Figure 1: Ages of migration and ages of death have little overlap](md_figures/figure1.png){fig-alt="Young people migrate, old people die" fig-align="center"}
+![Figure 1: Ages of migration and ages of death have little overlap](md_figures/figure1.png){fig-alt="Young people mainly migrate, old people mainly die" fig-align="center"}
 
@@
 If 2020 had the same single year of age death rates as 2019, and new migrant 2020 death rates were the same as the equivalently aged non-migratory population in 2019, then we can calculate the expected deaths from the 35,744 reduction in nett migration from 2019 to 2020.
-Using single age death rates the youthful migrating population would be expected to result in 85 fewer deaths for the 35,744 reduction in total population[^7].
+Using single age death rates, the youthful migrating population would be expected to result in 85 fewer deaths for the 35,744 reduction in total population[^7].
 
-[^7]: NZ migration data available from <https://infoshare.stats.govt.nz> in Tourism : International Travel and Migration - ITM : Table: Estimated migration by direction, age group and sex, 12/16-month rule (Annual-Dec).
+[^7]: NZ migration data available from https://infoshare.stats.govt.nz in Tourism : International Travel and Migration - ITM : Table: Estimated migration by direction, age group and sex, 12/16-month rule (Annual-Dec).
 NZ death data available in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
@@
 
-[^8]: NZ death data available from <https://infoshare.stats.govt.nz> in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
+[^8]: NZ death data available from https://infoshare.stats.govt.nz in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
 NZ population data available in Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec).
@@
 
-The assumption that aging is a long term process so should show up as a long slow process only applies if there are no sudden demographic shifts among age groups that alter the risk of death.
+The assumption that aging is a long-term process so should show up as a long slow process only applies if there are no sudden demographic shifts among age groups that alter the risk of death.
 The number of 75 and older residents, when compared to the 2012-2016 trend, was also accelerating in 2019.
@@
 
-![Figure 2: The rate of increase in elderly accelerated in 2019](md_figures/figure2.png){fig-alt="Aging increases deaths and the crude death rate" fig-align="center"}
+![Figure 2: The rate of increase in elderly accelerated in 2019](md_figures/figure2.png){fig-alt="The number of living elderly has been growing faster than trend prior to 2020" fig-align="center"}
 

### 2024-03-25 rawfiles/letter_to_ed.qmd
final (hopefully) version

@@
 ---
-title: Letter to Editor
-author: David Hood
 date: last-modified
@@
 ---
-
-*The timetable is: On Friday 22nd I will be distribution the final draft to people who have expressed an interest in co-signing, with an aim to send it to the journal early in the week beginning Monday 25th.*
 

### 2024-03-30 rawfiles/letter_to_ed.qmd
docx now pdf
more rewrite

@@
 date: last-modified
-format: gfm
+execute:
+  keep-md: true
+format:
+  pdf:
+    fig-pos: 'h'
 bibliography: references.bib
@@
 However, pragmatically, variations arising from the choice of standard population are likely to be immaterial to the results.
-Using single year age death rates by sex to 94 then 95+, and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -5% for 2020 to 2022.
-Using a standard reference population of 2021 when the borders were restricted gives a cumulative excess of -6%.
-Using a pre-COVID-19 standard population of 2019 gives a cumulative excess of -5%.
-Using the standard population of 1961, used by Stats NZ to maintain long-term continuity, cumulative excess is -6%[^4].
+Using single year age death rates by sex to 94 then 95+, and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -2% for 2020 to 2022.
+Using a standard reference population of 2021 when the borders were restricted gives a cumulative excess of -2%.
+Using a pre-COVID-19 standard population of 2019 gives a cumulative excess of -2%.
+Using the standard population of 1961, used by Stats NZ to maintain long-term continuity, cumulative excess is -2%[^4].
 
@@
 While many international studies use a 5 year baseline range to increase the number of countries for which weekly level data is available, in practice the results for New Zealand do not differ substantively between a 5 (2015-2019) and 7 (2013-2019) year baseline period.
-Using a 2022 standard population and a 2015-2019 baseline gives -5% excess mortality for 2020-2022, while using a 2013-2019 baseline gives -6%[^5].
+Using a 2022 standard population and a 2015-2019 baseline gives -2% excess mortality for 2020-2022, while using a 2013-2019 baseline gives -2%[^5].
 
@@
 
-To measure the accuracy of approximations of mortality, the difference of an approximation to age-standardised mortality methods can be measured.
-Earlier age-stratified methods give a cumulative excess mortality for 2020-2022 in the -4% to -6% range.
+To assess the accuracy of approximations of mortality, the difference of an approximation to age-standardised mortality methods can be measured.
+Earlier age-stratified methods give a cumulative excess mortality for 2020-2022 in the -2% to -4% range.
 
@@
 Though not mentioned in Gibson's paper, using total population (without aging) as a model term would normally be functionally equivalent to the commonly known Crude Mortality Rate (deaths among the total population / total living population).
-A 2015-2019 linear regression of the annual CMR gives an 2020-2022 cumulative excess of 1%[^6].
+A 2015-2019 linear regression of the annual CMR gives an 2020-2022 cumulative excess of 0%[^6].
 
@@
 Conversely, the expected increase in deaths due to aging can be calculated by aging the 2019 resident population by one year, and subtracting annual deaths of that age, and applying the single year death rates of 2019 (the same assumption as deaths expected from migration).
-For the resident population, there would be an expected rise of 837 deaths while the total population decreased by about 34,000[^8].
+For the resident population, ecluding under one yer olds, there would be an expected rise of 837 deaths while the total population decreased by about 33,963[^8].
 
@@
 The number of 75 and older residents, when compared to the 2012-2016 trend, was also accelerating in 2019.
-This matches a "sudden" demographic shift of a spike in the age distribution that began decades earlier and naturally aged into ranges that play a more significant role in total deaths.
+This matches a "sudden" demographic shift of an acceleration in the age distribution that began decades earlier and naturally aged into ranges that play a more significant role in total deaths.
 
 ![Figure 2: The rate of increase in elderly accelerated in 2019](md_figures/figure2.png){fig-alt="The number of living elderly has been growing faster than trend prior to 2020" fig-align="center"}
+But the dramatic increase in elderly only contributes to mortality as one of other interacting factors, such as the age specific risk environment.
+A more direct measure of importance is the difference in outcome between the Crude Mortality Rate (ignoring age structure) and the Age Standardised Mortality Rate (based on age startified death rates).
+The difference in excess mortality between those two methods is exactly and only the difference caused by using age specific death rates.
 
 Of more surprise is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the normal method of comparing death rates while ignoring the effects of age.
-While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same answers, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of Karlinsky & Kobak.
+While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same outcome, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of Karlinsky & Kobak.
 But, compared to the differences with methods acknowledging age, such problems are minor.
@@
 
-I suggest anyone interested in the topic of New Zealand mortality go explore population ages and death rates at Stats NZ themselves, and while there have a look at New Zealand death rates by age, or indeed the public, downloadable, standardised death rates.
-And then ask yourself, given the ready availability of high quality data, is the best default course of action to use the source that automatically compensates for the most confounders?
+Specific methods are used to facilitate particular goals, such as breadth of comparison.
+Methods can righty be critiqued when applied to other goals, such as country specific accuracy.
+But in making such critiques, it is important to provide the existing best methods for that goal.
+And if making specific claims of improved accuracy, benchmarking each method against the most accurate establishes the relative worth of different methods.
+
+David Hood,
+
+University of Otago
+
+## References

### 2024-03-30 supportFiles/footnotes4and5.R

@@ -1,77 +1,93 @@
 library(dplyr)
 library(tidyr)
 
-# Deaths, male/female single years then aggregate 95+
-DeathAges <- paste(2:99,"years")
-DeathAges <- c("Less than 1 year","1 year",DeathAges,"100 years and over")
-VSD <- read.csv("StatsNZData/VSD_DeathsDec_age_sex.csv", skip=1,
-                col.names = c("Year","Age","Male","Female")) |> 
-  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
-  fill(Year) |> 
-  filter(Age %in% DeathAges) |> 
-  group_by(Year,Age) |> #100 years + repeated in file so slicing 2nd
-  slice(1) |> 
-  ungroup() |> 
-  mutate(Age = ifelse(Age == "Less than 1 year","0 years",Age),
-         AgeN = as.numeric(gsub(" .*","",Age)),
-         AgeGroup = case_when(AgeN > 94 ~ 95,
-                              TRUE ~ AgeN),
+# Deaths, rearranged into long-style database form
+# Single years then 100+ by sex
+VSD_male <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
+  select(c(1,27:127)) |> 
+  rename(Year = X., X0.years = Less.than.1.year, X1.years=X1.year) |> 
+  filter(!is.na(X0.years)) |> 
+  gather(key="Agetext", value="Deaths", -Year) |> 
+  mutate(Ageshort = gsub("^X","",Agetext),
+         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
+         Sex="Male",
          Year = as.numeric(Year)) |> 
-  summarise(.by = c(Year,AgeGroup),
-            Male=sum(Male),
-            Female=sum(Female)) |> 
-  filter(Year > 2005) |> 
-  gather(key="Sex", value="Deaths", Male, Female) 
+  select(Year, Sex, AgeN, Deaths)
+VSD_female <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
+  select(c(1,154:254)) |> 
+  rename(Year = X., X0.years = Less.than.1.year.1, X1.years=X1.year.1) |> 
+  filter(!is.na(X0.years)) |> 
+  gather(key="Agetext", value="Deaths", -Year) |> 
+  mutate(Ageshort = gsub("^X","",Agetext),
[…]

### 2024-03-30 supportFiles/footnotes6.R

@@ -1,6 +1,6 @@
 library(dplyr)
 
-DMM <- read.csv("StatsNZData/DMM_CDRDec_TotPop.csv",
+DMM <- read.csv("RawData/DMM_CDRDec_TotPop.csv",
                 skip=1) |> 
   filter(!is.na(Total.Population)) |> 
   rename(Year=1,CDR=2) |> 
@@ -9,9 +9,10 @@ DMModel <- lm(CDR ~ Year,
               DMM |> filter(Year %in% 2015:2019))
 DMM$Expected <- DMModel$coefficients[1] +
   DMModel$coefficients[2] * DMM$Year
-DMM$excess <- 100 * DMM$CDR / DMM$Expected - 100
-cumexc <- sum(DMM$excess[DMM$Year %in% 2020:2022])
-print(round(cumexc,0))
+cumulatime <- DMM |> filter(Year >= 2020, Year <=2022)
+percentage = 100 * sum(cumulatime$CDR) /
+  sum(cumulatime$Expected) - 100
+print(round(percentage,0))
 
 
 

### 2024-03-30 supportFiles/footnotes7and8.R

@@ -7,114 +7,143 @@ library(tidyr)
 # ITM Migration
 # ITM has lowest open category of 90+
 # so sets the age limits for others.
-autoAge <- paste(0:89,"Years")
-autoAge <- c(autoAge,"90 Years and over")
-ITM <- read.csv("StatsNZData/ITM_agesex20192020.csv", skip=3,
-                col.names = c("Sex","Age","y19","y20")) |> 
-  mutate(Sex = ifelse(Sex == " ", NA_character_,Sex)) |> 
-  fill(Sex) |> 
-  filter(Age %in% autoAge) |> 
-  mutate(change = y20-y19,
-         AgeN = as.numeric(gsub(" Yea.*","",Age)))
+colist <- c(1,(c(1:90,127) * 2))
+ITMf <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
+  select(all_of(colist)) 
+names(ITMf) <- c("Year", paste0("X",0:90))
+ITMflong <- ITMf |> 
+  filter(!is.na(X20)) |> 
+  gather(key="AgeText", value="Net", -Year) |> 
+  mutate(Sex="Female")
+
+colist <- c(1,((c(1:90,127) * 2) +1))
+ITMm <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
+  select(all_of(colist)) 
+names(ITMm) <- c("Year", paste0("X",0:90))
+ITMmlong <- ITMm |> 
+  filter(!is.na(X20)) |> 
+  gather(key="AgeText", value="Net", -Year) |> 
+  mutate(Sex="Male")
+
+ITM <- bind_rows(ITMflong,ITMmlong) |> 
+  mutate(AgeN = as.numeric(gsub("X","",AgeText)),
+         Year = as.numeric(Year)) |> 
+  arrange(Sex, AgeN, Year) |> 
+  group_by(Sex,AgeN) |> 
+  mutate(followingYear = lead(Net),
+         change2following = followingYear - Net ) |> 
+  ungroup() |> 
+  select(Year,Sex,AgeN,Net,change2following)
 
 # VSD deaths
-autoAge <- paste(0:99,"years")
-autoAge <- c("Less than 1 year", "1 year", autoAge,"100 years and over")
-VSD <- read.csv("StatsNZData/VSD_DeathsDec_age_sex.csv", skip=1,
-                col.names = c("Year","Age","Male","Female")) |> 
[…]

### 2024-03-30 rawfiles/letter_to_ed.qmd
APA 7 references as per journal

@@
 bibliography: references.bib
+csl: apa.csl
 ---
@@
 
-![Figure 2: The rate of increase in elderly accelerated in 2019](md_figures/figure2.png){fig-alt="The number of living elderly has been growing faster than trend prior to 2020" fig-align="center"}
-But the dramatic increase in elderly only contributes to mortality as one of other interacting factors, such as the age specific risk environment.
-A more direct measure of importance is the difference in outcome between the Crude Mortality Rate (ignoring age structure) and the Age Standardised Mortality Rate (based on age startified death rates).
-The difference in excess mortality between those two methods is exactly and only the difference caused by using age specific death rates.
+![Figure 2: The rate of increase in elderly accelerated in 2019](md_figures/figure2.png){fig-alt="The number of living elderly has been growing faster than trend prior to 2020" fig-align="center"} But the dramatic increase in elderly only contributes to mortality as one of several interacting factors, such as the age specific risk environment.
+A more direct measure of the importance of age stratification is the difference in outcome between the Crude Mortality Rate (ignoring age structure) and the Age Standardised Mortality Rate (based on age stratified death rates).
+The difference in excess mortality between those two methods is exactly and only the difference caused by not using age specific death rates.
 
-Of more surprise is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the normal method of comparing death rates while ignoring the effects of age.
+Of more surprise than the comparison to age standartdised methods is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the normal method of comparing death rates while ignoring the effects of age.
 While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same outcome, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of Karlinsky & Kobak.
@@
 
-Specific methods are used to facilitate particular goals, such as breadth of comparison.
-Methods can righty be critiqued when applied to other goals, such as country specific accuracy.
-But in making such critiques, it is important to provide the existing best methods for that goal.
-And if making specific claims of improved accuracy, benchmarking each method against the most accurate establishes the relative worth of different methods.
+Specific methods are employed to achieve particular objectives, such as broadening the scope of comparison.
+It is reasonable to critique the use of such methods if they are applied to different objectives, such as ensuring accuracy on a country-specific level.
+However, in making such critiques, it is essential to identify the current best practices for achieving that specific objective.
+Additionally, when making claims about enhanced accuracy, it is crucial to benchmark each method against the most accurate standard to establish the comparative value of different approaches.
 

### 2024-03-31 rawfiles/letter_to_ed_docx.qmd
final... ish

--- 2024-03-30 rawfiles/letter_to_ed.qmd
@@
 date: last-modified
-execute:
-  keep-md: true
-format:
-  pdf:
-    fig-pos: 'h'
+title: "Comment on Cumulative Excess Deaths in New Zealand in the COVID-19 Era: Biases from Ignoring Changes in Population Growth Rates"
+format: docx
 bibliography: references.bib
@@
 
-To the editor and editorial board,
+This is a response to the [@gibson2024] paper in which, to give my summary of the paper, Gibson asserts that New Zealand authorities' measurement of mortality in the pandemic period is based on a flawed metric.
+Gibson's own metric, based on regression of deaths with an additive component of total population, is asserted as a more accurate model and produces approximately 4% excess mortality for the 2020-2022 period.
+If New Zealand authorities are using flawed metrics with significantly different results to reality, this would have profound governance and policy implications.
 
-I am formally writing to point out the significant errors in the [@gibson2024] paper, published in your journal.
-It is important both you and your readership are aware of its shortcomings before basing any decisions on the claims made.
-In particular, the paper asserts two claims:
+In particular, the paper asserts two broad claims:
 
--   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of [@karlinsky2021][^1].
+-   That claims of negative New Zealand mortality are underpinned by the "Our World In Data" projection graph using the method of [@karlinsky2021] [^1].
 -   That Gibson's method, producing positive excess mortality, is more effective than Karlinsky & Kobak in determining excess mortality as it adjusts for immigration-based changes in the rate of population growth.
@@
 
-> "...*the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ's are!)*" [@bloomfield2024]
+> '...*the key point being any comparisons both within country (over time) or between countries should be age standardised to provide the most accurate picture possible (and of course be based on high quality and complete underlying deaths data which we can say NZ's are!)*' [@bloomfield2024]
 
@@
 Death rates for an age stratified group, for a time period in a given location, measure the rate at which that age group dies.
-These can be directly compared to death rates in other time periods and locations to measure if the group experience is more or less deadly.
+These can be directly compared to death rates in other time periods and locations to measure if the group experience is more, or less, deadly.
 As it is a rate, it automatically adjusts for differences in the size of the age group.
@@
 However, pragmatically, variations arising from the choice of standard population are likely to be immaterial to the results.
-Using single year age death rates by sex to 94 then 95+, and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess of -2% for 2020 to 2022.
+Using single year age death rates by sex to 94 then 95+, and a linear baseline from 2013-2019, a standard reference population of 2023 gives a cumulative excess[^4] of -2% for 2020 to 2022.
 Using a standard reference population of 2021 when the borders were restricted gives a cumulative excess of -2%.
 Using a pre-COVID-19 standard population of 2019 gives a cumulative excess of -2%.
-Using the standard population of 1961, used by Stats NZ to maintain long-term continuity, cumulative excess is -2%[^4].
+Using the standard population of 1961, used by Stats NZ to maintain long-term continuity, cumulative excess is -2%[^5].
 
-[^4]: NZ annual deaths are available from https://infoshare.stats.govt.nz in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
+[^4]: Using the standard cumulative excess measure of 100\*((sum(actual)/sum(expected)) -1).
+
+[^5]: NZ annual deaths are available from https://infoshare.stats.govt.nz in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
 NZ mean annual population is available from Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec) : Mean year ended.
@@
 While many international studies use a 5 year baseline range to increase the number of countries for which weekly level data is available, in practice the results for New Zealand do not differ substantively between a 5 (2015-2019) and 7 (2013-2019) year baseline period.
-Using a 2022 standard population and a 2015-2019 baseline gives -2% excess mortality for 2020-2022, while using a 2013-2019 baseline gives -2%[^5].
+Using a 2022 standard population and a 2015-2019 baseline gives -2% excess mortality for 2020-2022, while using a 2013-2019 baseline gives -2%[^6].
 
-[^5]: Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
+[^6]: Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
 
@@
 Though not mentioned in Gibson's paper, using total population (without aging) as a model term would normally be functionally equivalent to the commonly known Crude Mortality Rate (deaths among the total population / total living population).
-A 2015-2019 linear regression of the annual CMR gives an 2020-2022 cumulative excess of 0%[^6].
+A 2015-2019 linear regression of the annual CMR gives an 2020-2022 cumulative excess of 0%[^7].
 
-[^6]: NZ crude death rates are available from https://infoshare.stats.govt.nz in Population : Death Rates - DMM Crude death rate (Maori and total population) (Annual-Dec).
+[^7]: NZ crude death rates are available from https://infoshare.stats.govt.nz in Population : Death Rates - DMM Crude death rate (Maori and total population) (Annual-Dec).
 Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
@@
 
-![Figure 1: Ages of migration and ages of death have little overlap](md_figures/figure1.png){fig-alt="Young people mainly migrate, old people mainly die" fig-align="center"}
+![Ages of migration and ages of death have little overlap](md_figures/figure1.png){fig-alt="2019 comparison of ages of migration (young) and ages of death (old)" fig-align="center"}
 
@@
 
-If 2020 had the same single year of age death rates as 2019, and new migrant 2020 death rates were the same as the equivalently aged non-migratory population in 2019, then we can calculate the expected deaths from the 35,744 reduction in nett migration from 2019 to 2020.
-Using single age death rates, the youthful migrating population would be expected to result in 85 fewer deaths for the 35,744 reduction in total population[^7].
+If 2020 had the same single year of age death rates as 2019, and new migrant 2020 death rates were the same as the equivalently aged non-migratory population in 2019, then we can calculate the expected deaths from the 35,744 reduction in net migration from 2019 to 2020.
+Using single age death rates, the youthful migrating population would be expected to result in 85 fewer deaths for the 35,744 reduction in total population[^8].
 
-[^7]: NZ migration data available from https://infoshare.stats.govt.nz in Tourism : International Travel and Migration - ITM : Table: Estimated migration by direction, age group and sex, 12/16-month rule (Annual-Dec).
+[^8]: NZ migration data available from https://infoshare.stats.govt.nz in Tourism : International Travel and Migration - ITM : Table: Estimated migration by direction, age group and sex, 12/16-month rule (Annual-Dec).
 NZ death data available in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
@@
 Conversely, the expected increase in deaths due to aging can be calculated by aging the 2019 resident population by one year, and subtracting annual deaths of that age, and applying the single year death rates of 2019 (the same assumption as deaths expected from migration).
-For the resident population, ecluding under one yer olds, there would be an expected rise of 837 deaths while the total population decreased by about 33,963[^8].
+For the resident population, excluding under one year olds, there would be an expected rise of 837 deaths while the total population decreased by about 33,963[^9].
 
-[^8]: NZ death data available from https://infoshare.stats.govt.nz in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
+[^9]: NZ death data available from https://infoshare.stats.govt.nz in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
 NZ population data available in Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec).
@@
 
-> *If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process.*
+> *'If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process.'*
 
 The assumption that aging is a long-term process so should show up as a long slow process only applies if there are no sudden demographic shifts among age groups that alter the risk of death.
-The number of 75 and older residents, when compared to the 2012-2016 trend, was also accelerating in 2019.
+As the pre-COVID-19 era, tautologically, the baseline for COVID-19 comparisons and trends fitted to it automatically best fit the data of the period, to see the pre-COVID-19 rise requires using a slightly earlier baseline.
+
+The number of 75 and older residents has been growing at an accelerating rate in the decade prior to COVID-19.
 This matches a "sudden" demographic shift of an acceleration in the age distribution that began decades earlier and naturally aged into ranges that play a more significant role in total deaths.
 
-![Figure 2: The rate of increase in elderly accelerated in 2019](md_figures/figure2.png){fig-alt="The number of living elderly has been growing faster than trend prior to 2020" fig-align="center"} But the dramatic increase in elderly only contributes to mortality as one of several interacting factors, such as the age specific risk environment.
+![The elderly population grew at an accelerating rate in the decade before COVID-19](md_figures/figure2.png){fig-alt="The number of living elderly has been growing faster than trend" fig-align="center"}
+
+But the dramatic increase in elderly is only one of several interacting factors contributing to mortality.
+Others are factors influencing the age specific risk environment.
 A more direct measure of the importance of age stratification is the difference in outcome between the Crude Mortality Rate (ignoring age structure) and the Age Standardised Mortality Rate (based on age stratified death rates).
-The difference in excess mortality between those two methods is exactly and only the difference caused by not using age specific death rates.
+The difference in excess mortality between those two methods of 2% to 4% is exactly, and solely, the difference caused by not using age specific death rates.
 
-Of more surprise than the comparison to age standartdised methods is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the normal method of comparing death rates while ignoring the effects of age.
+Of more surprise than the comparison to age standardised methods is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the normal method of comparing death rates while ignoring the effects of age.
 While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same outcome, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of Karlinsky & Kobak.
@@
 
-David Hood,
-
-University of Otago
-
 ## References

### 2024-03-31 rawfiles/letter_to_ed_repo_pdf.qmd
final... ish

--- 2024-03-31 rawfiles/letter_to_ed_docx.qmd
@@
 title: "Comment on Cumulative Excess Deaths in New Zealand in the COVID-19 Era: Biases from Ignoring Changes in Population Growth Rates"
-format: docx
+author: "David Hood"
+execute:
+  keep-md: true
+format:
+  pdf:
+    fig-pos: 'h'
 bibliography: references.bib
@@
 ---
+
+This is the Github repo for my letter of comment on Gibson's article.
+The letter references the code data files in the support folder available here.
+For context, the final draft of the letter is included below, though this final draft may differ slightly from the published version.
 
@@
 Death rates for an age stratified group, for a time period in a given location, measure the rate at which that age group dies.
-These can be directly compared to death rates in other time periods and locations to measure if the group experience is more, or less, deadly.
+These can be directly compared to death rates in other time periods and locations to measure if the group experience is more or less deadly.
 As it is a rate, it automatically adjusts for differences in the size of the age group.

### 2024-03-31 supportFiles/Fig1_immigration.R

@@ -1,73 +1,0 @@
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-source("theme.R")
-six_cols <- colorblind_pal()(6)
-
-VSD <- read.csv("StatsNZData/VSD_DeathsDec_age_sex.csv",
-                skip=1, col.names = c("Year","Age",
-                                      "Male","Female")) |> 
-  filter(!is.na(Male)) |> 
-  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
-  fill(Year) |> 
-  mutate(Year = as.numeric(Year)) |> 
-  filter(Year == 2019)|> 
-  slice(26:126) |> 
-  mutate(Deaths = Male + Female,
-         Age = ifelse(Age == "Less than 1 year",
-                      "0 years",Age),
-         AgeN = as.numeric(gsub(" yea.*","",Age)),
-         AgeN = ifelse(AgeN > 89, 90, AgeN)) |> 
-  summarise(.by=AgeN, Deaths=sum(Deaths))
-
-DPE <- read.csv("StatsNZData/DPE_PopDecMean_age_sex.csv",
-                skip=2, col.names = c("Year","Age",
-                                     "Male","Female"))  |> 
-  filter(!is.na(Male)) |> 
-  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
-  fill(Year) |> 
-  mutate(Year = as.numeric(Year)) |> 
-  filter(Year == 2019) |> 
-  slice(c(1:95,134)) |> 
-  mutate(Lives = Male + Female,
-         AgeN = as.numeric(gsub(" Yea.*","",Age)),
-         AgeN = ifelse(AgeN > 89, 90, AgeN)) |> 
-  summarise(.by=AgeN, Lives=sum(Lives))
-
-ITM <- read.csv("StatsNZData/ITM_2019AgeSex.csv", skip=4, 
-                col.names = c("Age","Female","Male")) |> 
-  slice(c(1:90,127)) |> 
-  mutate(Moves = as.numeric(Male) + as.numeric(Female),
-         AgeN = as.numeric(gsub(" Yea.*","",Age)),
-         AgeN = ifelse(AgeN > 89, 90, AgeN)) |> 
-  summarise(.by=AgeN, Moves=sum(Moves))
-
-Deaths <- VSD |> 
[…]

### 2024-03-31 supportFiles/Fig2_75plus.R

@@ -1,42 +1,0 @@
-# 75+ trend
-
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-source("theme.R")
-six_cols <- colorblind_pal()(6)
-
-autoAge <- paste(0:94,"Years")
-autoAge <- c(autoAge,"95 Years and Over")
-DPE <- read.csv("StatsNZData/DPE_PopDecMean_age_sex.csv", skip=2,
-                col.names = c("Year","Age","Male","Female")) |> 
-  mutate(Year = ifelse(Year == " ", NA_character_,Year)) |> 
-  fill(Year) |> 
-  filter(Age %in% autoAge) |> 
-  mutate(AgeN = as.numeric(gsub(" .*","",Age)),
-         Year = as.numeric(Year)) |> 
-  gather(key="Sex",value="MeanPopulation", Male:Female) |> 
-  filter(AgeN >=75) |> 
-  summarise(.by=Year,
-            Living75plus = sum(MeanPopulation))
-
-LifeModel <- lm(Living75plus ~ Year, data=DPE |> filter(Year %in% 2012:2016))
-DPE$Expected <- LifeModel$coefficients[1] + LifeModel$coefficients[2]*DPE$Year
-
-ggplot(DPE, aes(x=Year, y=Living75plus))+
-  geom_point() +
-  geom_line(aes(y=Expected), colour=six_cols[2], linetype=2) +
-  geom_line(data=DPE |> filter(Year < 2017), aes(y=Expected), colour=six_cols[2]) +
-  theme_davidhood() +
-  scale_x_continuous(breaks=c(2012,2016,2020))+
-  labs(title="Actual 75 years and older vs 2012-2016 trend",
-       y="Residents 75 years and Older", x="Year",
-       caption="Source: infoshare.stats.govt.nz
-Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec)")
-ggsave(filename="../rawfiles/md_figures/figure2.png",
-       height=4, width = 6, dpi=300, units = "in", bg = "white")
-ggsave(filename="../md_figures/figure2.png",
-       height=4, width = 6, dpi=300, units = "in", bg = "white")

### 2024-03-31 supportFiles/Figure 1 Migration Deaths.R

@@ -1,0 +1,136 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+
+# ITM Migration
+# ITM has lowest open category of 90+
+# so sets the age limits for others.
+colist <- c(1,(c(1:90,127) * 2))
+ITMf <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
+  select(all_of(colist)) 
+names(ITMf) <- c("Year", paste0("X",0:90))
+ITMflong <- ITMf |> 
+  filter(!is.na(X20)) |> 
+  gather(key="AgeText", value="Net", -Year) |> 
+  mutate(Sex="Female")
+
+colist <- c(1,((c(1:90,127) * 2) +1))
+ITMm <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
+  select(all_of(colist)) 
+names(ITMm) <- c("Year", paste0("X",0:90))
+ITMmlong <- ITMm |> 
+  filter(!is.na(X20)) |> 
+  gather(key="AgeText", value="Net", -Year) |> 
+  mutate(Sex="Male")
+
+ITM <- bind_rows(ITMflong,ITMmlong) |> 
+  mutate(AgeN = as.numeric(gsub("X","",AgeText)),
+         Year = as.numeric(Year)) |> 
+  filter(Year == 2019) |> 
+  summarise(.by=AgeN,
+            NetMigration = sum(Net))
+
+# VSD deaths
+
+VSD_male <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
+  select(c(1,27:127)) |> 
+  rename(Year = X., X0.years = Less.than.1.year, X1.years=X1.year) |> 
+  filter(!is.na(X0.years)) |> 
+  gather(key="Agetext", value="Deaths", -Year) |> 
+  mutate(Ageshort = gsub("^X","",Agetext),
+         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
+         Sex="Male",
+         Year = as.numeric(Year)) |> 
+  select(Year, Sex, AgeN, Deaths)
+VSD_female <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
+  select(c(1,154:254)) |> 
+  rename(Year = X., X0.years = Less.than.1.year.1, X1.years=X1.year.1) |> 
[…]

### 2024-03-31 supportFiles/theme.R

@@ -1,30 +1,24 @@
 six_cols <- colorblind_pal()(6)
-footer <- paste0("\nGraph made ", trimws(format(Sys.time(), format="%e %b %Y")),". Graphs: thoughtfulnz.quarto.pub/nzcovidreport/ Contact: thoughtfulnz on mastodon.nz or bsky.app\n")
 
 theme_davidhood <- function(){
-  theme_minimal(base_family="Open Sans",
+  theme_minimal(base_family="Arial",
                 base_size = 9) %+replace%  
-    theme(axis.line.x = element_line(linewidth=0.1),
-          axis.line.y = element_line(linewidth=0.1),
+    theme(axis.line.x = element_line(linewidth=0.2),
+          axis.line.y = element_line(linewidth=0.2),
           axis.ticks = element_line(linewidth=0.2),
-          axis.title.y.left = element_text(margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")),
-          strip.background = element_rect(fill= "#FFFFFF", colour="#EFEFEF"),
-          strip.text = element_text(margin = margin(t = 5, r = 5, b = 5, l = 5, unit = "pt")),
-          strip.placement = "inside",
+          axis.title.y.left = element_text(margin = margin(t = 2, r = 2, b = 2, l = 2, unit = "pt")),
           panel.background = element_rect(fill = "#FFFFFF", colour = "#FFFFFF"),
-          panel.spacing.x = unit(9, "points"),
-          panel.spacing.y = unit(3, "points"),
           panel.grid = element_blank(),
-          plot.title = element_text(lineheight = 1.18, size=11,
+          plot.title = element_text(lineheight = 1.18, size=10,
                                     margin=margin(t = 5, r = 0, b = 5, l = 15, unit = "pt"),
                                     hjust=0, vjust=0),
           plot.subtitle = element_text(lineheight = 1.18, size=8,
                                     margin=margin(t = 0, r = 15, b = 5, l = 15, unit = "pt"),
                                     hjust=0),
-          plot.background = element_rect(fill = "#FCFCFC", colour="#FCFCFC"),
+          plot.background = element_rect(fill = "#FFFFFF", colour="#FFFFFF"),
           plot.caption = element_text(margin=margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
                                       lineheight = 1.15,
-                                      size=7, hjust=1),
+                                      size=6.5, hjust=1),
           plot.caption.position = "plot")
   
 }

### 2024-03-31 rawfiles/letter_to_ed_repo_pdf.qmd
preprint submitted

@@
 The letter references the code data files in the support folder available here.
-For context, the final draft of the letter is included below, though this final draft may differ slightly from the published version.
+For context, the final draft/ preprint of the letter is included below, though this final draft may differ slightly from any published version.
+
+------------------------------------------------------------------------
 

### 2024-07-07 rawfiles/letter_to_ed_docx.qmd
reviewer changes

@@
 
--   Being based on trends of deaths, it enables a common, consistent basis comparison with countries that lack age stratified current population information.
+-   Being based on trends only of deaths, it enables a common, consistent basis comparison with countries that lack age stratified current population information, or indeed countries that lack any current population information.
 As Gibson notes, Karlinsky and Kobak were initially able to provide comparison data for 103 countries (later 127).
-By contrast the mortality.org has weekly death rates by age (suitable for age standard mortality comparisons) for 38 countries[^3].
+By contrast the mortality.org website has weekly death rates by age (suitable for age standard mortality comparisons) for 38 countries[^3].
 -   It makes weekly comparisons available.
@@
 
-While the existence of centuries of mortality measurement (not mentioned in Gibson's article) does not directly address if New Zealand medical professionals are aware that best practice mortality calculations recognise that people of different ages die at different rates when using the publicly available Our World in Data graphs in media pieces, it does establish that there is an established methodology, not mentioned in Gibson's article, underpinning claims of New Zealand's negative excess, and that method has been used for centuries, world-wide, across multiple disciplines.
-We can check if New Zealand medical professionals are aware of this by asking them.
+The existence of centuries of mortality measurement establishes that it is best practice in mortality calculations to recognise that people of different ages die at different rates not mentioned in Gibson's article.
+While New Zealand medical professionals have used the Karlinsky and Kobak method/ Our World in Data graphs in media presentations, it is not established if that was for the designed purpose while still being aware of more accurate methods or because K & K was considered definitive.
+We can check if New Zealand medical professionals are aware of best practice methods by asking them.
 
@@
 
+These methods use StatsNZ annual counts of resident deaths and resident population, rather than the weekly counts from mortality.org, as mortality.org takes the raw StatsNZ data and interpolates ages to the weekly totals based on the typical annual age distribution of deaths.
+This is why New Zealand data for weekly death counts at mortality.org contains decimal places.
+
+Using detailed annual StatsNZ age counts of resident deaths and population prevents inaccuracy in results from overly broad age groups.
+In 2019 the age specific death rate[^6] of 10-14 year olds (0.17 per 1000 residents of that age per year) approximately doubles by age 15-19 (0.41), and doubles again by 35-39 (0.83).
+More doubling takes place by 45-49 (1.95), 55-59 (4.45), and 65-69 (10.27).
+With the speed of doubling being even faster in older aged that make up the majority of deaths, mortality measurements can be very sensitive to even a few years of change in the population balance.
+
+[^6]: https://infoshare.stats.govt.nz/ Population : Death Rates - DMM : Age-specific death rates by sex, December years (total population) (Annual-Dec)
+
 Excess mortality is the difference between actual mortality and what is expected.
-The expected value is an estimate based on certain underlying premises, so excess mortality calculations using any method are sensitive what is considered "normal": the baseline.
+The expected value is an estimate based on certain underlying premises, so excess mortality calculations using any method are sensitive to what is considered "normal": the baseline.
 For Age Standardised Mortality the regular process is to use a linear trend of preceding years, and where trying to measure the impact of a major multi-year event like a pandemic, the linear trend of pre-pandemic years.
@@
 While many international studies use a 5 year baseline range to increase the number of countries for which weekly level data is available, in practice the results for New Zealand do not differ substantively between a 5 (2015-2019) and 7 (2013-2019) year baseline period.
-Using a 2022 standard population and a 2015-2019 baseline gives -2% excess mortality for 2020-2022, while using a 2013-2019 baseline gives -2%[^6].
+Using a 2022 standard population and a 2015-2019 baseline gives -2% excess mortality for 2020-2022, while using a 2013-2019 baseline gives -2%[^7].
 
-[^6]: Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
+[^7]: Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
 
@@
 
-Though not mentioned in Gibson's paper, using total population (without aging) as a model term would normally be functionally equivalent to the commonly known Crude Mortality Rate (deaths among the total population / total living population).
-A 2015-2019 linear regression of the annual CMR gives an 2020-2022 cumulative excess of 0%[^7].
+Though not mentioned in Gibson's paper, the commonest method for measuring deaths in relation to total population (without aging) is the Crude Mortality Rate (deaths among the total resident population / total living resident population).
+This is available for any country producing both an annual count of resident deaths and an annual resident population count.
+A 2015-2019 linear regression of the annual CMR[^8] gives an 2020-2022 cumulative excess of 0%.
 
-[^7]: NZ crude death rates are available from https://infoshare.stats.govt.nz in Population : Death Rates - DMM Crude death rate (Maori and total population) (Annual-Dec).
+[^8]: NZ crude death rates are available from https://infoshare.stats.govt.nz in Population : Death Rates - DMM Crude death rate (Maori and total population) (Annual-Dec).
 Raw data and analysis code at https://github.com/thoughtfulbloke/letter2ed in the support_files folder.
@@
 The reason for Gibson's model performing so poorly compared to Karlinsky & Kobak lies in the age imbalance of death rates.
-Young people frequently migrate and infrequently die.
+Young people frequently migrate and infrequently die (see Figure 1).
 Old people infrequently migrate and frequently die.
@@
 If 2020 had the same single year of age death rates as 2019, and new migrant 2020 death rates were the same as the equivalently aged non-migratory population in 2019, then we can calculate the expected deaths from the 35,744 reduction in net migration from 2019 to 2020.
-Using single age death rates, the youthful migrating population would be expected to result in 85 fewer deaths for the 35,744 reduction in total population[^8].
+Using single age death rates, the youthful migrating population would be expected to result in 85 fewer deaths for the 35,744 reduction in total population[^9].
 
-[^8]: NZ migration data available from https://infoshare.stats.govt.nz in Tourism : International Travel and Migration - ITM : Table: Estimated migration by direction, age group and sex, 12/16-month rule (Annual-Dec).
+[^9]: NZ migration data available from https://infoshare.stats.govt.nz in Tourism : International Travel and Migration - ITM : Table: Estimated migration by direction, age group and sex, 12/16-month rule (Annual-Dec).
 NZ death data available in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
@@
 Conversely, the expected increase in deaths due to aging can be calculated by aging the 2019 resident population by one year, and subtracting annual deaths of that age, and applying the single year death rates of 2019 (the same assumption as deaths expected from migration).
-For the resident population, excluding under one year olds, there would be an expected rise of 837 deaths while the total population decreased by about 33,963[^9].
+For the resident population, excluding under one year olds, there would be an expected rise of 837 deaths while the total population decreased by about 33,963[^10].
 
-[^9]: NZ death data available from https://infoshare.stats.govt.nz in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
+[^10]: NZ death data available from https://infoshare.stats.govt.nz in Population : Deaths - VSD : Deaths by age and sex (Annual-Dec).
 NZ population data available in Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec).
@@
 
-Gibson asserted that it was not important to incorporate aging:
-
-> *'If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process.'*
+Gibson asserted that it was not important to incorporate aging "If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process." (Gibson, 2024, pp.5)
 
@@
 
-The number of 75 and older residents has been growing at an accelerating rate in the decade prior to COVID-19.
+The number of 75 and older residents has been growing at an accelerating rate in the decade prior to COVID-19 (see Figure 2).
 This matches a "sudden" demographic shift of an acceleration in the age distribution that began decades earlier and naturally aged into ranges that play a more significant role in total deaths.
@@
 
+## Conclusion
+
 Specific methods are employed to achieve particular objectives, such as broadening the scope of comparison.

### 2024-07-07 supportFiles/Figure 1 Migration Deaths.R

@@ -121,7 +121,7 @@ ggplot(combo, aes(x=AgeN, ymax=Percent, fill=Series))+
   scale_fill_manual(values=c("#000000", "#CCCCCCAA"), name="Series:")+
   scale_x_continuous(breaks=c(0,30,60,90),
                      labels=c("under 1","30","60","90+")) +
-  labs(title="2019 Deaths vs. Net Migrantion",
+  labs(title="2019 Deaths vs. Net Migration",
        subtitle="as percentage of resident population",
        y="Percentage", x="Age",
        caption="Source: infoshare.stats.govt.nz: Deaths by age and sex (Annual-Dec)

### 2024-07-07 supportFiles/footnotes4and5.R

@@ -93,7 +93,7 @@ DMM <- read.csv("RawData/DMM_SDR_TotPop.csv",
   rename(Year=1,SDR=2) |> 
   mutate(Year = as.numeric(Year))
 DMModel <- lm(SDR ~ Year, 
-              DMM |> filter(Year %in% 2013:2019))
+              DMM |> filter(Year %in% 2015:2020))
 DMM$Expected <- DMModel$coefficients[1] +
   DMModel$coefficients[2] * DMM$Year
 cumulatime <- DMM |> filter(Year >= 2020, Year <=2022)

### 2024-07-07 rawfiles/letter_to_ed_docx.qmd
reviewer changes

@@
 
-This is a response to the [@gibson2024] paper in which, to give my summary of the paper, Gibson asserts that New Zealand authorities' measurement of mortality in the pandemic period is based on a flawed metric.
+For hundreds of years, correctly measuring mortality has been a concern of actuaries, demographers, and health researchers.
+Accurate measurement should form the basis of insurance risk pricing and government policy.
+Writing in this journal, Gibson [@gibson2024] asserts that New Zealand authorities' measurement of mortality in the pandemic period is based on a flawed metric.
 Gibson's own metric, based on regression of deaths with an additive component of total population, is asserted as a more accurate model and produces approximately 4% excess mortality for the 2020-2022 period.
@@
 
-For centuries, it has been established practice among actuaries, demographers, and health researchers to measure mortality using age-adjusted death rates where such data is available [@price1773].
+For centuries, it has been established professional practice to measure mortality using age-adjusted death rates where such data is available [@price1773].
 This method supersedes the use of Crude Mortality Rates[^2] by acknowledging that different ages have different risks of death, so comparing age-based risk of death on a uniform basis.
@@
 Age Standardised Mortality directly adjusts for both population growth and aging, as it is based on the number of deaths of the current resident population divided by the current resident population, stratified by age group and time period.
-This directly measures the population change effects Gibson is trying to model.
+This directly measures the population change effects Gibson is trying to model, and does so at the level of residents of that age in each time period.
 
@@
 The existence of centuries of mortality measurement establishes that it is best practice in mortality calculations to recognise that people of different ages die at different rates not mentioned in Gibson's article.
-While New Zealand medical professionals have used the Karlinsky and Kobak method/ Our World in Data graphs in media presentations, it is not established if that was for the designed purpose while still being aware of more accurate methods or because K & K was considered definitive.
+While New Zealand medical professionals have used the Karlinsky and Kobak method/ Our World in Data graphs in media presentations, it is not established if that was for the designed practical purposes while still being aware of more accurate methods or because K & K was considered definitive.
 We can check if New Zealand medical professionals are aware of best practice methods by asking them.
@@
 
-These methods use StatsNZ annual counts of resident deaths and resident population, rather than the weekly counts from mortality.org, as mortality.org takes the raw StatsNZ data and interpolates ages to the weekly totals based on the typical annual age distribution of deaths.
-This is why New Zealand data for weekly death counts at mortality.org contains decimal places.
-
-Using detailed annual StatsNZ age counts of resident deaths and population prevents inaccuracy in results from overly broad age groups.
+For calculating age specific death rates, using detailed annual StatsNZ age counts of resident deaths and population prevents inaccuracy in results from overly broad age groups.
 In 2019 the age specific death rate[^6] of 10-14 year olds (0.17 per 1000 residents of that age per year) approximately doubles by age 15-19 (0.41), and doubles again by 35-39 (0.83).
@@
 With the speed of doubling being even faster in older aged that make up the majority of deaths, mortality measurements can be very sensitive to even a few years of change in the population balance.
+Any method calculating cumulative excess using shorter individual time periods should, if correct, produce the same result for the same total time period.
 
@@
 
-Karlinsky & Kobak also appears superior for comparisons between countries, as Gibson's results are a worse approximation with ASMR based methods calculated by other countries statistical authorities.
+Karlinsky & Kobak also appears superior for comparisons between countries, as Gibson's results are a worse approximation with ASMR (Age Specific Mortality Rate) based methods calculated by other countries statistical authorities.
 The Australian Bureau of Statistic's age based mortality analysis has excess mortality of 3.4% across 2020-2022 [@australianbureauofstatistics2023].
@@
 
+From comparison to best practice ASMR methods, K&K are closer in result than Gibson's method.
+Further, given the results of ASMR, New Zealand and international professionals have a sound basis for determining New Zealand's excess mortality as negative for the covid period.
+
 ## References

### 2024-07-12 rawfiles/README.qmd
tidying repo to match final version also added Excel example

---
date: last-modified
author: "David Hood"
title: "Comment on Cumulative Excess Deaths in New Zealand in the COVID-19 Era: Biases from Ignoring Changes in Population Growth Rates"
format: gfm
bibliography: references.bib
csl: apa.csl
---

This is the Github repo for my letter of comment on Gibson's article. The letter references the code data files in the supportFiles folder available here. For context, the final draft/ preprint of the letter is included below, though this final draft may differ slightly from any published version.

While the data files and R code for generating the results and figures is available in the support folder, as not everyone is familiar with R, I am also providing an [Excel file in the supportFiles](supportFiles/ASMR_in_Excel.xlsx) folder which implements the normal Age Standardised Mortality steps so that people can review and test the logic themselves. It shows the steps of:

1)  Obtain annual resident population by age data from https://infoshare.stats.govt.nz : Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec) : Mean year ended, both Male and Female, 0 -94 then 95+, years 2013-2022

2)  Obtain annual resident death by age data from https://infoshare.stats.govt.nz : Population : Deaths - VSD : Deaths by age and sex (Annual-Dec) : both Male and Female, 0 -99 then 100+, years 2013-2022

3)  Combine the 95 to 100 + categories in the death data to the same age range as the population data (95+) in order to be able to match people of the same age in the same year

4)  calculate death rates for each age group of residents for each year.

5)  Multiple the individual age death rates in each year by a standard population, so that total death changes between years are considered on the basis of the same population and age structure in every year, to avoid mistaking non-fatal changes in population size and structure as changes in risk of death. In this example, the New Zealand population of 2022 is used

6)  Total the standard deaths for each year, to generate the number (based on death rates) of deaths if the population had been the same as the year of the standard population, but the age specific risk of death was the same as the historical period.

7)  Preform a regression on the years 2013-2019 to account for trends beyond changes to the population value. The expected value as a result of the regression corrects for historical trends such as the improvement in health understanding and care (as per the article 2013-2019 is considered the optimal baseline for Australasia without triggering structural bias)

8)  Project the regression trend into 2020 through 2022 to determine expected deaths given a population structure of the standard year. Compared them to actual standardised deaths

9)  Cumulative excess is sum(actual)/sum(expected) - 1 expressed as a percentage.

The letter of comment preprint included below, as well as providing a [PDF version of the preprint](letter_to_ed_pdf.pdf) among the files of this repo:

------------------------------------------------------------------------

{{< include base_content.qmd >}}

### 2024-07-12 rawfiles/base_content.qmd
tidying repo to match final version also added Excel example

--- 2024-07-07 rawfiles/letter_to_ed_repo_pdf.qmd
@@
 ---
-date: last-modified
-title: "Comment on Cumulative Excess Deaths in New Zealand in the COVID-19 Era: Biases from Ignoring Changes in Population Growth Rates"
-author: "David Hood"
-execute:
-  keep-md: true
-format:
-  pdf:
-    fig-pos: 'h'
 bibliography: references.bib
@@
 
-This is the Github repo for my letter of comment on Gibson's article.
-The letter references the code data files in the support folder available here.
-For context, the final draft/ preprint of the letter is included below, though this final draft may differ slightly from any published version.
-
-------------------------------------------------------------------------
-
 For hundreds of years, correctly measuring mortality has been a concern of actuaries, demographers, and health researchers.
 Accurate measurement should form the basis of insurance risk pricing and government policy.
-Writing in this journal, Gibson [@gibson2024] asserts that New Zealand authorities' measurement of mortality in the pandemic period is based on a flawed metric.
-Gibson's own metric, based on regression of deaths with an additive component of total population, is asserted as a more accurate model and produces approximately 4% excess mortality for the 2020-2022 period.
-If New Zealand authorities are using flawed metrics with significantly different results to reality, this would have profound governance and policy implications.
+Writing in this journal, Gibson [@gibson2024] asserts that New Zealand authorities’ measurement of mortality in the pandemic period is based on a flawed metric.
+Gibson’s own metric, based on regression of deaths with an additive component of total population, is asserted as a more accurate model and produces approximately 4% excess mortality for the 2020-2022 period.
+If New Zealand authorities are using flawed metrics with significantly different results to reality, the thousands of unnoticed deaths would have profound governance and policy implications.
 
-In particular, the paper asserts two broad claims:
+In particular, I interpret the paper as asserting two broad claims:
 
@@
 
-For centuries, it has been established professional practice to measure mortality using age-adjusted death rates where such data is available [@price1773].
+It has been established professional practice for centuries to measure mortality using age-adjusted death rates where such data is available [@price1773].
 This method supersedes the use of Crude Mortality Rates[^2] by acknowledging that different ages have different risks of death, so comparing age-based risk of death on a uniform basis.
@@
 
-While calculation of New Zealand's cumulative excess mortality for 2020-2022 can vary by a couple of percentage points depending on the exact method adopted, using an age-standardised death rates method the UK's Institute and Faculty of Actuaries finds New Zealand's excess mortality from week 10 2020 to the end of 2022 to be -4% (that there are 4% fewer deaths than would be expected based on pre-COVID-19 trends) [@instituteandfacultyofactuaries2024].
+While calculation of New Zealand’s cumulative excess mortality for 2020-2022 can vary by a couple of percentage points depending on the exact method adopted, using an age-standardised death rates method the UK’s Institute and Faculty of Actuaries finds New Zealand’s excess mortality from week 10 2020 to the end of 2022 to be -4% (that there are 4% fewer deaths than would be expected based on pre-COVID-19 trends) [@instituteandfacultyofactuaries2024].
 
-The Our World in Data graph, and the Karlinsky and Kobak method that underlies it, is an approximation (if it is a better or worse approximation to Gibson's method will be discussed later).
+The Our World in Data graph, and the Karlinsky and Kobak method that underlies it, is an approximation (if it is a better or worse approximation to Gibson’s method will be discussed later).
 While recognising the Our World In Data graph is an approximation, there are still several practically useful implicit features about it:
@@
 
-The existence of centuries of mortality measurement establishes that it is best practice in mortality calculations to recognise that people of different ages die at different rates not mentioned in Gibson's article.
-While New Zealand medical professionals have used the Karlinsky and Kobak method/ Our World in Data graphs in media presentations, it is not established if that was for the designed practical purposes while still being aware of more accurate methods or because K & K was considered definitive.
+Centuries of mortality measurement establishes that it is best practice in mortality calculations to recognise that people of different ages die at different rates.
+While New Zealand medical professionals have used the Karlinsky and Kobak method/ Our World in Data graphs in media presentations, it is not established if that was for the above practical benefits while still being aware of more accurate methods, or because K & K was considered definitive.
 We can check if New Zealand medical professionals are aware of best practice methods by asking them.
@@
 
-Though not mentioned in Gibson's paper, the commonest method for measuring deaths in relation to total population (without aging) is the Crude Mortality Rate (deaths among the total resident population / total living resident population).
+Though not mentioned in Gibson's paper, the commonest method for measuring deaths in relation to total population (without adjusting for aging) is the Crude Mortality Rate (deaths among the total resident population / total living resident population).
 This is available for any country producing both an annual count of resident deaths and an annual resident population count.
 A 2015-2019 linear regression of the annual CMR[^8] gives an 2020-2022 cumulative excess of 0%.
+This establishes that a method not adjusting for age, as might be logically expected in a country with an aging population, gives a higher value for excess than a method that adjusts for age.
 
@@
 
-Gibson's decision not to apply total population but not age to the model minimises an expected increase in deaths of an approximate order of magnitude greater than the effects of migration.
+Gibson’s decision not to apply total population but not age to the model minimises an expected increase in deaths of an approximate order of magnitude greater than the effects of migration.
 But all these complexities are avoided by incorporating both migration and aging by using age standardised mortality.
 
-Gibson asserted that it was not important to incorporate aging "If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process." (Gibson, 2024, pp.5)
+Gibson asserted that it was not important to incorporate aging “If societal aging was a cause of the increased number of deaths, as Gabel and Knox (2023) put forward as a reason for the excess mortality in 2022, it should also show up prior to the COVID-19 era, given that societal aging is a long-term process.” (Gibson, 2024, pp.5)
 
 The assumption that aging is a long-term process so should show up as a long slow process only applies if there are no sudden demographic shifts among age groups that alter the risk of death.
-As the pre-COVID-19 era, tautologically, the baseline for COVID-19 comparisons and trends fitted to it automatically best fit the data of the period, to see the pre-COVID-19 rise requires using a slightly earlier baseline.
+In the pre-COVID-19 era, tautologically, the baseline for COVID-19 comparisons and trends are automatically best fit the data of the pre-covid period, to see the pre-COVID-19 rise is aided by using a slightly earlier baseline.
 
 The number of 75 and older residents has been growing at an accelerating rate in the decade prior to COVID-19 (see Figure 2).
-This matches a "sudden" demographic shift of an acceleration in the age distribution that began decades earlier and naturally aged into ranges that play a more significant role in total deaths.
+This matches a “sudden” demographic shift of an acceleration in the age distribution that began decades earlier and naturally aged into ranges that play a more significant role in total deaths.
 
@@
 A more direct measure of the importance of age stratification is the difference in outcome between the Crude Mortality Rate (ignoring age structure) and the Age Standardised Mortality Rate (based on age stratified death rates).
-The difference in excess mortality between those two methods of 2% to 4% is exactly, and solely, the difference caused by not using age specific death rates.
+The difference in excess mortality between those two methods of 2% to 4% is the difference caused by not using age specific death rates.
 
-Of more surprise than the comparison to age standardised methods is that Gibson's results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the normal method of comparing death rates while ignoring the effects of age.
-While there is not enough information in the paper to exactly replicate Gibson's method and achieve the same outcome, part of the reason for the difference seems likely to be the unusual step of treating population as an additive contributor to deaths in his modification of Karlinsky & Kobak.
+Of more surprise than the comparison to age standardised methods is that Gibson’s results for 2020-22 less well approximate the crude mortality rate than Karlinsky & Kobak, since the crude mortality rate is the normal method of comparing death rates while ignoring the effects of age.
+This may be the result of the unusual step of treating population as an additive contributor to deaths in his modification of Karlinsky & Kobak.
 But, compared to the differences with methods acknowledging age, such problems are minor.
@@
 
-From comparison to best practice ASMR methods, K&K are closer in result than Gibson's method.
-Further, given the results of ASMR, New Zealand and international professionals have a sound basis for determining New Zealand's excess mortality as negative for the covid period.
+From comparison to best practice ASMR methods, K&K are closer in result than Gibson’s method.
+Further, given the results of ASMR, New Zealand and international professionals have a sound basis for determining New Zealand’s excess mortality as negative for the covid period.
 

### 2024-07-12 rawfiles/supportFiles/Figure 1 Migration Deaths.R

@@ -1,0 +1,136 @@
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+
+# ITM Migration
+# ITM has lowest open category of 90+
+# so sets the age limits for others.
+colist <- c(1,(c(1:90,127) * 2))
+ITMf <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
+  select(all_of(colist)) 
+names(ITMf) <- c("Year", paste0("X",0:90))
+ITMflong <- ITMf |> 
+  filter(!is.na(X20)) |> 
+  gather(key="AgeText", value="Net", -Year) |> 
+  mutate(Sex="Female")
+
+colist <- c(1,((c(1:90,127) * 2) +1))
+ITMm <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
+  select(all_of(colist)) 
+names(ITMm) <- c("Year", paste0("X",0:90))
+ITMmlong <- ITMm |> 
+  filter(!is.na(X20)) |> 
+  gather(key="AgeText", value="Net", -Year) |> 
+  mutate(Sex="Male")
+
+ITM <- bind_rows(ITMflong,ITMmlong) |> 
+  mutate(AgeN = as.numeric(gsub("X","",AgeText)),
+         Year = as.numeric(Year)) |> 
+  filter(Year == 2019) |> 
+  summarise(.by=AgeN,
+            NetMigration = sum(Net))
+
+# VSD deaths
+
+VSD_male <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
+  select(c(1,27:127)) |> 
+  rename(Year = X., X0.years = Less.than.1.year, X1.years=X1.year) |> 
+  filter(!is.na(X0.years)) |> 
+  gather(key="Agetext", value="Deaths", -Year) |> 
+  mutate(Ageshort = gsub("^X","",Agetext),
+         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
+         Sex="Male",
+         Year = as.numeric(Year)) |> 
+  select(Year, Sex, AgeN, Deaths)
+VSD_female <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
+  select(c(1,154:254)) |> 
+  rename(Year = X., X0.years = Less.than.1.year.1, X1.years=X1.year.1) |> 
[…]

### 2024-07-12 rawfiles/supportFiles/Figure 2 Elderly.R

@@ -1,0 +1,54 @@
+# 75+ trend
+
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(ggplot2)
+library(ggthemes)
+library(ggtext)
+source("theme.R")
+six_cols <- colorblind_pal()(6)
+
+DPE_male <- read.csv("RawData/DPE_PopDecMean_age_sex.csv", skip=3) |> 
+  select(c(1,131)) |> 
+  rename(Year = 1, seventyfivePlus = X75.Years.and.Over) |> 
+  filter(!is.na(seventyfivePlus)) |> 
+  mutate(Year = as.numeric(Year),
+         Sex = "Male")
+DPE_female <- read.csv("RawData/DPE_PopDecMean_age_sex.csv", skip=3) |> 
+  select(c(1,267)) |> 
+  rename(Year = 1, seventyfivePlus = X75.Years.and.Over.1) |> 
+  filter(!is.na(seventyfivePlus)) |> 
+  mutate(Year = as.numeric(Year),
+         Sex = "Female")
+
+DPE75plus <- bind_rows(DPE_male, DPE_female) |> 
+  summarise(.by=Year,
+            Subpopulation = sum(seventyfivePlus)) |> 
+  filter(Year >2006)
+
+###
+LifeModel <- lm(Subpopulation ~ Year, 
+                data=DPE75plus |> filter(Year %in% 2012:2016))
+DPE75plus$Expected <- LifeModel$coefficients[1] + 
+  LifeModel$coefficients[2] * DPE75plus$Year
+
+ggplot(DPE75plus, aes(x=Year, y=Subpopulation))+
+  geom_line(aes(y=Expected), colour="#AAAAAA", 
+            linetype=2)+
+  geom_line(data=DPE75plus |> filter(Year %in% 2012:2016), 
+            aes(y=Expected), 
+            colour="#AAAAAA") +
+  geom_point() +
+  theme_davidhood() +
+  scale_x_continuous(breaks=c(2010,2015,2020))+
+  labs(title="NZ Resident population 75 plus years\nvs 2012-2016 trend",
+       y="75 plus poopulation", x="Year",
+       caption="Source: infoshare.stats.govt.nz : Population : Population Estimates - DPE : 
+Estimated Resident Population by Age and Sex (1991+) (Annual-Dec)")
+ggsave(filename="../rawfiles/md_figures/figure2.tiff",
[…]

### 2024-07-12 rawfiles/supportFiles/footnotes5and7.R

@@ -1,0 +1,108 @@
+library(dplyr)
+library(tidyr)
+
+# Deaths, rearranged into long-style database form
+# Single years then 100+ by sex
+VSD_male <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
+  select(c(1,27:127)) |> 
+  rename(Year = X., X0.years = Less.than.1.year, X1.years=X1.year) |> 
+  filter(!is.na(X0.years)) |> 
+  gather(key="Agetext", value="Deaths", -Year) |> 
+  mutate(Ageshort = gsub("^X","",Agetext),
+         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
+         Sex="Male",
+         Year = as.numeric(Year)) |> 
+  select(Year, Sex, AgeN, Deaths)
+VSD_female <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
+  select(c(1,154:254)) |> 
+  rename(Year = X., X0.years = Less.than.1.year.1, X1.years=X1.year.1) |> 
+  filter(!is.na(X0.years)) |> 
+  gather(key="Agetext", value="Deaths", -Year) |> 
+  mutate(Ageshort = gsub("^X","",Agetext),
+         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
+         Sex="Female",
+         Year = as.numeric(Year)) |> 
+  select(Year, Sex, AgeN, Deaths)
+
+VSD <- bind_rows(VSD_male, VSD_female)
+
+# Population Male/Female single years then 95+
+DPE_male <- read.csv("RawData/DPE_PopDecMean_age_sex.csv", skip=3) |> 
+  select(c(1,2:96,135)) |> 
+  rename(Year = X.) |> 
+  filter(!is.na(X0.Years)) |> 
+  gather(key="Agetext", value="Population", -Year) |> 
+  mutate(Ageshort = gsub("^X","",Agetext),
+         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
+         Sex="Male",
+         Year = as.numeric(Year)) |> 
+  select(Year, Sex, AgeN, Population)
+DPE_female <- read.csv("RawData/DPE_PopDecMean_age_sex.csv", skip=3) |> 
+  select(c(1,138:232,271)) |> 
+  rename(Year = X.) |> 
+  filter(!is.na(X0.Years.1)) |> 
+  gather(key="Agetext", value="Population", -Year) |> 
[…]

### 2024-07-12 rawfiles/supportFiles/footnotes8.R

@@ -1,0 +1,19 @@
+library(dplyr)
+
+DMM <- read.csv("RawData/DMM_CDRDec_TotPop.csv",
+                skip=1) |> 
+  filter(!is.na(Total.Population)) |> 
+  rename(Year=1,CDR=2) |> 
+  mutate(Year = as.numeric(Year))
+DMModel <- lm(CDR ~ Year, 
+              DMM |> filter(Year %in% 2015:2019))
+DMM$Expected <- DMModel$coefficients[1] +
+  DMModel$coefficients[2] * DMM$Year
+cumulatime <- DMM |> filter(Year >= 2020, Year <=2022)
+percentage = 100 * sum(cumulatime$CDR) /
+  sum(cumulatime$Expected) - 100
+print(round(percentage,0))
+
+
+
+

### 2024-07-12 rawfiles/supportFiles/footnotes9and10.R

@@ -1,0 +1,149 @@
+# working out that if 2020 had the same death rates by age as 2019
+# and the loss in migration, how much would the death rate go up by.
+
+library(dplyr)
+library(tidyr)
+
+# ITM Migration
+# ITM has lowest open category of 90+
+# so sets the age limits for others.
+colist <- c(1,(c(1:90,127) * 2))
+ITMf <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
+  select(all_of(colist)) 
+names(ITMf) <- c("Year", paste0("X",0:90))
+ITMflong <- ITMf |> 
+  filter(!is.na(X20)) |> 
+  gather(key="AgeText", value="Net", -Year) |> 
+  mutate(Sex="Female")
+
+colist <- c(1,((c(1:90,127) * 2) +1))
+ITMm <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
+  select(all_of(colist)) 
+names(ITMm) <- c("Year", paste0("X",0:90))
+ITMmlong <- ITMm |> 
+  filter(!is.na(X20)) |> 
+  gather(key="AgeText", value="Net", -Year) |> 
+  mutate(Sex="Male")
+
+ITM <- bind_rows(ITMflong,ITMmlong) |> 
+  mutate(AgeN = as.numeric(gsub("X","",AgeText)),
+         Year = as.numeric(Year)) |> 
+  arrange(Sex, AgeN, Year) |> 
+  group_by(Sex,AgeN) |> 
+  mutate(followingYear = lead(Net),
+         change2following = followingYear - Net ) |> 
+  ungroup() |> 
+  select(Year,Sex,AgeN,Net,change2following)
+
+# VSD deaths
+
+VSD_male <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
+  select(c(1,27:127)) |> 
+  rename(Year = X., X0.years = Less.than.1.year, X1.years=X1.year) |> 
+  filter(!is.na(X0.years)) |> 
+  gather(key="Agetext", value="Deaths", -Year) |> 
+  mutate(Ageshort = gsub("^X","",Agetext),
+         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
+         Sex="Male",
+         Year = as.numeric(Year)) |> 
+  select(Year, Sex, AgeN, Deaths)
[…]

### 2024-07-12 rawfiles/supportFiles/theme.R

@@ -1,0 +1,24 @@
+six_cols <- colorblind_pal()(6)
+
+theme_davidhood <- function(){
+  theme_minimal(base_family="Arial",
+                base_size = 9) %+replace%  
+    theme(axis.line.x = element_line(linewidth=0.2),
+          axis.line.y = element_line(linewidth=0.2),
+          axis.ticks = element_line(linewidth=0.2),
+          axis.title.y.left = element_text(margin = margin(t = 2, r = 2, b = 2, l = 2, unit = "pt")),
+          panel.background = element_rect(fill = "#FFFFFF", colour = "#FFFFFF"),
+          panel.grid = element_blank(),
+          plot.title = element_text(lineheight = 1.18, size=10,
+                                    margin=margin(t = 5, r = 0, b = 5, l = 15, unit = "pt"),
+                                    hjust=0, vjust=0),
+          plot.subtitle = element_text(lineheight = 1.18, size=8,
+                                    margin=margin(t = 0, r = 15, b = 5, l = 15, unit = "pt"),
+                                    hjust=0),
+          plot.background = element_rect(fill = "#FFFFFF", colour="#FFFFFF"),
+          plot.caption = element_text(margin=margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
+                                      lineheight = 1.15,
+                                      size=6.5, hjust=1),
+          plot.caption.position = "plot")
+  
+}

### 2024-07-12 rawfiles/README.qmd
explained how to download excel file

@@
 
-While the data files and R code for generating the results and figures is available in the support folder, as not everyone is familiar with R, I am also providing an [Excel file in the supportFiles](supportFiles/ASMR_in_Excel.xlsx) folder which implements the normal Age Standardised Mortality steps so that people can review and test the logic themselves.
+While the data files and R code for generating the results and figures is available in the support folder, as not everyone is familiar with R, I am also providing an [Excel file in the supportFiles](supportFiles/ASMR_in_Excel.xlsx) folder (use the download raw files button in the upper right) which implements the normal Age Standardised Mortality steps so that people can review and test the logic themselves.
 It shows the steps of:

### 2024-07-12 rawfiles/README.qmd
reworded explanation on how to download excel file

--- 2024-07-12 rawfiles/README.qmd
@@
 
-While the data files and R code for generating the results and figures is available in the support folder, as not everyone is familiar with R, I am also providing an [Excel file in the supportFiles](supportFiles/ASMR_in_Excel.xlsx) folder which implements the normal Age Standardised Mortality steps so that people can review and test the logic themselves.
+While the data files and R code for generating the results and figures is available in the support folder, as not everyone is familiar with R, I am also providing an [Excel file in the supportFiles](supportFiles/ASMR_in_Excel.xlsx) folder (use the download raw files button in the upper right of the Excel file screen) which implements the normal Age Standardised Mortality steps so that people can review and test the logic themselves.
 It shows the steps of:

### 2024-07-13 rawfiles/README.qmd
changed support files folder name to exactly match text in article

--- 2024-07-12 rawfiles/README.qmd
@@
 This is the Github repo for my letter of comment on Gibson's article.
-The letter references the code data files in the supportFiles folder available here.
+The letter references the code data files in the support_files folder available here.
 For context, the final draft/ preprint of the letter is included below, though this final draft may differ slightly from any published version.
 
-While the data files and R code for generating the results and figures is available in the support folder, as not everyone is familiar with R, I am also providing an [Excel file in the supportFiles](supportFiles/ASMR_in_Excel.xlsx) folder which implements the normal Age Standardised Mortality steps so that people can review and test the logic themselves.
+While the data files and R code for generating the results and figures is available in the support_files folder, as not everyone is familiar with R, I am also providing an [Excel file in the support_files](support_files/ASMR_in_Excel.xlsx) folder (use the download raw files button in the upper right of the Excel file screen) which implements the normal Age Standardised Mortality steps so that people can review and test the logic themselves.
 It shows the steps of:
 
-1)  Obtain annual resident population by age data from https://infoshare.stats.govt.nz : Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec) : Mean year ended, both Male and Female, 0 -94 then 95+, years 2013-2022
+1)  Obtain annual resident population by age data from https://infoshare.stats.govt.nz : Population : Population Estimates - DPE : Estimated Resident Population by Age and Sex (1991+) (Annual-Dec) : Mean year ended, both Male and Female, 0 -94 then 95+, years 2013-2022.
+Because you need to know how many living people of each age are resident in an area in a given time period to calculate age specific death rates.
 
-2)  Obtain annual resident death by age data from https://infoshare.stats.govt.nz : Population : Deaths - VSD : Deaths by age and sex (Annual-Dec) : both Male and Female, 0 -99 then 100+, years 2013-2022
+2)  Obtain annual resident death by age data from https://infoshare.stats.govt.nz : Population : Deaths - VSD : Deaths by age and sex (Annual-Dec) : both Male and Female, 0 -99 then 100+, years 2013-2022.
+Because you need to know how many people die of each age who are resident in an area in a given time period to calculate age specific death rates.
 
-3)  Combine the 95 to 100 + categories in the death data to the same age range as the population data (95+) in order to be able to match people of the same age in the same year
+3)  Combine the 95 to 100 + categories in the death data to the same age range as the population data (95+) in order to be able to match people of the same age in the same year.
+Because you need to be able to match the age categories between deaths and population.
 
-4)  calculate death rates for each age group of residents for each year.
+4)  Calculate death rates for each age group of residents for each year.
+Because age specific death rates of those resident in an area for a time period enables population size independent comparisons over time.
 
 5)  Multiple the individual age death rates in each year by a standard population, so that total death changes between years are considered on the basis of the same population and age structure in every year, to avoid mistaking non-fatal changes in population size and structure as changes in risk of death.
-In this example, the New Zealand population of 2022 is used
+In this example, the New Zealand population of 2022 is used.
+Because to get a whole of population death figure from death rates requires specifying a population the death rates belong to.
 
 6)  Total the standard deaths for each year, to generate the number (based on death rates) of deaths if the population had been the same as the year of the standard population, but the age specific risk of death was the same as the historical period.
+Because the desired outcome is to understand if there were more or fewer deaths independent of population change.
 
 7)  Preform a regression on the years 2013-2019 to account for trends beyond changes to the population value.
-The expected value as a result of the regression corrects for historical trends such as the improvement in health understanding and care (as per the article 2013-2019 is considered the optimal baseline for Australasia without triggering structural bias)
+The expected value as a result of the regression corrects for historical trends such as the improvement in health understanding and care (as per the article 2013-2019 is considered the optimal baseline for Australasia without triggering structural bias).
+Because life expectancy changes over time.
 
-8)  Project the regression trend into 2020 through 2022 to determine expected deaths given a population structure of the standard year. Compared them to actual standardised deaths
+8)  Project the regression trend into 2020 through 2022 to determine expected deaths given a population structure of the standard year.
+Compared them to actual standardised deaths.
+Because projecting a regression trend for a short period is an established counter-factual to use as an expected value for comparisons of unexpected change- testing if actual results are changing different to the past trend.
 

### 2024-07-13 rawfiles/supportFiles/Figure 1 Migration Deaths.R

@@ -1,136 +1,0 @@
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-
-# ITM Migration
-# ITM has lowest open category of 90+
-# so sets the age limits for others.
-colist <- c(1,(c(1:90,127) * 2))
-ITMf <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
-  select(all_of(colist)) 
-names(ITMf) <- c("Year", paste0("X",0:90))
-ITMflong <- ITMf |> 
-  filter(!is.na(X20)) |> 
-  gather(key="AgeText", value="Net", -Year) |> 
-  mutate(Sex="Female")
-
-colist <- c(1,((c(1:90,127) * 2) +1))
-ITMm <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
-  select(all_of(colist)) 
-names(ITMm) <- c("Year", paste0("X",0:90))
-ITMmlong <- ITMm |> 
-  filter(!is.na(X20)) |> 
-  gather(key="AgeText", value="Net", -Year) |> 
-  mutate(Sex="Male")
-
-ITM <- bind_rows(ITMflong,ITMmlong) |> 
-  mutate(AgeN = as.numeric(gsub("X","",AgeText)),
-         Year = as.numeric(Year)) |> 
-  filter(Year == 2019) |> 
-  summarise(.by=AgeN,
-            NetMigration = sum(Net))
-
-# VSD deaths
-
-VSD_male <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
-  select(c(1,27:127)) |> 
-  rename(Year = X., X0.years = Less.than.1.year, X1.years=X1.year) |> 
-  filter(!is.na(X0.years)) |> 
-  gather(key="Agetext", value="Deaths", -Year) |> 
-  mutate(Ageshort = gsub("^X","",Agetext),
-         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
-         Sex="Male",
-         Year = as.numeric(Year)) |> 
-  select(Year, Sex, AgeN, Deaths)
-VSD_female <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
-  select(c(1,154:254)) |> 
-  rename(Year = X., X0.years = Less.than.1.year.1, X1.years=X1.year.1) |> 
[…]

### 2024-07-13 rawfiles/supportFiles/Figure 2 Elderly.R

@@ -1,54 +1,0 @@
-# 75+ trend
-
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggplot2)
-library(ggthemes)
-library(ggtext)
-source("theme.R")
-six_cols <- colorblind_pal()(6)
-
-DPE_male <- read.csv("RawData/DPE_PopDecMean_age_sex.csv", skip=3) |> 
-  select(c(1,131)) |> 
-  rename(Year = 1, seventyfivePlus = X75.Years.and.Over) |> 
-  filter(!is.na(seventyfivePlus)) |> 
-  mutate(Year = as.numeric(Year),
-         Sex = "Male")
-DPE_female <- read.csv("RawData/DPE_PopDecMean_age_sex.csv", skip=3) |> 
-  select(c(1,267)) |> 
-  rename(Year = 1, seventyfivePlus = X75.Years.and.Over.1) |> 
-  filter(!is.na(seventyfivePlus)) |> 
-  mutate(Year = as.numeric(Year),
-         Sex = "Female")
-
-DPE75plus <- bind_rows(DPE_male, DPE_female) |> 
-  summarise(.by=Year,
-            Subpopulation = sum(seventyfivePlus)) |> 
-  filter(Year >2006)
-
-###
-LifeModel <- lm(Subpopulation ~ Year, 
-                data=DPE75plus |> filter(Year %in% 2012:2016))
-DPE75plus$Expected <- LifeModel$coefficients[1] + 
-  LifeModel$coefficients[2] * DPE75plus$Year
-
-ggplot(DPE75plus, aes(x=Year, y=Subpopulation))+
-  geom_line(aes(y=Expected), colour="#AAAAAA", 
-            linetype=2)+
-  geom_line(data=DPE75plus |> filter(Year %in% 2012:2016), 
-            aes(y=Expected), 
-            colour="#AAAAAA") +
-  geom_point() +
-  theme_davidhood() +
-  scale_x_continuous(breaks=c(2010,2015,2020))+
-  labs(title="NZ Resident population 75 plus years\nvs 2012-2016 trend",
-       y="75 plus poopulation", x="Year",
-       caption="Source: infoshare.stats.govt.nz : Population : Population Estimates - DPE : 
-Estimated Resident Population by Age and Sex (1991+) (Annual-Dec)")
-ggsave(filename="../rawfiles/md_figures/figure2.tiff",
[…]

### 2024-07-13 rawfiles/supportFiles/footnotes5and7.R

@@ -1,108 +1,0 @@
-library(dplyr)
-library(tidyr)
-
-# Deaths, rearranged into long-style database form
-# Single years then 100+ by sex
-VSD_male <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
-  select(c(1,27:127)) |> 
-  rename(Year = X., X0.years = Less.than.1.year, X1.years=X1.year) |> 
-  filter(!is.na(X0.years)) |> 
-  gather(key="Agetext", value="Deaths", -Year) |> 
-  mutate(Ageshort = gsub("^X","",Agetext),
-         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
-         Sex="Male",
-         Year = as.numeric(Year)) |> 
-  select(Year, Sex, AgeN, Deaths)
-VSD_female <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
-  select(c(1,154:254)) |> 
-  rename(Year = X., X0.years = Less.than.1.year.1, X1.years=X1.year.1) |> 
-  filter(!is.na(X0.years)) |> 
-  gather(key="Agetext", value="Deaths", -Year) |> 
-  mutate(Ageshort = gsub("^X","",Agetext),
-         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
-         Sex="Female",
-         Year = as.numeric(Year)) |> 
-  select(Year, Sex, AgeN, Deaths)
-
-VSD <- bind_rows(VSD_male, VSD_female)
-
-# Population Male/Female single years then 95+
-DPE_male <- read.csv("RawData/DPE_PopDecMean_age_sex.csv", skip=3) |> 
-  select(c(1,2:96,135)) |> 
-  rename(Year = X.) |> 
-  filter(!is.na(X0.Years)) |> 
-  gather(key="Agetext", value="Population", -Year) |> 
-  mutate(Ageshort = gsub("^X","",Agetext),
-         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
-         Sex="Male",
-         Year = as.numeric(Year)) |> 
-  select(Year, Sex, AgeN, Population)
-DPE_female <- read.csv("RawData/DPE_PopDecMean_age_sex.csv", skip=3) |> 
-  select(c(1,138:232,271)) |> 
-  rename(Year = X.) |> 
-  filter(!is.na(X0.Years.1)) |> 
-  gather(key="Agetext", value="Population", -Year) |> 
[…]

### 2024-07-13 rawfiles/supportFiles/footnotes8.R

@@ -1,19 +1,0 @@
-library(dplyr)
-
-DMM <- read.csv("RawData/DMM_CDRDec_TotPop.csv",
-                skip=1) |> 
-  filter(!is.na(Total.Population)) |> 
-  rename(Year=1,CDR=2) |> 
-  mutate(Year = as.numeric(Year))
-DMModel <- lm(CDR ~ Year, 
-              DMM |> filter(Year %in% 2015:2019))
-DMM$Expected <- DMModel$coefficients[1] +
-  DMModel$coefficients[2] * DMM$Year
-cumulatime <- DMM |> filter(Year >= 2020, Year <=2022)
-percentage = 100 * sum(cumulatime$CDR) /
-  sum(cumulatime$Expected) - 100
-print(round(percentage,0))
-
-
-
-

### 2024-07-13 rawfiles/supportFiles/footnotes9and10.R

@@ -1,149 +1,0 @@
-# working out that if 2020 had the same death rates by age as 2019
-# and the loss in migration, how much would the death rate go up by.
-
-library(dplyr)
-library(tidyr)
-
-# ITM Migration
-# ITM has lowest open category of 90+
-# so sets the age limits for others.
-colist <- c(1,(c(1:90,127) * 2))
-ITMf <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
-  select(all_of(colist)) 
-names(ITMf) <- c("Year", paste0("X",0:90))
-ITMflong <- ITMf |> 
-  filter(!is.na(X20)) |> 
-  gather(key="AgeText", value="Net", -Year) |> 
-  mutate(Sex="Female")
-
-colist <- c(1,((c(1:90,127) * 2) +1))
-ITMm <- read.csv("RawData/ITM_Dec_age_sex.csv", skip=4) |> 
-  select(all_of(colist)) 
-names(ITMm) <- c("Year", paste0("X",0:90))
-ITMmlong <- ITMm |> 
-  filter(!is.na(X20)) |> 
-  gather(key="AgeText", value="Net", -Year) |> 
-  mutate(Sex="Male")
-
-ITM <- bind_rows(ITMflong,ITMmlong) |> 
-  mutate(AgeN = as.numeric(gsub("X","",AgeText)),
-         Year = as.numeric(Year)) |> 
-  arrange(Sex, AgeN, Year) |> 
-  group_by(Sex,AgeN) |> 
-  mutate(followingYear = lead(Net),
-         change2following = followingYear - Net ) |> 
-  ungroup() |> 
-  select(Year,Sex,AgeN,Net,change2following)
-
-# VSD deaths
-
-VSD_male <- read.csv("RawData/VSD_Deaths_Dec_age_sex.csv", skip=2) |> 
-  select(c(1,27:127)) |> 
-  rename(Year = X., X0.years = Less.than.1.year, X1.years=X1.year) |> 
-  filter(!is.na(X0.years)) |> 
-  gather(key="Agetext", value="Deaths", -Year) |> 
-  mutate(Ageshort = gsub("^X","",Agetext),
-         AgeN = as.numeric(gsub("[^0123456789].*","",Ageshort)),
-         Sex="Male",
-         Year = as.numeric(Year)) |> 
-  select(Year, Sex, AgeN, Deaths)
[…]

### 2024-07-13 rawfiles/supportFiles/theme.R

@@ -1,24 +1,0 @@
-six_cols <- colorblind_pal()(6)
-
-theme_davidhood <- function(){
-  theme_minimal(base_family="Arial",
-                base_size = 9) %+replace%  
-    theme(axis.line.x = element_line(linewidth=0.2),
-          axis.line.y = element_line(linewidth=0.2),
-          axis.ticks = element_line(linewidth=0.2),
-          axis.title.y.left = element_text(margin = margin(t = 2, r = 2, b = 2, l = 2, unit = "pt")),
-          panel.background = element_rect(fill = "#FFFFFF", colour = "#FFFFFF"),
-          panel.grid = element_blank(),
-          plot.title = element_text(lineheight = 1.18, size=10,
-                                    margin=margin(t = 5, r = 0, b = 5, l = 15, unit = "pt"),
-                                    hjust=0, vjust=0),
-          plot.subtitle = element_text(lineheight = 1.18, size=8,
-                                    margin=margin(t = 0, r = 15, b = 5, l = 15, unit = "pt"),
-                                    hjust=0),
-          plot.background = element_rect(fill = "#FFFFFF", colour="#FFFFFF"),
-          plot.caption = element_text(margin=margin(t = 5, r = 5, b = 5, l = 5, unit = "pt"),
-                                      lineheight = 1.15,
-                                      size=6.5, hjust=1),
-          plot.caption.position = "plot")
-  
-}
